---
title: "P4_Metagenomics_analyses_Final"
output: html_document
date: "2026-01-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl) #v1.4.5
library(phyloseq) # v1.50.0
library(vegan) # v.2.7.2
library(tibble) # v3.3.1 
library(stringr) # v1.6.0 
library(flextable) # v0.9.11
library(officer) # v0.7.3
library(tableone) # v0.13.2
library(rlang) # v1.1.7
library(dplyr) # v1.2.0
library(lme4) # v1.1.38
library(emmeans) # v2.0.1
library(rstatix) # v0.7.3
library(ggplot2) # v4.0.2
library(ggpubr) # v0.6.3
library(doParallel) # v1.0.17
library(foreach) # v1.5.2
library(permute) # v0.9.10
library(tidytext) # v0.4.3
library(ggh4x) # v0.3.1
library(Maaslin2) # v1.20.0
library(randomForest) # v4.7.1.2
library(NatParksPalettes) #v0.2.0
library(forcats) #v1.0.1
library(ggExtra) #v.0.11.0

source("GM_development_functions.R")
source("~/Documents/GitHub/Preterm_Plasmidome/GM_development/customtab.R")

meta_data_path <- "~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260816_supplementarytable1_metadata.xlsx"


cat_var <- c("Sample_ID", "Patient", "Cohort", "Study","abx_group_updated", "BSI_exp_type", "Necrotizing_enterocolitis", "Received_antimicrobials", "Sex","Delivery_route","Maternal_ethnicity","Maternal_race","Site","Final_disposition",
             "Para", "Maternal_smoking",	"Maternal_alcohol",	"Maternal_substance_abuse",	"Maternal_preeclampsia",	"Maternal_diabetes",	"Maternal_chorioamnionitis","Maternal_hypertension", "mat_inpt_antimicrobe","mat_outpt_antimicrobial","Multiple_births", "PMA_bin",
             "BSI_status","Organism","Antimicrobial_1",	"Antimicrobial_2",	"Antimicrobial_3",	"Antimicrobial_4",
             "SIP_status", "DOL_bin")

num_var <- c("DOL","MOL","abx_exposure_count","Gestational_age_weeks", "Birthweight","Maternal_age_at_delivery","Gravida","antiviral_exposure", "aminoglycoside_exposure","penicillin_exposure","macrolide_exposure","firstgencephalosporin_exposure",
             "fourthgencephalosporin_exposure","thirdgencephalosporin_exposure","secondgencephalosporin_exposure","lincosamide_exposure",
             "antifungal_exposure","carbapenem_exposure","other_antimicrobials_exposure","other_non_antimicrobials_exposure","glycopeptide_exposure","med_exposure_count", "morphine_exposure", "fentanyl_exposure", "surfactant_exposure", "acetaminophen_exposure", "caffeine_exposure", "cholecalciferol_vit_d__exposure",	'vitamin_a_exposure',	'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	'bumetanide_bumax__exposure',	'other_exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	'sodium_bicarbonate_exposure','hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure','cum_days_formula','cum_days_milk','frac_days_formula', 'frac_days_milk',
             "Antimicrobial_duration_1",	"Antimicrobial_duration_2",	"Antimicrobial_duration_3",	"Antimicrobial_duration_4")

custom_colors_groups <- c("None" = "#8E949F", "Early only" = "#44637D", "Early and after" = "#AAC9ED", "Near-term" = "#613921")
custom_colors_names <- c("ABN" = "#8E949F", "EEO" = "#44637D", "EAS" = "#AAC9ED", "Near-term" = "#613921")
```

# Load data
```{r}
# Read the metaphlan3 table
mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/250318_P4_GM_dev_metaphlan3_merged_n2585_v2.txt", header = T, sep="\t", skip=1, check.names = FALSE))

# R adds an "X" to column names that start with a number. Steps to remove this below
#colnames(mphlan) <- gsub("^X([0-9])", "\\1", colnames(mphlan))
colnames(mphlan) <- as.character(colnames(mphlan))

metadata <- read_excel(meta_data_path)

## Processing data for phyloseq
rownames(mphlan) <- mphlan$clade_name
mphlan$clade_name <- NULL

## Create species data for phyloseq 
# keep rows with species info 
#mphlan.species <- mphlan
mphlan.species <- mphlan[grepl('s__', rownames(mphlan)), ]
# Drop rows with SBG (species level genome bins)
mphlan.species <- mphlan.species[!grepl('t__', rownames(mphlan.species)), ]
mphlan.species <- mphlan.species %>% dplyr::select_if(~sum(.) != 0)
mphlan.species <- as.matrix(mphlan.species)

## Create taxa table
species <- data.frame(Names = rownames(mphlan.species))
species <- data.frame(do.call('rbind', strsplit(as.character(species$Names), '|', fixed = TRUE)))
rownames(species) <- rownames(mphlan.species)
colnames(species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
species <- as.matrix(species)

## Metadata
rows_metadata <- rownames(metadata)[colnames(mphlan.species) %in% metadata$Sample_ID_R_final]
metadata <- data.frame(metadata)
rownames(metadata) <- metadata$Sample_ID_R_final

# Set categorical variables to factor
metadata[,cat_var] <- lapply(metadata[,cat_var], factor)
# Set numerical variables to numeric
metadata[,num_var] <- lapply(metadata[,num_var], as.numeric)
print(dim(metadata))

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(mphlan.species), rownames(metadata)))

ps <- phyloseq(otu_table(mphlan.species, taxa_are_rows = TRUE),
               sample_data(metadata),
               tax_table(species)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps.filt <- filter_taxa(ps, function(x) mean(x) > 0.01, TRUE)
ps.filt  <- transform_sample_counts(ps.filt, function(x) (x / sum(x))*100)

### Filter to DOL 0-60 and remove After only infants ########
ps_filt_dol60 <- ps.filt %>% subset_samples(DOL <= 60) 
ps_filt_dol60_meta <- data.frame(sample_data(ps_filt_dol60))

sample_data(ps_filt_dol60)$DOL_bin <- factor(sample_data(ps_filt_dol60)$DOL_bin)
sample_data(ps_filt_dol60)$abx_group_updated <- factor(sample_data(ps_filt_dol60)$abx_group_updated,
                                               levels = c("Early and after","Early only","None"))

###################### Merge with P4 cross-sectional cohort ###########
p4_mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/251020_P4_merged_metaphlan_n146.txt", header = T, sep="\t", check.names = FALSE))

## Processing data for phyloseq
rownames(p4_mphlan) <- p4_mphlan$clade_name
p4_mphlan$clade_name <- NULL

new_names <- colnames(p4_mphlan)
new_names <- gsub("^X", "", new_names)
new_names <- gsub("_", ".", new_names)
new_names <- paste0("p4_", new_names)

colnames(p4_mphlan) <- new_names

## Create species data for phyloseq 
# keep rows with species info 
p4_mphlan.species <- p4_mphlan[grepl('s__', rownames(p4_mphlan)), ]
# Drop rows with SBG (species level genome bins)
p4_mphlan.species <- p4_mphlan.species[!grepl('t__', rownames(p4_mphlan.species)), ]
p4_mphlan.species <- p4_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
p4_mphlan.species <- as.matrix(p4_mphlan.species)

## Create taxa table
p4_species <- data.frame(Names = rownames(p4_mphlan.species))
p4_species <- data.frame(do.call('rbind', strsplit(as.character(p4_species$Names), '|', fixed = TRUE)))
rownames(p4_species) <- rownames(p4_mphlan.species)
colnames(p4_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
p4_species <- as.matrix(p4_species)

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(p4_mphlan.species), rownames(metadata)))

ps_p4 <- phyloseq(otu_table(p4_mphlan.species, taxa_are_rows = TRUE),
               sample_data(metadata),
               tax_table(p4_species)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps_p4.filt <- filter_taxa(ps_p4, function(x) mean(x) > 0.01, TRUE)
ps_p4.filt  <- transform_sample_counts(ps_p4.filt, function(x) (x / sum(x))*100)

ps_p4_meta <- data.frame(phyloseq::sample_data(ps_p4.filt))

ps_p4.filt <- ps_p4.filt %>% subset_samples(abx_group_updated != "After only")
ps_p4.filt_meta <- data.frame(sample_data(ps_p4.filt))

sample_data(ps_p4.filt)$DOL_bin <- factor(sample_data(ps_p4.filt)$DOL_bin)
sample_data(ps_p4.filt)$abx_group_updated <- factor(sample_data(ps_p4.filt)$abx_group_updated,
                                               levels = c("Early and after","Early only","None"))

merged_ps <- merge_phyloseq(ps_filt_dol60, ps_p4.filt)
merged_ps_meta <- data.frame(sample_data(merged_ps))

merged_ps_meta$abx_group_names <- factor(merged_ps_meta$abx_group_names, levels = c("EAS", "EEO", "ABN"))
sample_data(merged_ps) <- merged_ps_meta


###### Filter out samples from infants that had NEC or BSI ########
merged_ps_filt <- merged_ps %>% subset_samples(Necrotizing_enterocolitis == "No") %>% subset_samples(BSI_status == "Negative") 
merged_ps_meta <- data.frame(sample_data(merged_ps_filt))

#save(merged_ps_filt, file = "260701_merged_ps_filt.RData")

merged_ps_filt_meta <- data.frame(sample_data(merged_ps_filt))

```

### Load near-term data
```{r}
drewg_meta <- read_excel(meta_data_path)


drewg_mphln <- mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/251016_near-term_n162_metaphlan3_merged.txt", header = T, sep="\t", skip=1, check.names = FALSE))
rownames(drewg_mphln) <- drewg_mphln$clade_name
drewg_mphln$clade_name <- NULL

drewg_mphlan.species <- drewg_mphln[grepl('s__', rownames(drewg_mphln)), ]
# Drop rows with SBG (species level genome bins)
drewg_mphlan.species <- drewg_mphlan.species[!grepl('t__', rownames(drewg_mphlan.species)), ]
drewg_mphlan.species <- drewg_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
drewg_mphlan.species <- as.matrix(drewg_mphlan.species)

## Create taxa table
drewg_species <- data.frame(Names = rownames(drewg_mphlan.species))
drewg_species <- data.frame(do.call('rbind', strsplit(as.character(drewg_species$Names), '|', fixed = TRUE)))
rownames(drewg_species) <- rownames(drewg_mphlan.species)
colnames(drewg_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
drewg_species <- as.matrix(drewg_species)

## Metadata
rows_metadata <- rownames(drewg_meta)[colnames(drewg_mphlan.species) %in% drewg_meta$Sample_ID_R_final]
drewg_meta <- data.frame(drewg_meta)
rownames(drewg_meta) <- drewg_meta$Sample_ID_R_final

# Set categorical variables to factor
drewg_meta[,cat_var] <- lapply(drewg_meta[,cat_var], factor)
# Set numerical variables to numeric
drewg_meta[,num_var] <- lapply(drewg_meta[,num_var], as.numeric)
print(dim(drewg_meta))

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we are only analyzing near-term samples from Gasparrini et al. 2019 that were collected before DOL 60  
print(setdiff(colnames(drewg_mphlan.species), rownames(drewg_meta)))

# 160 samples and 341 species
ps_term <- phyloseq(otu_table(drewg_mphlan.species, taxa_are_rows = TRUE),
               sample_data(drewg_meta),
               tax_table(drewg_species)) 

# 160 samples and 221 species
ps_term <- filter_taxa(ps_term, function(x) mean(x) > 0.01, TRUE)
ps_term  <- transform_sample_counts(ps_term, function(x) (x / sum(x))*100)

ps_term_meta <- data.frame(phyloseq::sample_data(ps_term))

## Merge preterm and term phyloseq objects #########
ps_infants <- merge_phyloseq(merged_ps_filt, ps_term)
ps_infants_meta <- data.frame(sample_data(ps_infants))

patient_summary <- ps_infants_meta %>%
  group_by(abx_group_names, Site) %>%
  summarize(n_unique_patients = n_distinct(Patient))

patient_summary2 <- ps_infants_meta %>%
  group_by(abx_group_names, Study) %>%
  summarize(n_unique_patients = n_distinct(Patient))
write.csv(patient_summary2, "260729_patient_summary.csv", row.names = F)

#save(ps_infants, file = "260616_ps_infants.RData")

#### Create metadata file for Supplementary Table 1 ######
#write.csv(ps_infants_meta, "260626_GM_Dev_Metadata_Supplementary_Table1_noBSI.csv", row.names = F)



```

Table 1
```{r}
library(tibble) # v3.3.1 for rownames_to_column() function
library(stringr) # v1.6.0 for str_squish()
library(officer) # v0.7.3
library(flextable) # v0.9.11
library(tableone) # v0.13.2
library(dplyr)

source("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/customtab.R")

meta_pt <- ps_infants_meta 

# Set categorical variables as factors 
categorical_vars <- c('Patient','abx_group_names', 'Sex', 'Delivery_route', 'Maternal_chorioamnionitis','Maternal_preeclampsia', 'BSI_status','SIP_status', 'mat_inpt_antimicrobe')

meta_pt[,categorical_vars] <- lapply(meta_pt[,categorical_vars], factor)
meta_pt$Delivery_route <- factor(meta_pt$Delivery_route, levels = c("Cesarean Section", "Vaginal"))


numerical_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks","frac_days_formula","cum_days_formula", 'cum_days_milk', 'frac_days_milk')
meta_pt[,numerical_vars] <- lapply(meta_pt[,numerical_vars], as.numeric)


## All numerical variables are non-normal so we'll use median (IQR) instead of mean (SD)
normality_results <- lapply(numerical_vars, function(v) {
  p <- shapiro.test(meta_pt[[v]])$p.value
  data.frame(variable = v, p_value = p)
}) %>% bind_rows()

# Replace NAs to Unknown since they are basically the same thing
meta_pt$Maternal_chorioamnionitis[meta_pt$Maternal_chorioamnionitis == "NA"] <- "Unknown"
meta_pt$Maternal_chorioamnionitis <- droplevels(meta_pt$Maternal_chorioamnionitis)
meta_pt <- meta_pt %>%
  mutate(Study_simplified = ifelse(Study == "This_study", "Current study", "Previous studies")) 

sample_data(ps_infants) <- meta_pt
ps_infants_meta <- data.frame(sample_data(ps_infants))

# Generate metadata comparison table 
comp_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks",'Site', 'Sex','Study_simplified', 'Delivery_route','BSI_status','SIP_status', 'Maternal_chorioamnionitis','Maternal_preeclampsia', 'mat_inpt_antimicrobe', "frac_days_formula","cum_days_formula", 'cum_days_milk', 'frac_days_milk')


tab_pt <- CreateTableOne(vars=comp_vars,
                           strata='abx_group_names', 
                           data=meta_pt)
tab_plot_pt <- print(tab_pt, 
  showAllLevels = TRUE,
  quote        = FALSE,
  noSpaces     = TRUE,
  test         = TRUE,
  contDigits   = 1,
  nonnormal    = numerical_vars,  
  printToggle  = FALSE,
  dropEqual    = TRUE,
  explain      = FALSE)

# Convert to dataframe
tab_pt_df <- as.data.frame(tab_plot_pt) %>% rownames_to_column(var = "Variable")
write.csv(tab_pt_df, "260811_noBSI_mat_inpt_antimicrobe/260811_Table1.csv", row.names = F)
```

## Supplemenary Figure 1A
```{r}
merged_ps_filt_meta$acute_abx_exposure_count

recent_abx_df <- merged_ps_filt_meta %>%
  mutate(recent_abx_exposure = if_else(acute_abx_exposure_count > 0, TRUE, FALSE)) %>% filter(recent_abx_exposure == T)

recent_abx_plot <- ggplot(recent_abx_df, aes(x = DOL, fill = abx_group_names)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.7, color = "black", linewidth = 0.2) +
  scale_fill_manual(values = c(custom_colors_names)) +  
  labs(
    x = "Day of life (DOL)",
    y = "Number of samples with recent antibiotic exposure",
    fill = "Antibiotic exposure group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank())
ggsave("260630_noBSI_v2/260724_P4_GM_Dev_recent_abx_exposure_dist.pdf", plot = recent_abx_plot, bg = "transparent", width = 8, height = 8)
```



## Alpha diversity (Related to Fig. 1A, Supplementary Figs 1B-D)
```{r}
library(rlang) # v1.1.7
library(dplyr) # v1.2.0
library(lme4) # v1.1.38
library(emmeans) # v2.0.1
library(rstatix) # v0.7.3
library(ggplot2) # v4.0.2
library(ggpubr) # v0.6.3

ps.filt.trans.term <- transform_sample_counts(ps_infants, function(x) trunc(x*100000)) #### ps_infants includes all term infants
adiv_incl_term <- estimate_richness(ps.filt.trans.term, measures = c("Observed", "Shannon"))
ps.filt.trans <- transform_sample_counts(merged_ps_filt, function(x) trunc(x*100000))
adiv <- estimate_richness(ps.filt.trans, measures=c('Observed', 'Shannon'))

#### Merge alpha diversity with metadata file ################################
sample_data(ps_infants)$Richness <- adiv_incl_term$Observed
sample_data(ps_infants)$Shannon <- adiv_incl_term$Shannon
sample_data(merged_ps_filt)$Richness <- adiv$Observed
sample_data(merged_ps_filt)$Shannon <- adiv$Shannon


alpha_merged_meta <- data.frame(sample_data(ps_infants))
alpha_merged_meta_pterm <- data.frame(sample_data(merged_ps_filt))
########################################## ------------------------- #################################

counts_by_bin <- alpha_merged_meta %>%
  filter(!is.na(Richness), !is.na(abx_group_names)) %>%
  mutate(abx_group_names = droplevels(abx_group_names)) %>%
  count(DOL_bin, abx_group_names, name = "n") %>%
  group_by(DOL_bin) %>% mutate(groups_present = n_distinct(abx_group_names)) %>% ungroup()
write.csv(counts_by_bin, "260810_noBSI_fentanyl_inpt_antimicrob/260810_adiv_comparisons_abx_exposure_n_by_group_names.csv", row.names = F)

## There are repeat samples and differences in DOL so we need to use linear models to account for these confounders
patient_repeats_per_DOL_bin <- alpha_merged_meta %>%
  group_by(DOL_bin, Patient) %>% summarise(n_samples = n(), .groups = "drop")

## Determining co-linearity between variables ##
cor.test(alpha_merged_meta$Gestational_age_weeks, alpha_merged_meta$Birthweight, method = "spearman") #rho=0.86, p < 2.2e-16

library(car)

# Fit a standard linear model first (VIF is calculated on lm, not lmer)
richness_vif_model <- lm(Richness ~ Gestational_age_weeks + abx_group_names + Delivery_route + Sex + 
                  SIP_status + Maternal_chorioamnionitis + Maternal_preeclampsia + frac_days_formula + frac_days_milk  + mat_inpt_antimicrobe, data = alpha_merged_meta)
vif(richness_vif_model) ## All variables seem to be safe to be included together and have a GVIF^(1/(2*Df)) value < √5 ≈ 2.24

########### Richness ##################
### Perform stat test and plot #######
lmer_emmeans_richness_updated <- perform_emmeans_lmer(df = alpha_merged_meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                                "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness_updated$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_richness.csv", row.names = F)

## Plot alpha diversity barplot binned by DOL 
richness_abx_dol <- ggplot(alpha_merged_meta, aes(x = abx_group_names, y = Richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Species Richness", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8))

# Access the contrast table
lmer_emmeans_richness_updated$stat.test

# Add emmeans results to plot
richness_abx_dol_emm_lmer <- richness_abx_dol + stat_pvalue_manual(
  lmer_emmeans_richness_updated$stat.test,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin")
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_Richness_abx_group_DOL_bp_emmeans.pdf", plot = richness_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)


sensitivity_results_richness <- compare_random_effects_sensitivity(
  df            = alpha_merged_meta,
  facet_var     = "DOL_bin",
  x_var         = "abx_group_names",
  y_var         = "Richness",
  fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                    "SIP_status", "Maternal_chorioamnionitis",
                    "Maternal_preeclampsia", "frac_days_formula", "frac_days_milk")
)
#write.csv(sensitivity_results_richness, "260627_richness_sensitivity_results_just_no_study_site.csv")


########### Shannon ##################
### Perform stat test and plot #######
lmer_emmeans_shannon <- perform_emmeans_lmer(df = alpha_merged_meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                                "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_shannon.csv", row.names = F)

shannon_abx_dol <- ggplot(alpha_merged_meta, aes(x = abx_group_names, y = Shannon, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Shannon Diversity", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
    theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8))



shannon_abx_dol_emm_lmer <- shannon_abx_dol + stat_pvalue_manual(
  lmer_emmeans_shannon$stat.test,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin")

ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_Shannon_abx_group_DOL_bp_emmeans.pdf", plot = shannon_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)

###### Facet by abx group #######
lmer_emmeans_shannon_DOL <- perform_emmeans_lmer(df = alpha_merged_meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                                "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")

shannon_abx_dol2 <- ggplot(alpha_merged_meta, aes(x = DOL_bin, y = Shannon, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ abx_group_names, nrow = 2, strip.position = "top") +
  labs(
    x = "DOL bin", 
    y = "Shannon Diversity", 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 18, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_Shannon_abx_group_facet_DOL_bp.pdf", plot = shannon_abx_dol2, bg = "transparent", width = 20, height = 10)

lmer_emmeans_shannon_DOL <- lmer_emmeans_shannon_DOL$stat.test %>% tidyr::separate(contrast, into = c("group1", "group2"), sep = "\\) - \\(") %>%
  mutate(
    group1 = gsub("^\\(|\\)$", "", group1),
    group2 = gsub("^\\(|\\)$", "", group2)
  )
write.csv(lmer_emmeans_shannon_DOL, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_shannon_by_DOL.csv", row.names = F)


lmer_emmeans_richness_DOL <- perform_emmeans_lmer(df = alpha_merged_meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Richness",
                                                  fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                                "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")

richness_abx_dol2 <- ggplot(alpha_merged_meta, aes(x = DOL_bin, y = Richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ abx_group_names, nrow = 2, strip.position = "top") +
  labs(
    x = "DOL bin", 
    y = "Species Richness", 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 18, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_Richness_abx_group_facet_DOL_bp.pdf", plot = richness_abx_dol2, bg = "transparent", width = 20, height = 10)

lmer_emmeans_richness_DOL <- lmer_emmeans_richness_DOL$stat.test %>% tidyr::separate(contrast, into = c("group1", "group2"), sep = "\\) - \\(") %>%
  mutate(
    group1 = gsub("^\\(|\\)$", "", group1),
    group2 = gsub("^\\(|\\)$", "", group2)
  )
write.csv(lmer_emmeans_richness_DOL, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_richness_by_DOL.csv", row.names = F)
```

ARG Comparison (Related to Fig 1B, Supplementary Fig. 1E, G)
```{r}
library(rlang) # v1.1.7
library(dplyr) # v1.2.0
library(lme4) # v1.1.38
library(emmeans) # v2.0.1
library(rstatix) # v0.7.3
library(ggplot2) # v4.0.2
library(ggpubr) # v0.6.3
library(reshape2) #v1.4.5

#look at overall ARG burden between groups
##load shortbred output
arg.full <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/250905_P4_GM_Dev_ShortBRED_merged_output_n2585.txt"))
arg.full <- arg.full %>% mutate(Sample_ID_R_final = gsub("-", ".", paste0("p4_", Sample)))
arg.p4 <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251022_GM_Dev_P4_shortBRED_joint_output.txt"))
arg.p4 <- arg.p4 %>% mutate(Sample_ID_R_final = paste0("p4_", gsub("_", ".", Sample)))
arg_df <- rbind(arg.full, arg.p4)

####### Plotting PCoA plot with Bray-curtis dissimilarities of ARG profiles #####
arg_wide <- arg_df %>%
  group_by(Sample_ID_R_final, Family) %>%
  summarise(Count = sum(Count), .groups = "drop") %>%   # in case of duplicate Sample-Family rows
  pivot_wider(names_from = Sample_ID_R_final, values_from = Count, values_fill = 0) %>%
  column_to_rownames("Family")

# phyloseq expects a matrix, with taxa (Family) as rows and samples as columns
arg_mat <- as.matrix(arg_wide)

# ---------------------------------------------------------------
# 2. Build the phyloseq object
# ---------------------------------------------------------------
ARG <- otu_table(arg_mat, taxa_are_rows = TRUE)

ps_arg <- phyloseq(otu_table(ARG, taxa_are_rows = taxa_are_rows(ARG)),
  sample_data(ps_infants_meta))

# Remove samples that have zero total counts
ps_arg.filt <- prune_samples(sample_sums(ps_arg) > 0, ps_arg) # 2202 samples by 9293 ARGs

# Remove ARGs that are zero across all samples
ps_arg.filt <- prune_taxa(taxa_sums(ps_arg.filt) > 0, ps_arg.filt) # 2202 samples by 1758 ARGs

ps_arg.filt <- filter_taxa(ps_arg.filt, function(x) mean(x) > 0.01, TRUE) # 2202 samples by 1490 ARGs
ps_arg.filt  <- transform_sample_counts(ps_arg.filt, function(x) (x / sum(x))*100)

# Now compute Bray-Curtis on the filtered object
bc_dist <- phyloseq::distance(ps_arg.filt, method = "bray")
bc_matrix <- as.matrix(bc_dist)
head(bc_matrix)

# ---------------------------------------------------------------
# 4. PCoA ordination
# ---------------------------------------------------------------
pcoa_ord <- ordinate(ps_arg.filt, method = "PCoA", distance = bc_dist)

# Percent variance explained by each axis
eig_vals <- pcoa_ord$values$Relative_eig
pc1_var <- round(eig_vals[1] * 100, 1)
pc2_var <- round(eig_vals[2] * 100, 1)


##### Plot ARG PCoA ######
pcoa_gm_dev_arg_abx <- phyloseq::plot_ordination(ps_arg.filt, pcoa_ord,
                                             type="samples", color= "abx_group_names")+
  geom_point(shape=17, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_names) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", linewidth = 3))

pcoa_gm_dev_arg_abx <- ggMarginal(pcoa_gm_dev_arg_abx, type='boxplot', groupFill=TRUE, size=10)
ggsave("260727_GM_Dev_PCoA_ARGs_abx_exposure_group.pdf", plot = pcoa_gm_dev_arg_abx, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_ARGs_abx_count <- phyloseq::plot_ordination(ps_arg.filt, pcoa_ord,
                                             type="samples", color= "abx_exposure_count")+
  geom_point(shape=17, size=3)+
  scale_color_gradientn(colors = natparks.pals("Arches")) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 16),
        legend.text = element_text(size=14),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))
ggsave("260727_GM_Dev_PCoA_ARGs_abx_exposure_count.pdf", plot = pcoa_gm_dev_ARGs_abx_count, bg = "transparent", width = 9, height = 6)


########## Alpha diversity plots #################
arg.wide <- as.data.frame(dcast(arg_df, Sample_ID_R_final ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg.wide) <- arg.wide$Sample_ID_R_final
arg.wide <- arg.wide[, c(-1)]

##sum counts to get RPKM per sample
arg.wide$RPKM <- rowSums(arg.wide)

##get ARG richness
arg.wide$Richness <- specnumber(arg.wide[, 1:ncol(arg.wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg.wide$Sample_ID_R_final <- row.names(arg.wide)
# Subset arg.wide to columns we want
arg.want <- arg.wide %>%
  select(c(Sample_ID_R_final, RPKM, Richness)) %>% dplyr::rename(ShortBRED_richness = Richness)

# Merge with metadata
arg.df <- merge(ps_infants_meta, arg.want, by = "Sample_ID_R_final")

##### Read in term ARG data ######
arg_term <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251016_DrewG_term_n162_shortbred_merged.txt"))

arg_term_wide <- as.data.frame(dcast(arg_term, Sample ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg_term_wide) <- arg_term_wide$Sample
arg_term_wide <- arg_term_wide[, c(-1)]

##sum counts to get RPKM per sample
arg_term_wide$RPKM <- rowSums(arg_term_wide)

##get ARG richness
arg_term_wide$Richness <- specnumber(arg_term_wide[, 1:ncol(arg_term_wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg_term_wide$Sample <- row.names(arg_term_wide)
# Subset arg.wide to columns we want
arg_term_want <- arg_term_wide %>%
  select(c(Sample, RPKM, Richness)) %>% dplyr::rename(Sample_ID_R_final = Sample, ShortBRED_richness = Richness)

# Merge with metadata
term_arg.df <- merge(ps_infants_meta, arg_term_want, by = "Sample_ID_R_final") 

#### Merge with preterm ARG data ###
arg.df2 <- arg.df %>% 
  select(Sample_ID_R_final, Patient, Study, Site, DOL_bin, Gestational_age_weeks, abx_group_updated, abx_group_names, RPKM, ShortBRED_richness)
term_arg.df2 <- term_arg.df %>% select(Sample_ID_R_final, Patient, Study, Site, DOL_bin, Gestational_age_weeks, abx_group_updated, abx_group_names, RPKM, ShortBRED_richness)

arg_full <- rbind(arg.df2, term_arg.df2)
write.csv(arg_full, "260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_ARG_data.csv", row.names = F)

##### Merge ARG data with metadata for Supplementary table 1 ##################################
arg_meta <- arg_full %>% select(Sample_ID_R_final, RPKM, ShortBRED_richness) %>% dplyr::rename(ARG_abundance_RPKM = RPKM,
                                                                                               ARG_richness = ShortBRED_richness)
arg_merged_meta <- merge(alpha_merged_meta, arg_meta, by = "Sample_ID_R_final", all = TRUE)

######################### ---------------------------------- ###################################

###### Plot and compare using linear mixed models ###########
lmer_emmeans_arg_rpkm <- perform_emmeans_lmer(df = arg_merged_meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "ARG_abundance_RPKM",
                                              fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                                "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_arg_rpkm$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_ARG_rpkm.csv", row.names = F)

counts_by_bin_arg <- arg_full %>%
  filter(!is.na(RPKM), !is.na(abx_group_names)) %>%
  mutate(abx_group_names = droplevels(abx_group_names)) %>%
  count(DOL_bin, abx_group_names, name = "n") %>%
  group_by(DOL_bin) %>% mutate(groups_present = n_distinct(abx_group_names)) %>% ungroup()
write.csv(counts_by_bin, "260811_noBSI_mat_inpt_antimicrobe/260811_abx_exposure_n_by_group_shortbred_comparison.csv", row.names = F)

shortbred_hits_rpkm <- ggplot(arg_full, aes(x = abx_group_names, y = RPKM, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Total ShortBRED hits (RPKM)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

lmer_emmeans_arg_rpkm_stats <- lmer_emmeans_arg_rpkm$stat.test %>%
  mutate(
    contrast_clean = trimws(contrast),
    group1 = str_match(contrast_clean, "^\\(?([^()]+?)\\)?\\s*-\\s*\\(?[^()]+?\\)?$")[, 2],
    group2 = str_match(contrast_clean, "^\\(?[^()]+?\\)?\\s*-\\s*\\(?([^()]+?)\\)?$")[, 2],
    group1 = trimws(group1),
    group2 = trimws(group2))

shortbred_hits_rpkm_stat  <- shortbred_hits_rpkm + stat_pvalue_manual(
  lmer_emmeans_arg_rpkm_stats,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin")
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_ARG_RPKM_abx_group_DOL_bp_emmeans.pdf", plot = shortbred_hits_rpkm_stat, bg = "transparent", width = 20, height = 10)

########### Total ShortBRED Hits (Richness) #########
lmer_emmeans_arg_richness <- perform_emmeans_lmer(df = arg_merged_meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "ARG_richness",
                                              fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                                "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_arg_richness$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_ARG_richness.csv", row.names = F)

shortbred_hits_richness <- ggplot(arg_full, aes(x = abx_group_names, y = ShortBRED_richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Total ShortBRED hits (Unique ARGs)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_ARG_Richness_abx_group_DOL_bp_emmeans.pdf", plot = shortbred_hits_richness, bg = "transparent", width = 20, height = 10)

###################-------------Richness normalized ARGs------------#######################
arg_merged_meta$norm_arg_abund <- arg_merged_meta$ARG_abundance_RPKM / arg_merged_meta$Richness

lmer_emmeans_norm_arg_richness <- perform_emmeans_lmer(df = arg_merged_meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "norm_arg_abund",
                                              fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status","Maternal_chorioamnionitis",
                                                          "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_formula",
                                                                "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_norm_arg_richness$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_richness_norm_ARG_rpkm.csv", row.names = F)

norm_arg_abund <- ggplot(arg_merged_meta, aes(x = abx_group_names, y = norm_arg_abund, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Richness Normalized Total ShortBRED Hits (RPKM)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 20, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/260810_P4_GM_Dev_Richness_Norm_ARG_rpkm_abx_group_.pdf", plot = norm_arg_abund, bg = "transparent", width = 20, height = 10)
```

Repeat measures PERMANOVA (Related to Fig. 1C, D and Supplementary Fig. 1F)
This section was run through a high-throughput computing facility 
```{r}
library(readxl)
library(vegan)
library(phyloseq)
library(tidyr)
library(dplyr)   

#source("GM_development_functions.R")
#source("Repeat_measures_PERMANOVA_function.R")
#load("260701_merged_ps_filt.RData")
dim(merged_ps_filt)

cat_var <- c("Sample_ID", "Patient", "Cohort", "abx_group", "abx_group_updated","BSI_exp_type", "Necrotizing_enterocolitis", "Received_antimicrobials", 
"Sequencing_runs","duplicate_count","Cohorts","Sex","Delivery_route","Maternal_ethnicity","Maternal_race","Site","Final_disposition",
             "Para", "Maternal_smoking",	"Maternal_alcohol",	"Maternal_substance_abuse",	"Maternal_preeclampsia",	"Maternal_diabetes",	
             "Maternal_chorioamnionitis","Maternal_hypertension", "mat_inpt_antimicrobe","mat_outpt_antimicrobial","Multiple_births", "SIP_status")

num_var <- c("DOL","MOL","acute_abx_exposure_count","abx_exposure_count_updated","Gestational_age_weeks", "Birthweight","Maternal_age_at_delivery","Gravida","acute_antiviral_exposure", 
"acute_aminoglycoside_exposure","acute_penicillin_exposure","acute_macrolide_exposure","acute_firstgencephalosporin_exposure",
             "acute_fourthgencephalosporin_exposure","acute_thirdgencephalosporin_exposure","acute_secondgencephalosporin_exposure","acute_lincosamide_exposure",
             "acute_antifungal_exposure","acute_carbapenem_exposure","acute_other_non_antimicrobials_exposure",
             "acute_glycopeptide_exposure","acute_med_exposure_count", "acute_morphine_exposure", "acute_fentanyl_exposure", "acute_surfactant_exposure", 
             "acute_acetaminophen_exposure", "acute_caffeine_exposure", "acute_cholecalciferol_vit_d__exposure",	'acute_vitamin_a_exposure',	
             'acute_famotidine_pepcid__exposure',	'acute_iron_exposure',	'acute_calcium_gluconate_exposure',	'acute_dopamine_exposure','acute_adenosine_exposure',	
             'acute_cosyntropin_cortrosyn_acth__exposure','acute_bumetanide_bumax__exposure',	'acute_other_exposure',	'acute_indomethacin_exposure',
             'acute_multi_vitamin_with_iron_exposure',	'acute_lasix_exposure',	'acute_atropine_exposure',	'acute_dobutamine_exposure',
             'acute_sodium_chloride_exposure',	'acute_levothyroxine_exposure',	'acute_dexamethasone_exposure',	'acute_phenobarbitol_exposure',	
             'acute_hydrocortisone_exposure',	'acute_ibuprofen_exposure',	'acute_chlorathiazide_diuril__exposure',	'acute_midazolam_versed__exposure',	
             'acute_potassium_chloride_exposure',	'acute_hepatitis_b_immune_globulin_exposure',	'acute_citric_acid_sodium_citrate_bicitra__exposure',	
             'acute_pancuronium_or_vecurionium_exposure',	'acute_insulin_exposure',	'acute_immunoglobulin_other_exposure',	'acute_multi_vitamin_no_iron_exposure', 
             'acute_levocarnitine_exposure','acute_digoxin_exposure','acute_hydraluronidase_hydase__exposure','acute_lorazapam_ativan__exposure',
             'acute_sodium_bicarbonate_exposure','acute_hydrochlorathiazide_hydrodiuril__exposure',
             'acute_spironolactone_hydrochlorathiazide_aldactazide__exposure','acute_tocopherol_acetate_vit_e__exposure','cum_days_formula',
             'cum_days_milk','frac_days_formula', 'frac_days_milk', "antiviral_exposure", "aminoglycoside_exposure","penicillin_exposure","macrolide_exposure","firstgencephalosporin_exposure",
             "fourthgencephalosporin_exposure","thirdgencephalosporin_exposure","secondgencephalosporin_exposure","lincosamide_exposure",
             "antifungal_exposure","carbapenem_exposure","other_antimicrobials_exposure","other_non_antimicrobials_exposure","glycopeptide_exposure",
             "med_exposure_count", "morphine_exposure", "fentanyl_exposure", "surfactant_exposure", "acetaminophen_exposure", "caffeine_exposure", 
             "cholecalciferol_vit_d__exposure",	'vitamin_a_exposure',	'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	
             'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	'bumetanide_bumax__exposure',	'other_exposure',	
             'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	
             'sodium_chloride_exposure',	'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	
             'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	'potassium_chloride_exposure',	
             'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	
             'insulin_exposure',	'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	
             'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	'sodium_bicarbonate_exposure','hydrochlorathiazide_hydrodiuril__exposure',	
             'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure')


### Run Repeat measure PERMANOVA #####
ps_dol60_meta <- data.frame(phyloseq::sample_data(merged_ps_filt))
ps_dol60_meta$Specimen_ID <- NULL
ps_dol60_meta$BSI_exp_type <- NULL
summary(ps_dol60_meta$abx_group_updated)

########## Collapsing medication columns for repeat measures PERMANOVA ############
# Define the category -> variable mapping
med_categories <- list(
  opioid_analgesic_sedative = c("morphine_exposure", "fentanyl_exposure",
                                 "midazolam_versed__exposure", "lorazapam_ativan__exposure"),
  nonopioid_analgesic       = c("acetaminophen_exposure"),
  pda_closure_nsaid         = c("indomethacin_exposure", "ibuprofen_exposure"),
  respiratory_support       = c("surfactant_exposure", "caffeine_exposure"),
  vitamins_nutritional      = c("cholecalciferol_vit_d__exposure", "vitamin_a_exposure",
                                 "iron_exposure", "multi_vitamin_with_iron_exposure",
                                 "multi_vitamin_no_iron_exposure", "levocarnitine_exposure"),
  gi_acid_suppression       = c("famotidine_pepcid__exposure",
                                 "citric_acid_sodium_citrate_bicitra__exposure"),
  electrolyte_mineral       = c("calcium_gluconate_exposure", "sodium_chloride_exposure",
                                 "potassium_chloride_exposure", "sodium_bicarbonate_exposure"),
  cv_inotrope_vasopressor   = c("dopamine_exposure", "dobutamine_exposure"),
  antiarrhythmic_cardiac    = c("digoxin_exposure", "atropine_exposure", "adenosine_exposure"),
  diuretics                 = c("bumetanide_bumax__exposure", "lasix_exposure",
                                 "chlorathiazide_diuril__exposure",
                                 "spironolactone_hydrochlorathiazide_aldactazide__exposure"),
  corticosteroids           = c("dexamethasone_exposure", "hydrocortisone_exposure"),
  neuromuscular_blocker     = c("pancuronium_or_vecurionium_exposure"),
  endocrine                 = c("levothyroxine_exposure", "insulin_exposure"),
  immunologic_biologic      = c("hepatitis_b_immune_globulin_exposure", "immunoglobulin_other_exposure"),
  anticonvulsant            = c("phenobarbitol_exposure"),
  other_misc                = c("hydraluronidase_hydase__exposure")
)

# Check that every variable in the mapping actually exists in df,
# and that every column in df is accounted for
all_mapped_vars <- unlist(med_categories)
missing_from_df <- setdiff(all_mapped_vars, names(ps_dol60_meta))
if (length(missing_from_df) > 0) {
  warning("These variables are in the mapping but not found in df: ",
          paste(missing_from_df, collapse = ", "))
}

# Build the collapsed dataframe
df_collapsed <- ps_dol60_meta  # keep original columns/other metadata intact

for (cat_name in names(med_categories)) {
  vars_in_cat <- intersect(med_categories[[cat_name]], names(ps_dol60_meta))
  if (length(vars_in_cat) == 0) next
  df_collapsed[[cat_name]] <- rowSums(ps_dol60_meta[, vars_in_cat, drop = FALSE], na.rm = TRUE)
}

# Optionally drop the original 39 individual columns, keeping only the collapsed ones
df_collapsed <- df_collapsed %>%
  select(-all_of(all_mapped_vars))

# Quick sanity check: compare sparsity before/after
sparsity_before <- sapply(ps_dol60_meta[, all_mapped_vars], function(x) mean(x == 0))
sparsity_after  <- sapply(df_collapsed[, names(med_categories)], function(x) mean(x == 0))

data.frame(
  category = names(med_categories),
  pct_zero_after = round(sparsity_after * 100, 1)
)

ps_dol60_meta <- df_collapsed 

### Determine metadata of interest 
med_category_names <- c("opioid_analgesic_sedative", "nonopioid_analgesic", "pda_closure_nsaid", "respiratory_support", "vitamins_nutritional", "gi_acid_suppression", "electrolyte_mineral", "cv_inotrope_vasopressor", "antiarrhythmic_cardiac", "diuretics", "corticosteroids", "neuromuscular_blocker", "endocrine", "immunologic_biologic", "anticonvulsant", "other_misc")
meta_of_interest <- c("Patient", 'Gestational_age_weeks', 'Birthweight', 'Sex', 'Delivery_route', 
                      'Site', 'Final_disposition', 'Maternal_race', 'Maternal_age_at_delivery', 'Gravida', 'Para', 'Maternal_smoking', 
                      'Maternal_alcohol', 'Maternal_substance_abuse', 'Maternal_preeclampsia', 'Maternal_diabetes', 
                      'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe', 'mat_outpt_antimicrobial', 
                      'Multiple_births', med_category_names)


# Rows without any NA
ps_dol60_meta_filt <- ps_dol60_meta %>% filter(if_all(all_of(meta_of_interest), ~ !is.na(.x)))
summary(ps_dol60_meta_filt$abx_group_updated)

# Rows with at least one NA --> 31 samples filtered out 
ps_dol60_meta_NAs <- ps_dol60_meta %>% filter(if_any(all_of(meta_of_interest), ~ is.na(.x)))
#write.csv(ps_dol60_meta_NAs,"251008_filtered samples.csv", row.names = F)

ps_filt2_dol60 <- phyloseq::subset_samples(merged_ps_filt, phyloseq::sample_names(merged_ps_filt) %in% ps_dol60_meta_filt$Sample_ID_R_final)

## Calculate bray curtis distance 
gm_dev_dol60.dist <- phyloseq::distance(ps_filt2_dol60, method="bray")

# Create subject_data which is a dataframe with per-subject metadata or data specific to each subject that don't change over time
gm_dev_subject_data <- ps_dol60_meta_filt %>%
  select(Patient, Gestational_age_weeks, Birthweight, Sex, Delivery_route, Site, Final_disposition, Maternal_race, 
  Maternal_age_at_delivery, Gravida, Para, Maternal_smoking, Maternal_alcohol, Maternal_substance_abuse, Maternal_preeclampsia, Maternal_diabetes, 
  Maternal_chorioamnionitis, Maternal_hypertension, mat_inpt_antimicrobe, mat_outpt_antimicrobial, Multiple_births)

gm_dev_subject_data <- gm_dev_subject_data %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(num_var, names(.))), as.numeric))

inherits(gm_dev_dol60.dist, "dist")
gm_dev_subject_data2 <- unique(gm_dev_subject_data)

rownames(gm_dev_subject_data2) <- gm_dev_subject_data2$Patient
gm_dev_subject_data2$Patient <- NULL

# Create sample_data file which is a dataframe with per-sample metadata
## do the rownames here also have to match?
gm_dev_sample_data <- ps_dol60_meta_filt %>%
  select(DOL, Cohort, abx_group_updated, SIP_status, Study, cum_days_formula,cum_days_milk,frac_days_formula,frac_days_milk,acute_abx_exposure_count,acute_antiviral_exposure,
  acute_aminoglycoside_exposure,acute_penicillin_exposure,acute_macrolide_exposure,acute_firstgencephalosporin_exposure,acute_fourthgencephalosporin_exposure,
  acute_thirdgencephalosporin_exposure,acute_secondgencephalosporin_exposure,acute_lincosamide_exposure,acute_antifungal_exposure,
  acute_carbapenem_exposure,acute_other_non_antimicrobials_exposure,acute_glycopeptide_exposure,
  acute_med_exposure_count,acute_morphine_exposure,acute_fentanyl_exposure,acute_surfactant_exposure,acute_acetaminophen_exposure,acute_caffeine_exposure,
  acute_cholecalciferol_vit_d__exposure,acute_vitamin_a_exposure,acute_famotidine_pepcid__exposure,acute_iron_exposure, acute_calcium_gluconate_exposure,
  acute_dopamine_exposure, acute_adenosine_exposure, acute_cosyntropin_cortrosyn_acth__exposure, acute_bumetanide_bumax__exposure, acute_other_exposure,
  acute_indomethacin_exposure, acute_multi_vitamin_with_iron_exposure, acute_lasix_exposure, acute_atropine_exposure, acute_dobutamine_exposure, 
  acute_sodium_chloride_exposure, acute_levothyroxine_exposure, acute_dexamethasone_exposure, acute_phenobarbitol_exposure, acute_hydrocortisone_exposure,
  acute_ibuprofen_exposure, acute_chlorathiazide_diuril__exposure, acute_midazolam_versed__exposure, acute_potassium_chloride_exposure, 
  acute_hepatitis_b_immune_globulin_exposure, acute_citric_acid_sodium_citrate_bicitra__exposure, acute_pancuronium_or_vecurionium_exposure, 
  acute_insulin_exposure, acute_immunoglobulin_other_exposure,	acute_multi_vitamin_no_iron_exposure, 
  acute_levocarnitine_exposure,acute_digoxin_exposure,acute_hydraluronidase_hydase__exposure,acute_lorazapam_ativan__exposure,
  acute_sodium_bicarbonate_exposure,acute_hydrochlorathiazide_hydrodiuril__exposure,
  acute_spironolactone_hydrochlorathiazide_aldactazide__exposure, acute_tocopherol_acetate_vit_e__exposure, abx_exposure_count_updated, antiviral_exposure, 
  aminoglycoside_exposure,penicillin_exposure,macrolide_exposure,firstgencephalosporin_exposure,fourthgencephalosporin_exposure,
  thirdgencephalosporin_exposure,secondgencephalosporin_exposure,lincosamide_exposure,antifungal_exposure,carbapenem_exposure,
  other_antimicrobials_exposure,other_non_antimicrobials_exposure,glycopeptide_exposure,med_exposure_count,morphine_exposure, 
  fentanyl_exposure,surfactant_exposure,acetaminophen_exposure,caffeine_exposure,cholecalciferol_vit_d__exposure,vitamin_a_exposure,	
  famotidine_pepcid__exposure,iron_exposure,calcium_gluconate_exposure,dopamine_exposure,adenosine_exposure,cosyntropin_cortrosyn_acth__exposure,	
  bumetanide_bumax__exposure,other_exposure,indomethacin_exposure,multi_vitamin_with_iron_exposure,lasix_exposure,atropine_exposure,
  dobutamine_exposure,sodium_chloride_exposure,levothyroxine_exposure,dexamethasone_exposure,phenobarbitol_exposure,hydrocortisone_exposure,	
  ibuprofen_exposure,chlorathiazide_diuril__exposure,midazolam_versed__exposure,potassium_chloride_exposure,hepatitis_b_immune_globulin_exposure,	
  citric_acid_sodium_citrate_bicitra__exposure,pancuronium_or_vecurionium_exposure,insulin_exposure,immunoglobulin_other_exposure,
  multi_vitamin_no_iron_exposure,levocarnitine_exposure,digoxin_exposure,hydraluronidase_hydase__exposure,lorazapam_ativan__exposure,
  sodium_bicarbonate_exposure,hydrochlorathiazide_hydrodiuril__exposure,spironolactone_hydrochlorathiazide_aldactazide__exposure,
  tocopherol_acetate_vit_e__exposure, opioid_analgesic_sedative, nonopioid_analgesic, pda_closure_nsaid, respiratory_support, vitamins_nutritional, gi_acid_suppression, electrolyte_mineral, cv_inotrope_vasopressor, antiarrhythmic_cardiac, diuretics, corticosteroids, neuromuscular_blocker,endocrine, immunologic_biologic, anticonvulsant, other_misc)

num_var <- c(num_var, med_category_names)
       
gm_dev_sample_data <- gm_dev_sample_data %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(num_var, names(.))), as.numeric))

# Create subject vector
gm_dev_subject_vector <- as.character(ps_dol60_meta_filt$Patient)
names(gm_dev_subject_vector) <- rownames(ps_dol60_meta_filt)

set.seed(42)
baseline_diet_var <- c('DOL','Gestational_age_weeks',	'Birthweight','Sex','Delivery_route',	'Cohort', 'Study', 'SIP_status','frac_days_milk','cum_days_formula', 
'cum_days_milk', 'frac_days_formula')
mat_var <- c('Maternal_race','Maternal_age_at_delivery',	'Gravida',	'Para',	'Maternal_smoking',	'Maternal_alcohol',	'Maternal_substance_abuse',	'Maternal_preeclampsia',
'Maternal_diabetes',	'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe',	'mat_outpt_antimicrobial',	'Multiple_births')
abx_var <- c('abx_group_updated','acute_abx_exposure_count','acute_antiviral_exposure', 'acute_aminoglycoside_exposure','acute_penicillin_exposure',
'acute_macrolide_exposure','acute_firstgencephalosporin_exposure','acute_fourthgencephalosporin_exposure','acute_thirdgencephalosporin_exposure',
'acute_secondgencephalosporin_exposure','acute_lincosamide_exposure','acute_antifungal_exposure','acute_carbapenem_exposure','acute_other_non_antimicrobials_exposure','acute_glycopeptide_exposure',
'abx_exposure_count_updated','antiviral_exposure',	'aminoglycoside_exposure',	'penicillin_exposure',	'macrolide_exposure',	
'firstgencephalosporin_exposure',	'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	
'secondgencephalosporin_exposure','lincosamide_exposure',	'antifungal_exposure',	'carbapenem_exposure',
'other_antimicrobials_exposure',	'other_non_antimicrobials_exposure',	'glycopeptide_exposure')
medications_var <- c('acute_med_exposure_count','acute_morphine_exposure','acute_fentanyl_exposure','acute_surfactant_exposure',
'acute_acetaminophen_exposure','acute_caffeine_exposure','acute_cholecalciferol_vit_d__exposure','acute_vitamin_a_exposure',
'acute_famotidine_pepcid__exposure','acute_iron_exposure', 'acute_calcium_gluconate_exposure', 'acute_dopamine_exposure', 'acute_adenosine_exposure', 
'acute_cosyntropin_cortrosyn_acth__exposure', 'acute_bumetanide_bumax__exposure', 'acute_other_exposure', 'acute_indomethacin_exposure', 
'acute_multi_vitamin_with_iron_exposure', 'acute_lasix_exposure', 'acute_atropine_exposure', 'acute_dobutamine_exposure', 'acute_sodium_chloride_exposure',
'acute_levothyroxine_exposure', 'acute_dexamethasone_exposure', 'acute_phenobarbitol_exposure', 'acute_hydrocortisone_exposure', 'acute_ibuprofen_exposure', 
'acute_chlorathiazide_diuril__exposure', 'acute_midazolam_versed__exposure', 'acute_potassium_chloride_exposure', 'acute_hepatitis_b_immune_globulin_exposure', 
'acute_citric_acid_sodium_citrate_bicitra__exposure', 'acute_pancuronium_or_vecurionium_exposure', 'acute_insulin_exposure', 'acute_immunoglobulin_other_exposure',	'acute_multi_vitamin_no_iron_exposure', 
'acute_levocarnitine_exposure','acute_digoxin_exposure','acute_hydraluronidase_hydase__exposure','acute_lorazapam_ativan__exposure',
'acute_sodium_bicarbonate_exposure','acute_hydrochlorathiazide_hydrodiuril__exposure',
'acute_spironolactone_hydrochlorathiazide_aldactazide__exposure','acute_tocopherol_acetate_vit_e__exposure', 
'med_exposure_count','morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	
'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	
'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure','bumetanide_bumax__exposure',
'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	
'sodium_chloride_exposure',	'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	
'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	'potassium_chloride_exposure',	
'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	
'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	
'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure')

mtdat_check <- cbind(
  gm_dev_subject_data2[gm_dev_subject_vector, , drop = FALSE],
  gm_dev_sample_data
)

metadata_order <- c(baseline_diet_var, mat_var, abx_var)
mtdat_check <- mtdat_check[, metadata_order, drop = FALSE]  # your current 56 vars

med_check <- mtdat_check[, medications_var, drop = FALSE]
med_cols_to_keep <- colSums(med_check) > 110
# Med cols that have a prevalence > 110
med_keep <- med_check[, med_cols_to_keep]
med_keep_names <- colnames(med_keep)

# Re-check aliasing on the CURRENT variable set (not the old 145/57 one)
alias_check <- alias(
  lm(as.formula(paste("rep(1, nrow(mtdat_check)) ~", 
                       paste(metadata_order, collapse = " + "))), 
     data = mtdat_check)
)
alias_check$Complete   # is this really NULL/empty now?

# Rank check
mm <- model.matrix(
  as.formula(paste("~", paste(metadata_order, collapse = " + "))),
  data = mtdat_check
)
qr(mm)$rank
ncol(mm)



# srun -c 24 --time=12:00:00 --mem=32G --pty bash -i 
# eval $( spack load --sh /7anzvct ) # load R
# Rscript <script_file_name>
# print("Testing variables")
# # baseline_diet_var and abx_var, mat_var checked
# all(medications_var %in% union(colnames(gm_dev_subject_data2), colnames(gm_dev_sample_data)))
# missing_var <- medications_var[!medications_var %in% union(colnames(gm_dev_subject_data2), colnames(gm_dev_sample_data))]
# missing_var
#all(baseline_diet_var  %in% union(colnames(gm_dev_subject_data2), colnames(gm_dev_sample_data)))


gm_dev_repeat_perm_abx <- PERMANOVA_repeat_measures_margin(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(baseline_diet_var, mat_var, abx_var),
  permutations=999, ncores=20)

# Apply Benjamini-Hochberg p-value adjustments (FDR) 
gm_dev_repeat_perm_abx$p_adj_BH <- p.adjust(gm_dev_repeat_perm_abx[["Pr(>F)"]], method = "BH")
write.csv(gm_dev_repeat_perm_abx, "260702_GM_Dev_repeat_permanova_BH_adj_results_abx.csv")
print("Done with PERMANOVA on abx, maternal and baseline and diet variables")


gm_dev_repeat_perm_med <- PERMANOVA_repeat_measures_original(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(baseline_diet_var, mat_var, medications_var),
  permutations=999, ncores=20)


# Apply Benjamini-Hochberg p-value adjustments (FDR) 
gm_dev_repeat_perm_med$p_adj_BH <- p.adjust(gm_dev_repeat_perm_med[["Pr(>F)"]], method = "BH")
write.csv(gm_dev_repeat_perm_med, "260702_GM_Dev_repeat_permanova_BH_adj_results_med.csv")
print("Done with PERMANOVA on medication, maternal and baseline and diet variables")

```

Creating Summary of Medication/Antibiotics/maternal factors exposure table for supplementary materials 
```{r}
library(tibble) # v3.3.1 for rownames_to_column() function
library(stringr) # v1.6.0 for str_squish()
library(officer) # v0.7.3
library(flextable) # v0.9.11
library(tableone) # v0.13.2
library(dplyr)

source("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/customtab.R")


### Convert medication exposure variables (numerical) to binary yes/no columns ####
med_vars <- c(
  "morphine_exposure", "fentanyl_exposure", "midazolam_versed__exposure", "lorazapam_ativan__exposure",
  "acetaminophen_exposure",
  "indomethacin_exposure", "ibuprofen_exposure",
  "surfactant_exposure", "caffeine_exposure",
  "cholecalciferol_vit_d__exposure", "vitamin_a_exposure", "tocopherol_acetate_vit_e__exposure",
  "iron_exposure", "multi_vitamin_with_iron_exposure", "multi_vitamin_no_iron_exposure",
  "levocarnitine_exposure",
  "famotidine_pepcid__exposure", "citric_acid_sodium_citrate_bicitra__exposure",
  "calcium_gluconate_exposure", "sodium_chloride_exposure", "potassium_chloride_exposure",
  "sodium_bicarbonate_exposure",
  "dopamine_exposure", "dobutamine_exposure",
  "digoxin_exposure", "atropine_exposure", "adenosine_exposure",
  "bumetanide_bumax__exposure", "lasix_exposure", "chlorathiazide_diuril__exposure",
  "hydrochlorathiazide_hydrodiuril__exposure", "spironolactone_hydrochlorathiazide_aldactazide__exposure",
  "dexamethasone_exposure", "hydrocortisone_exposure",
  "pancuronium_or_vecurionium_exposure",
  "levothyroxine_exposure", "insulin_exposure", "cosyntropin_cortrosyn_acth__exposure",
  "hepatitis_b_immune_globulin_exposure", "immunoglobulin_other_exposure",
  "phenobarbitol_exposure",
  "hydraluronidase_hydase__exposure"
)

# Create a new dataframe with binary exposure columns, named with a "_binary" suffix
nicu_df_binary <- merged_ps_filt_meta

for (var in med_vars) {
  new_var <- paste0(var, "_binary")
  nicu_df_binary[[new_var]] <- ifelse(merged_ps_filt_meta[[var]] > 0, 1, 0)
  # NOTE: NA counts will propagate as NA here (ifelse(NA > 0, ...) = NA),
  # so missing exposure data stays missing rather than silently becoming "0".
  nicu_df_binary[[new_var]] <- factor(nicu_df_binary[[new_var]], levels = c(0, 1))
}

# Optional: quick check of how many patients were exposed to each medication
binary_vars <- paste0(med_vars, "_binary")
exposure_summary <- sapply(nicu_df_binary[, binary_vars], function(x) {
  c(n_exposed = sum(x == 1, na.rm = TRUE),
    n_not_exposed = sum(x == 0, na.rm = TRUE),
    n_missing = sum(is.na(x)))
})
t(exposure_summary)


mat_var <- c('Maternal_race','Gravida',	'Maternal_smoking',	'Maternal_alcohol',	'Maternal_substance_abuse',	'Maternal_preeclampsia',
'Maternal_diabetes',	'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe',	'mat_outpt_antimicrobial',	'Multiple_births')

# Generate metadata comparison table 
comp_vars <- c(binary_vars, mat_var)

na_counts <- sapply(nicu_df_binary[, comp_vars], function(x) sum(is.na(x)))
# Count of literal "NA" strings per column
na_string_counts_all <- sapply(nicu_df_binary[, comp_vars], function(x) sum(x == "NA", na.rm = TRUE))
na_string_counts_all[na_string_counts_all > 0]

nicu_df_binary <- data.frame(nicu_df_binary)

tab_pt <- CreateTableOne(vars=comp_vars,
                           strata='abx_group_names', 
                           data=nicu_df_binary)
tab_plot_pt <- print(tab_pt, 
  showAllLevels = TRUE,
  quote        = FALSE,
  noSpaces     = TRUE,
  test         = TRUE,
  contDigits   = 1,
  printToggle  = FALSE,
  dropEqual    = TRUE,
  explain      = FALSE)

# Convert to dataframe
tab_pt_df <- as.data.frame(tab_plot_pt) %>% rownames_to_column(var = "Variable")
write.csv(tab_pt_df, "260630_noBSI_v2/260805_Medication_Maternal_vars_Table.csv", row.names = F)

```


Beta Diversity PCoA Plots (Related to Fig. 1C, Supplementary Figs. 1F,H)
```{r}
######## Plot ordination plot #####
gm_dev_ordination <- phyloseq::ordinate(merged_ps_filt, method='PCoA', distance='bray')

pcoa_gm_dev <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                         type="samples", color= "DOL_bin")+
  geom_point(shape=19, size=3)+
  #scale_color_manual(values = dol_bin_color_vector) + 
  scale_color_manual(values=natparks.pals("Arches"))+ 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev <- ggMarginal(pcoa_gm_dev, type='boxplot', groupFill=TRUE, size=10)
ggsave("260724_GM_Dev_PCoA_DOL_bin.pdf", plot = pcoa_gm_dev, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_abx <- phyloseq::plot_ordination(merged_ps_filt, gm_dev_ordination,
                                             type="samples", color= "abx_group_names")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_names) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", linewidth = 3))

pcoa_gm_dev_abx <- ggMarginal(pcoa_gm_dev_abx, type='boxplot', groupFill=TRUE, size=10)
ggsave("260724_GM_Dev_PCoA_abx_exposure_group.pdf", plot = pcoa_gm_dev_abx, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_abx_count <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                             type="samples", color= "abx_exposure_count")+
  geom_point(shape=19, size=3)+
  scale_color_gradientn(colors = natparks.pals("Arches")) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 16),
        legend.text = element_text(size=14),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))
ggsave("260724_GM_Dev_PCoA_abx_exposure_count.pdf", plot = pcoa_gm_dev_abx_count, bg = "transparent", width = 9, height = 6)

### By site (for revision)
pcoa_gm_dev_site <- phyloseq::plot_ordination(merged_ps_filt, gm_dev_ordination,
                                             type="samples", color= "Site")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values=natparks.pals("Saguaro"))+
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", linewidth = 3))

pcoa_gm_dev_site <- ggMarginal(pcoa_gm_dev_site, type='boxplot', groupFill=TRUE, size=10)
ggsave("260724_GM_Dev_PCoA_Site.pdf", plot = pcoa_gm_dev_site, bg = "transparent", width = 9, height = 6)

### Including term infants #####
gm_dev_ordination2 <- phyloseq::ordinate(ps_infants, method='PCoA', distance='bray')

pcoa_gm_dev_abx_term <- phyloseq::plot_ordination(ps_infants, gm_dev_ordination2,
                                             type="samples", color= "abx_group_names")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_names) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev_abx_term <- ggMarginal(pcoa_gm_dev_abx_term, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_abx_exposure_group_tern_joined.pdf", plot = pcoa_gm_dev_abx_term, bg = "transparent", width = 9, height = 6)

```

Batch effect tests (Revision): Finding samples to analyze
```{r}

table(ps_infants_meta_all$Necrotizing_enterocolitis)
this_study_infants <- ps_infants_meta_all %>% filter(Study == "This_study")

## Finding overlapping samples between P4 cross-sectional and NEC control infants
nec <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Kailun_NEC_Cohort/210429_nec_metadata.xlsx")
nec$Date_of_isolation <- as.Date(as.character(nec$Date_of_isolation))
nec$DOB <- as.Date(as.character(nec$DOB))
nec_control <- nec %>%
  filter(NEC_status == 0)
nec_control$Patient <- NULL
nec_control <- nec_control %>%
  dplyr::rename(Patient = Patient_modified)

p4_nec <- inner_join(this_study_infants, nec_control, by = c("Patient", "DOL"))
write.csv(p4_nec, "260603_p4_samples_repeat_seq_NEC_control.csv", row.names = F)
### BSI
bsi <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/DrewS_BSI_paper/Drewmetadatamerge/241021_DrewStoolSamplesBSIPaper.xlsx")
p4_bsi <- inner_join(this_study_infants, bsi, by = c("Patient", "DOL"))
write.csv(p4_bsi, "260603_p4_samples_repeat_seq_bsi.csv", row.names = F)

bsi_new <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metadata/BSI/250124_DrewS_other_BSI_samples.xlsx")
bsi_new$Patient <- as.factor(bsi_new$Patient)
bsi$Patient <- as.factor(bsi$Patient)

p4_bsi_new <- inner_join(this_study_infants, bsi_new, by = c("Patient", "DOL"))
write.csv(p4_bsi_new, "260603_p4_samples_repeat_seq_bsi_new.csv", row.names = F)


## Finding antibiotic exposed samples that were not previously sequenced
ps_infants_meta_all

molly <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Scripts/240515_molly_sample_base.xlsx")
molly_df <- molly %>% select(-c(PostmenstrualAgeDays, PostmenstrualAgeWeeks)) %>% mutate(DOL_floor = floor(DOL)) %>% select(-c(DOL)) %>% dplyr::rename(DOL = DOL_floor)

gasparrini <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Scripts/240515_gasparrini_sample_base.xlsx")
gasparrini_df <- gasparrini %>% select(-c(PostmenstrualAgeDays, PostmenstrualAgeWeeks)) %>% mutate(DOL_floor = floor(DOL)) %>% select(-c(DOL)) %>% dplyr::rename(DOL = DOL_floor)

bsi_df2 <- bsi %>% select(-c(Sample, Subject_id, DOL)) %>% 
  dplyr::rename(DOL = DOL_excelround, PaperID = Sample_id) %>% mutate(Cohort = "BSI")
bsi_new_df <- bsi_new %>% select(-c(Sequencing_runs, Path)) %>% dplyr::rename(PaperID = Sample_ID) %>% 
  mutate(Cohort = "BSI_new") %>% mutate(DOL = floor(DOL))

nec_control_df <- nec_control %>% select(Patient, DNA_Sample_ID, DOL) %>% dplyr::rename(PaperID = DNA_Sample_ID) %>% mutate(Cohort = "NEC_control")

all_pterm_samples <- rbind(molly_df, gasparrini_df, bsi_df2, bsi_new_df, nec_control_df)

####### Antibiotic exposed Samples not sequenced in previous studies
abx_exposed_infants <- ps_infants_meta_all %>% filter(abx_group_names == "EAS" | abx_group_names == "EEO")

new_seq_abx_exposed <- abx_exposed_infants %>% anti_join(all_pterm_samples, by = c("Patient", "DOL"))
patient_overlap <- new_seq_abx_exposed %>% semi_join(all_pterm_samples, by = "Patient")
p4_abx_exposed_samples <- patient_overlap
write.csv(p4_abx_exposed_samples, "260603_p4_new_abx_exposed_samples.csv", row.names = F)
previous_abx_exposed_samples_from_same_patient <- all_pterm_samples %>% semi_join(new_seq_abx_exposed, by = "Patient")

```

Batch effect tests: Taxonomic abundance tests
```{r}
# Read the metaphlan3 table
batch_mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/Batch_effect_test/260604_P4_testing_batch_effects_mphln3_merged_n28_v1.txt", header = T, sep="\t", skip=1, check.names = FALSE))

# R adds an "X" to column names that start with a number. Steps to remove this below
colnames(batch_mphlan) <- as.character(colnames(batch_mphlan))
names(batch_mphlan) <- gsub("\\.", "-", names(batch_mphlan))

batch_metadata <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/Batch_effect_test/260605_P4_batch_effect_tests_metadata.xlsx")

additional_meta <- merged_ps_filt_meta %>% dplyr::select(Sample_ID_R_final, Gestational_age_weeks, Birthweight, Sex, Delivery_route, DOL_bin)

batch_metadata2 <- merge(batch_metadata, additional_meta, by = "Sample_ID_R_final")

## Processing data for phyloseq
rownames(batch_mphlan) <- batch_mphlan$clade_name
batch_mphlan$clade_name <- NULL

## Create species data for phyloseq 
# keep rows with species info 
#mphlan.species <- mphlan
batch_mphlan.species <- batch_mphlan[grepl('s__', rownames(batch_mphlan)), ]
# Drop rows with SBG (species level genome bins)
batch_mphlan.species <- batch_mphlan.species[!grepl('t__', rownames(batch_mphlan.species)), ]
batch_mphlan.species <- batch_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
batch_mphlan.species <- as.matrix(batch_mphlan.species)

## Create taxa table
batch_species <- data.frame(Names = rownames(batch_mphlan.species))
batch_species <- data.frame(do.call('rbind', strsplit(as.character(batch_species$Names), '|', fixed = TRUE)))
rownames(batch_species) <- rownames(batch_mphlan.species)
colnames(batch_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
batch_species <- as.matrix(batch_species)

## Metadata
rows_metadata <- rownames(batch_metadata2)[colnames(batch_mphlan.species) %in% batch_metadata2$Sample_ID_R]
batch_metadata2 <- data.frame(batch_metadata2)
rownames(batch_metadata2) <- batch_metadata2$Sample_ID_R

# Set categorical variables to factor
batch_cat_var <- c("Sample_ID_R_final", "LTS_name", "Sample_ID_R", "Patient", "Study","abx_group_names", "Sample_type", "Delivery_route", 
                   "DOL_bin", "Sex")
batch_num_var <- c("DOL", "Gestational_age_weeks", "Birthweight")
batch_metadata2[,batch_cat_var] <- lapply(batch_metadata2[,batch_cat_var], factor)
# Set numerical variables to numeric
batch_metadata2[,batch_num_var] <- lapply(batch_metadata2[,batch_num_var], as.numeric)
print(dim(batch_metadata2))

# Create Study_simplified column
batch_metadata2 <- batch_metadata2 %>%
  mutate(Study_simplified = ifelse(Study == "This_study", "Current study", "Previous studies")) 


# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(batch_mphlan.species), rownames(batch_metadata2)))

ps_batch <- phyloseq(otu_table(batch_mphlan.species, taxa_are_rows = TRUE),
               sample_data(batch_metadata2),
               tax_table(batch_species)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps_batch.filt <- filter_taxa(ps_batch, function(x) mean(x) > 0.01, TRUE)
ps_batch.filt  <- transform_sample_counts(ps_batch.filt, function(x) (x / sum(x))*100)


#####################---------------Plotting PCoA Plots to test batch effects --------##################################
######## Plot ordination plot #####

batch_ordination <- phyloseq::ordinate(ps_batch.filt, method='PCoA', distance='bray')

pcoa_infant_id <- phyloseq::plot_ordination(ps_batch.filt, batch_ordination,
                                            type = "samples", color = "Patient", shape = "abx_group_names") +
  geom_point(size = 3) +
  scale_color_manual(values = c(
  "#E6194B","#3CB44B","#FFE119","#4363D8","#F58231",
  "#911EB4","#42D4F4","#F032E6","#BFEF45","#FABED4",
  "#469990","#DCBEFF","#9A6324","#808000")) +
  scale_shape_manual(values = c(16, 17, 15, 18)) + 
  theme_classic() +
  theme(axis.text.y = element_text(size = 20),
        axis.text.x = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.position = 'left',
        plot.title = element_text(size = 15),
        panel.border = element_rect(fill = NA, colour = "black", linewidth = 3))

ggsave("260608_Batch_effect_test_PCoA_infant_ID_shape_by_original_abx_group.pdf", plot = pcoa_infant_id, bg = "transparent", width = 9, height = 6)


pcoa_infant_id2 <- phyloseq::plot_ordination(ps_batch.filt, batch_ordination,
                                            type = "samples", color = "Patient", shape = "Sample_type") +
  geom_point(size = 3) +
  scale_color_manual(values = c(
  "#E6194B","#3CB44B","#FFE119","#4363D8","#F58231",
  "#911EB4","#42D4F4","#F032E6","#BFEF45","#FABED4",
  "#469990","#DCBEFF","#9A6324","#808000")) +
  scale_shape_manual(values = c(16, 17)) + 
  theme_classic() +
  theme(axis.text.y = element_text(size = 20),
        axis.text.x = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.position = 'left',
        plot.title = element_text(size = 15),
        panel.border = element_rect(fill = NA, colour = "black", linewidth = 3))

ggsave("260608_Batch_effect_test_PCoA_infant_ID_shape_by_sample_type.pdf", plot = pcoa_infant_id2, bg = "transparent", width = 9, height = 6)

pcoa_infant_batch <- phyloseq::plot_ordination(ps_batch.filt, batch_ordination,
                                            type = "samples", color = "abx_group_names", shape = "Study_simplified") +
  geom_point(size = 3) +
  scale_color_manual(values = custom_colors_names) +
  scale_shape_manual(values = c(16, 17)) + 
  theme_classic() +
  theme(axis.text.y = element_text(size = 20),
        axis.text.x = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.position = 'left',
        plot.title = element_text(size = 15),
        panel.border = element_rect(fill = NA, colour = "black", linewidth = 3))
ggsave("260724_Batch_effect_test_PCoA_infant_ID_shape_by_batch2.pdf", plot = pcoa_infant_batch, bg = "transparent", width = 9, height = 6)

pcoa_coords <- as.data.frame(batch_ordination$vectors)

write.csv(pcoa_coords, "260608_Batch_effect_test_PCoA_coordiantes.csv", row.names = T)

##################--------------------Alpha diversity comparisons------------------##################################
ps_batch.trans <- transform_sample_counts(ps_batch.filt, function(x) trunc(x*100000))
batch_adiv <- estimate_richness(ps_batch.trans, measures=c('Observed', 'Shannon'))

sample_data(ps_batch.filt)$Richness <- batch_adiv$Observed
sample_data(ps_batch.filt)$Shannon <- batch_adiv$Shannon

#### Merge alpha diversity with metadata file ################################
batch_alpha_div <- data.frame(phyloseq::sample_data(ps_batch.filt)) %>% select(Sample_ID_R, Richness, Shannon)
batch_metadata_adiv <- merge(batch_metadata2, batch_alpha_div, by = "Sample_ID_R", all = TRUE) 

# Create Study_simplified column
batch_metadata_adiv <- batch_metadata_adiv %>%
  mutate(Study_simplified = ifelse(Study == "This_study", "Current study", "Previous studies")) %>% arrange(Richness)

## Plot alpha diversity barplot binned by DOL 
batch_richness_nofacet <- ggplot(batch_metadata_adiv, aes(x = Study_simplified, y = Richness)) +
  # Connect same patient across studies
  geom_line(aes(group = Patient), color = "grey70", linewidth = 0.5) +
  # Boxplot without fill to keep lines visible
  geom_boxplot(outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5, width = 0.4) +
  geom_point(aes(color = abx_group_names, shape = Sample_type), size = 3) +
  scale_shape_manual(values = c(17, 19)) + 
  scale_color_manual(values=custom_colors_names)+
  scale_x_discrete(limits = c("Previous studies", "Current study")) +
  labs(
    x = NULL, 
    y = "Species Richness", 
    color = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.position = "right")
ggsave("260608_Batch_effect_Richness_nofacet.pdf", plot = batch_richness_nofacet, bg = "transparent", width = 10, height = 8)

batch_richness <- ggplot(batch_metadata_adiv, aes(x = Study_simplified, y = Richness)) +
  # Connect same patient across studies
  geom_line(aes(group = Patient), color = "grey70", linewidth = 0.5) +
  # Boxplot without fill to keep lines visible
  geom_boxplot(outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5, width = 0.4) +
  geom_point(aes(color = abx_group_names, shape = Sample_type), size = 3) +
  scale_shape_manual(values = c(17, 19)) + 
  scale_color_manual(values=custom_colors_names)+
  scale_x_discrete(limits = c("Previous studies", "Current study")) +
  labs(
    x = NULL, 
    y = "Species Richness", 
    color = "Antibiotic Exposure Group",
  ) +
  facet_wrap(~ DOL_bin, strip.position = "top") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.position = "right",
    panel.border = element_rect(fill = NA, colour = "black", linewidth = 2),
    strip.text = element_text(size = 18, face = "bold"),  
    strip.background = element_rect(fill = "grey90", colour = "black", linewidth = 1))
ggsave("260608_Batch_effect_Richness_facet.pdf", plot = batch_richness, bg = "transparent", width = 10, height = 8)


## Plot alpha diversity barplot binned by DOL 
batch_shannon <- ggplot(batch_metadata_adiv, aes(x = Study_simplified, y = Shannon)) +
  # Connect same patient across studies
  geom_line(aes(group = Patient), color = "grey70", linewidth = 0.5) +
  # Boxplot without fill to keep lines visible
  geom_boxplot(outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5, width = 0.4) +
  geom_point(aes(color = abx_group_names, shape = Sample_type), size = 3) +
  scale_shape_manual(values = c(17, 19)) + 
  scale_color_manual(values=custom_colors_names)+
  scale_x_discrete(limits = c("Previous studies", "Current study")) +
  labs(
    x = NULL, 
    y = "Shannon Diversity", 
    color = "Antibiotic Exposure Group",
  ) +
  facet_wrap(~ DOL_bin, strip.position = "top") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.position = "right",
    panel.border = element_rect(fill = NA, colour = "black", linewidth = 2),
    strip.text = element_text(size = 18, face = "bold"),  
    strip.background = element_rect(fill = "grey90", colour = "black", linewidth = 1))
ggsave("260608_Batch_effect_Shannon_facet.pdf", plot = batch_shannon, bg = "transparent", width = 10, height = 8)

batch_shannon_nofacet <- ggplot(batch_metadata_adiv, aes(x = Study_simplified, y = Shannon)) +
  # Connect same patient across studies
  geom_line(aes(group = Patient), color = "grey70", linewidth = 0.5) +
  # Boxplot without fill to keep lines visible
  geom_boxplot(outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5, width = 0.4) +
  geom_point(aes(color = abx_group_names, shape = Sample_type), size = 3) +
  scale_shape_manual(values = c(17, 19)) + 
  scale_color_manual(values=custom_colors_names)+
  scale_x_discrete(limits = c("Previous studies", "Current study")) +
  labs(
    x = NULL, 
    y = "Shannon Diversity", 
    color = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.position = "right")
ggsave("260609_Batch_effect_Shannon_nofacet.pdf", plot = batch_shannon_nofacet, bg = "transparent", width = 10, height = 8)

### Statistical tests
# Reshape to wide format so each patient has one row
batch_richness_wide <- batch_metadata_adiv %>%
  select(Patient, Study_simplified, Richness) %>%
  tidyr::pivot_wider(names_from = Study_simplified, values_from = Richness)

# Run paired Wilcoxon signed-rank test
batch_richness_pairedwilcox <- wilcox.test(batch_richness_wide$`Current study`, 
            batch_richness_wide$`Previous studies`, 
            paired = TRUE,
            exact = FALSE)   # set to FALSE to avoid errors with ties

batch_shannon_wide <- batch_metadata_adiv %>%
  select(Patient, Study_simplified, Shannon) %>%
  tidyr::pivot_wider(names_from = Study_simplified, values_from = Shannon)

# Run paired Wilcoxon signed-rank test
batch_shannon_pairedwilcox <- wilcox.test(batch_shannon_wide$`Current study`, 
            batch_shannon_wide$`Previous studies`, paired = TRUE, exact = FALSE)   # set to FALSE to avoid errors with ties

```


Repeat Measure PERMANOVA Summary Figure (Related to Fig. 1D)
```{r}
DOL <- c("DOL")
study <- c("Study")
baseline_var <- c('Gestational_age_weeks','Birthweight','Sex','Delivery_route',	'Site', 'SIP_status')
diet_var <- c('cum_days_formula', 'cum_days_milk', 'frac_days_formula','frac_days_milk')
mat_var <- c('Maternal_race','Maternal_age_at_delivery',	'Gravida',	'Para',	'Maternal_smoking',	'Maternal_alcohol',	'Maternal_substance_abuse',	'Maternal_preeclampsia',
'Maternal_diabetes',	'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe',	'mat_outpt_antimicrobial',	'Multiple_births')
abx_var <- c('abx_group_updated','abx_exposure_count','aminoglycoside_exposure',	'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	
'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure','lincosamide_exposure',		'carbapenem_exposure',
	'other_antimicrobials_exposure', 'glycopeptide_exposure','acute_abx_exposure_count', 'acute_aminoglycoside_exposure', 'acute_penicillin_exposure','acute_macrolide_exposure', 'acute_firstgencephalosporin_exposure','acute_fourthgencephalosporin_exposure',
'acute_thirdgencephalosporin_exposure','acute_secondgencephalosporin_exposure','acute_lincosamide_exposure','acute_carbapenem_exposure','acute_other_non_antimicrobials_exposure','acute_glycopeptide_exposure')
medications_var <- c('med_exposure_count', 'other_non_antimicrobials_exposure',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure', 'antifungal_exposure', 'acute_med_exposure_count','acute_morphine_exposure','acute_fentanyl_exposure','acute_surfactant_exposure',
'acute_acetaminophen_exposure','acute_caffeine_exposure','acute_cholecalciferol_vit_d__exposure',
'acute_vitamin_a_exposure','acute_famotidine_pepcid__exposure','acute_iron_exposure',
'acute_calcium_gluconate_exposure','acute_dopamine_exposure','acute_bumetanide_bumax__exposure',
'acute_other_exposure','acute_indomethacin_exposure','acute_multi_vitamin_with_iron_exposure',
'acute_lasix_exposure','acute_atropine_exposure','acute_dobutamine_exposure',
'acute_sodium_chloride_exposure','acute_levothyroxine_exposure','acute_dexamethasone_exposure',
'acute_phenobarbitol_exposure','acute_hydrocortisone_exposure','acute_ibuprofen_exposure',
'acute_chlorathiazide_diuril__exposure','acute_midazolam_versed__exposure',
'acute_potassium_chloride_exposure','acute_citric_acid_sodium_citrate_bicitra__exposure',
'acute_pancuronium_or_vecurionium_exposure','acute_insulin_exposure', "acute_multi_vitamin_no_iron_exposure", "acute_levocarnitine_exposure", "acute_digoxin_exposure", "acute_lorazapam_ativan__exposure", "acute_sodium_bicarbonate_exposure",
'acute_antifungal_exposure', 'acute_antiviral_exposure', 'antiviral_exposure')

#### Repeat measures PERMANOVA results were compiled manually ######
# results <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260630_noBSI_v2/beta-diversity/PERMANOVA/260713_GM_Dev_repeat_permanova_results_combined.xlsx")
# results$p_adj_BH <- p.adjust(results[["Pr(>F)"]], method = "BH")
# results$p_adj_BH <- as.numeric(results$p_adj_BH)
# write.csv(results, "260713_GM_Dev_repeat_PERMANOVA_results_combined_p_adj.csv", row.names = F)

results <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260630_noBSI_v2/beta-diversity/PERMANOVA/260713_GM_Dev_repeat_PERMANOVA_results_combined_p_adj.xlsx")

results <- results %>%
  mutate(significant = p_adj_BH < 0.05,
         Category = case_when(Variable %in% DOL ~ "DOL",
                              Variable %in% study ~ "Study",
                              Variable %in% baseline_var ~ "Baseline",
                              Variable %in% diet_var ~ "Diet",
                              Variable %in% mat_var ~ "Maternal",
                              Variable %in% abx_var ~ "Antibiotic",
                              Variable %in% medications_var ~ "Non-antibiotic medication"))
var_order <- c("DOL","Baseline","Study","Diet", "Maternal","Antibiotic", "Non-antibiotic medication")

results$Category <- factor(results$Category, levels = var_order)

results_plot <- results %>% mutate(R2_percent = R2 * 100,
                                   sig_label = case_when(p_adj_BH <= 0.001 ~ "***",
                                                         p_adj_BH <= 0.01 ~ "**",
                                                         p_adj_BH <= 0.05 ~ "*")) %>%
  filter(p_adj_BH < 0.05) %>%
  arrange(desc(R2_percent)) %>%
  mutate(Label = factor(Label, levels = unique(Label))) %>%
  ungroup()

perm_cat_vector = c("DOL" = "#3A1F46","Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

permanova_sign_var_p <- ggplot(results_plot, aes(x = Label, y = R2_percent, fill = Category)) +
  geom_col(width = 0.8, color = "black") + 
  geom_text(aes(label = sig_label, y = R2_percent + 0.15), size = 5, vjust = 0) +
  scale_fill_manual(values=perm_cat_vector) + 
  labs(x = NULL, y = "Variance explained (%)", fill = NULL) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    axis.line.x = element_line(color = "black", linewidth = 0.6),
    axis.ticks.x = element_line(color = "black", linewidth = 0.4),
    plot.margin = ggplot2::margin(t = 10, r = 10, b = 40, l = 10))  # extra space for category bar
ggsave("260716_GM_Dev_repeat_measure_PERMANOVA_sign_var.pdf", plot = permanova_sign_var_p, bg = "transparent", width = 12, height = 6)
```


Run Linear Mixed Models (MaAslin2) to test association between species and metadata (Related to Fig. 2A-C, Supplementary Fig. 2)
```{r}
############### ----------------------------- Taxa ----------------------##################################################

# results_plot can be found in the code block above
repeat_permanova_sign_var <- results_plot$Variable

pterm_infants_otu <- data.frame(otu_table(merged_ps_filt), check.names = FALSE) 
pterm_infants_meta <- data.frame(sample_data(merged_ps_filt), check.names = FALSE)

cat_var <- c("Patient",'Sex','Delivery_route', 'SIP_status', 'Site','Maternal_substance_abuse','Maternal_preeclampsia','Maternal_diabetes', 'Maternal_chorioamnionitis', 'Maternal_hypertension','mat_inpt_antimicrobe', 'mat_outpt_antimicrobial','Multiple_births','Cohort', 'Site', 
             'abx_group_names', 'Study')

num_var <- c('Gestational_age_weeks', 'Birthweight','DOL','abx_exposure_count', 'antiviral_exposure', 'antiviral_exposure',	'aminoglycoside_exposure',	
'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure',	
'lincosamide_exposure',	'antifungal_exposure',	'carbapenem_exposure',	'other_antimicrobials_exposure',	'other_non_antimicrobials_exposure','glycopeptide_exposure', 'med_exposure_count',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure', 
  'cum_days_formula', 'cum_days_milk', 'frac_days_formula','frac_days_milk', "acute_fentanyl_exposure", "acute_cholecalciferol_vit_d__exposure", "acute_dopamine_exposure", "acute_levothyroxine_exposure", "acute_hydrocortisone_exposure", "acute_insulin_exposure", "acute_surfactant_exposure", "acute_iron_exposure", "acute_indomethacin_exposure", "acute_dexamethasone_exposure",
"acute_citric_acid_sodium_citrate_bicitra__exposure", "acute_lorazapam_ativan__exposure")

pterm_infants_meta <- pterm_infants_meta %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(results_plot$Variable, names(.))), as.numeric))

pterm_infants_meta$abx_group_names <- factor(pterm_infants_meta$abx_group_names, levels = c("ABN", "EEO", "EAS"))

maaslin_taxa_results <- Maaslin2(input_data = pterm_infants_otu,
                                input_metadata = pterm_infants_meta,
                                transform = "LOG",
                                normalization = "NONE",
                                min_abundance = 0.0001,
                                min_prevalence = 0.1,
                                output = "~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260810_noBSI_fentanyl_inpt_antimicrob/beta-diversity/MaAslin2/Taxa/",
                                fixed_effects = c("abx_group_names","Delivery_route","SIP_status", "Sex",
                                                  "fentanyl_exposure","Maternal_chorioamnionitis",
                                                  "Maternal_preeclampsia","mat_inpt_antimicrobe","frac_days_milk",
                                                  repeat_permanova_sign_var),
                                random_effects = c("Patient", "Site", "Study"))

############ --------------- Coefficient Plots ------------------------------#############################################
taxa_sign_out <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260810_noBSI_fentanyl_inpt_antimicrob/beta-diversity/MaAslin2/Taxa/significant_results.xlsx")

num_samples <- 2208

taxa_sign_out_filt <- taxa_sign_out %>%
  filter(qval < 0.05) %>%
  mutate(
    # grab text AFTER "g__" up to the next dot
    genus  = str_extract(feature, "(?<=g__)[^.]+"),
    genus  = if_else(is.na(genus) | genus == "", "Unknown", genus),
    value  = as.factor(value),
    feature = gsub(".*s__", "", feature)
  )

unique_taxa <- unique(taxa_sign_out_filt$feature)
unique_genera <- unique(taxa_sign_out_filt$genus)

maaslin_taxa_abxGroup <- taxa_sign_out %>% filter(metadata == "abx_group_names") %>% 
  #filter(qval < 0.05) %>%
  mutate(
    # grab text AFTER "g__" up to the next dot
    genus  = str_extract(feature, "(?<=g__)[^.]+"),
    genus  = if_else(is.na(genus) | genus == "", "Unknown", genus),
    value  = as.factor(value),
    feature = gsub(".*s__", "", feature)
  ) %>%
  arrange(desc(coef))

#obtain min and max coefficient values (needed for plotting later)
min <- min(maaslin_taxa_abxGroup$coef)
max <- max(maaslin_taxa_abxGroup$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
maaslin_taxa_abxGroup <- maaslin_taxa_abxGroup %>% mutate(species = fct_reorder(feature, coef))

maaslin_taxa_abxGroup$coefmax <- maaslin_taxa_abxGroup$coef + maaslin_taxa_abxGroup$stderr
maaslin_taxa_abxGroup$coefmin <- maaslin_taxa_abxGroup$coef - maaslin_taxa_abxGroup$stderr

maaslin_taxa_abxGroup$CI95upper <- maaslin_taxa_abxGroup$coef + 1.96*maaslin_taxa_abxGroup$stderr
maaslin_taxa_abxGroup$CI95lower <- maaslin_taxa_abxGroup$coef - 1.96*maaslin_taxa_abxGroup$stderr

maaslin_taxa_abxGroup$value <- factor(maaslin_taxa_abxGroup$value, levels = c("EEO", "EAS"))

# Plot coefficients
x_min_lim <- (floor(min / 5) * 6) 
x_max_lim <- (ceiling(max / 5) * 6) 
colors_list <- c("#44637D","#AAC9ED")

abxgroup_coeff_p <- ggplot(maaslin_taxa_abxGroup, aes(x=coef, y=species,color=value))+
  geom_point(size = 3)+
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.1), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  scale_color_manual(values=colors_list)+
  theme_linedraw()+
  theme(axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=17),
        axis.title.x = element_text(size=17),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.position = "bottom",
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", linewidth=1))+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  xlab('Coefficient') + ggtitle("Antibiotic Exposure Group")
ggsave("260810_noBSI_fentanyl_inpt_antimicrob/beta-diversity/MaAslin2/Taxa/Plots/260810_GM_dev_abx_group_MaAslin2_taxa_coeff.pdf", plot = abxgroup_coeff_p, bg = "transparent", width = 10, height = 8)

#### Generate prevalence plot ###
### Access species abundance data 
abund_gmdev <- psmelt(merged_ps_filt)

abund_gmdev <- abund_gmdev %>%  mutate(
    genus = str_extract(OTU, "g__[^|]+"),  # extract 'g__Genus'
    genus = str_remove(genus, "g__"),          # remove 'g__' prefix
    genus = if_else(is.na(genus), "Unknown", genus),
    OTU=gsub(".*s__", "", OTU))

abund_abx_group <- subset(abund_gmdev, OTU %in% maaslin_taxa_abxGroup$feature)

# Re-order feature abs abundances according to coefficient value
# This order is preserved in filename, which was previously ordered: 
order_abx_group <- data.frame(OTU = unique(maaslin_taxa_abxGroup$feature))
# sort = F makes sure that the order is based on order_index df which is ordered by coefficient value
abund_abx_group.df <- merge(order_abx_group, abund_abx_group, by="OTU", sort = F)
abund_abx_group.df$OTU <- as.factor(abund_abx_group.df$OTU)
abund_abx_group.df <- abund_abx_group.df %>% mutate(OTU = fct_relevel(OTU, order_abx_group$OTU))

# For each significant feature in each sample, set AbunNotZero == 1 if its 
# abundance is greater than 0. 
abund_abx_group.df$AbunNotZero <- 1*(abund_abx_group.df$Abundance > 0)

# Summarize pathway prevalence by variable of interest
abund_abx_group.df.n0 <- abund_abx_group.df %>% 
  group_by(OTU, abx_group_names) %>%
  dplyr::summarise(N = n(), SumNotZero = sum(AbunNotZero)) %>%
  mutate(PercentNotZero = 100*(SumNotZero / N)) %>%
  ungroup()
write.csv(abund_abx_group.df.n0, "260716_GM_dev_abx_group_maaslin2_species_prevalence.csv", 
          row.names = F)

prev_p_abx_group <- ggplot(abund_abx_group.df.n0, aes(x = abx_group_names, y=OTU, fill=PercentNotZero))+
  geom_tile()+
  scale_fill_distiller(palette ='RdBu', type='seq', breaks = c(0,50,100), 'Prevalence (%)')+
  theme_classic()+
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(size=17, angle=45, hjust=1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=17),
        legend.text = element_text(size = 18),
        panel.border = element_rect(fill=NA, colour="black", size=1))+
  scale_y_discrete(limits=rev)
ggsave("260716_GM_dev_abx_group_MaAslin2_taxa_prev.pdf", plot = prev_p_abx_group, bg = "transparent", width = 6, height = 8)

combined_abx_group <- abxgroup_coeff_p | prev_p_abx_group 
ggsave("260716_GM_dev_MaAslin2_taxa_plot.pdf", plot = combined_abx_group, bg = "transparent", width = 16, height = 6)

############---------------- Other numerical variables ------------------------------------------#########
library(tidytext) # v0.4.3
library(stringr) # v1.6.0
library(ggh4x) # v0.3.1

perm_cat_vector = c("DOL" = "#3A1F46", "Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

taxa_sign_out_num <- taxa_sign_out_filt %>% filter(metadata != "abx_group_names") %>%
  mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 
write.csv(taxa_sign_out_num, "260716_sign_MaAslin2_taxa.csv", row.names = F)

taxa_sign_out_num <- read_excel("260716_sign_MaAslin2_taxa.xlsx")
taxa_sign_out_num <- taxa_sign_out_num %>% mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 


taxa_sign_out_num$metadata_category <- factor(taxa_sign_out_num$metadata_category, levels = names(perm_cat_vector))

labels_df <- results_plot %>% select(Variable, Label)
taxa_sign_out_num2 <- merge(taxa_sign_out_num, labels_df, by = "Variable")

#obtain min and max coefficient values (needed for plotting later)
min <- min(taxa_sign_out_num2$coef)
max <- max(taxa_sign_out_num2$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
#maaslin_taxa <- taxa_sign_out_num %>% mutate(species = fct_reorder(feature, coef))

taxa_sign_out_num2$coefmax <- taxa_sign_out_num2$coef + taxa_sign_out_num2$stderr
taxa_sign_out_num2$coefmin <- taxa_sign_out_num2$coef - taxa_sign_out_num2$stderr

taxa_sign_out_num2$CI95upper <- taxa_sign_out_num2$coef + 1.96*taxa_sign_out_num2$stderr
taxa_sign_out_num2$CI95lower <- taxa_sign_out_num2$coef - 1.96*taxa_sign_out_num2$stderr

taxa_sign_out_num2$direction <- factor(taxa_sign_out_num2$direction, levels = c("Positive", "Negative"))

genus_color_vector <- c(
  "Veillonella"      = "#A8A55C",  
  "Streptococcus"      = "#68434E",  
  "Clostridium"    = "#B07159",  
  "Klebsiella"     = "#2F4B7A",  
  "Enterobacter"      = "#8C2B0E",  
  "Staphylococcus"      = "#B47E83",  
  "Enterococcus"  = "#F2A450",  
  "Escherichia"    = "#9F7E59",  
  "Cutibacterium"          = "#6EA57B", 
  "Finegoldia"       = "#8DA0CB",  
  "Clostridioides"          = "#088096"  
)

taxa_dol_coeff <- taxa_sign_out_num2 %>% filter(metadata_category == "DOL") %>% ggplot(., aes(x = coef, y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.09), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#3A1F46", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260721_GM_dev_MaAslin2_taxa_DOL_var_coeff_plot.pdf", plot = taxa_dol_coeff, bg = "transparent", width = 5, height = 4)

taxa_baseline_coeff <- taxa_sign_out_num2 %>% filter(metadata_category == "Baseline") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.1), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#7F4B89", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260721_GM_dev_MaAslin2_taxa_baseline_var_coeff_plot.pdf", plot = taxa_baseline_coeff, bg = "transparent", width = 4, height = 3)

taxa_diet_coeff <- taxa_sign_out_num2 %>% filter(metadata_category == "Diet") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.1), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#F3DAE4", color = "black"),
    strip.text = element_text(size = 11, face = "bold", color = "black"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.box = "horizontal")
ggsave("260721_GM_dev_MaAslin2_taxa_diet_var_coeff_plot.pdf", plot = taxa_diet_coeff, bg = "transparent", width = 12, height = 4)

taxa_abx_coeff <- taxa_sign_out_num2 %>% filter(metadata_category == "Antibiotic") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.1), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#E3A5D6", color = "black"),
    strip.text = element_text(size = 12, face = "bold", color = "black"),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.box = "horizontal")
ggsave("260721_GM_dev_MaAslin2_taxa_abx_var_coeff_plot.pdf", plot = taxa_abx_coeff, bg = "transparent", width = 18, height = 6)

taxa_med_coeff <- taxa_sign_out_num2 %>% filter(metadata_category == "Non-antibiotic medication") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.1), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#7C5467", color = "black"),
    strip.text = element_text(size = 12, face = "bold", color = "white"),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 12, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260721_GM_dev_MaAslin2_taxa_med_var_coeff_plot.pdf", plot = taxa_med_coeff, bg = "transparent", width = 22, height = 10)

###################### ---------------------------- ShortBRED-------------------###########################################
## Merge term infant ARGs with class
arg.map <- read.delim('~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/ShortBRED_mapping.txt',header = F)
arg_df_pterm <- arg_df %>% filter(Sample_ID_R_final %in% ps_infants_meta$Sample_ID_R_final)

## Merge preterm infant ARGs with class
arg.byclass <- merge(arg_df_pterm, arg.map, by.x = 'Family', by.y = 'V1') %>% 
  dplyr::rename(Class = V2)

arg.byclass_preterm <- arg.byclass %>% 
  dplyr::group_by(Sample_ID_R_final, Class) %>% 
  dplyr::summarise(Count = sum(Count))

arg.byclass_preterm.wide <- dcast(arg.byclass_preterm, Sample_ID_R_final ~ Class, value.var = 'Count')
arg.byclass_preterm.wide[is.na(arg.byclass_preterm.wide)] <- 0
row.names(arg.byclass_preterm.wide) <- arg.byclass_preterm.wide$Sample_ID_R_final
arg.byclass_preterm.wide$Sample_ID_R_final <- NULL


# Create metadata file
arg_meta <- ps_infants_meta %>% select(Sample_ID_R_final, Patient, Study, Site, DOL, DOL_bin, Gestational_age_weeks, Birthweight, Sex, Delivery_route, abx_exposure_count, abx_group_updated, abx_group_names) %>% filter(abx_group_names != "Near-term")

arg_meta$abx_group_names <- factor(arg_meta$abx_group_names, levels = c("ABN", "EEO", "EAS"))

############--------- Run MaAslin2 -------------------------------############################

shortbred_preterm <- Maaslin2(input_data = arg.byclass_preterm.wide,
                                input_metadata = arg_meta,
                                transform = "LOG",
                                normalization = "NONE",
                                min_abundance = 0.0001,
                                min_prevalence = 0.1,
                                output = "~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260630_noBSI_v2/beta-diversity/MaAslin2/ShortBRED/repeat_permanova_sign_vars",
                                fixed_effects = c("abx_group_names, ABN",repeat_permanova_sign_var),
                                random_effects = c("Patient", "Site", "Study"))


########## Create ARG coefficient plots #############
arg_sign_out <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260630_noBSI_v2/beta-diversity/MaAslin2/ShortBRED/repeat_permanova_sign_vars/significant_results.xlsx")

num_samples <- 2208

arg_sign_out_filt <- arg_sign_out %>% filter(qval < 0.05) 

perm_cat_vector = c("DOL" = "#3A1F46", "Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

arg_sign_out_num <- arg_sign_out_filt %>% filter(metadata != "abx_group_names") %>%
  mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 
write.csv(arg_sign_out_num, "260722_sign_MaAslin2_ARG_Class.csv", row.names = F)

arg_sign_out_num <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/260630_noBSI_v2/beta-diversity/MaAslin2/ShortBRED/repeat_permanova_sign_vars/260722_sign_MaAslin2_ARG_Class.xlsx")
arg_sign_out_num <- arg_sign_out_num %>% mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 


arg_sign_out_num$metadata_category <- factor(arg_sign_out_num$metadata_category, levels = names(perm_cat_vector))


#obtain min and max coefficient values (needed for plotting later)
min <- min(arg_sign_out_num$coef)
max <- max(arg_sign_out_num$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
#maaslin_taxa <- taxa_sign_out_num %>% mutate(species = fct_reorder(feature, coef))

arg_sign_out_num$coefmax <- arg_sign_out_num$coef + arg_sign_out_num$stderr
arg_sign_out_num$coefmin <- arg_sign_out_num$coef - arg_sign_out_num$stderr

arg_sign_out_num$CI95upper <- arg_sign_out_num$coef + 1.96*arg_sign_out_num$stderr
arg_sign_out_num$CI95lower <- arg_sign_out_num$coef - 1.96*arg_sign_out_num$stderr

arg_sign_out_num$direction <- factor(arg_sign_out_num$direction, levels = c("Positive", "Negative"))
write.csv(arg_sign_out_num, "260722_MaAslin2_ARG_num_sign.csv", row.names = F)

arg_dol_coeff <- arg_sign_out_num %>% filter(metadata_category == "DOL") %>% ggplot(., aes(x = coef, y = feature)) +
  geom_point(aes(color = "#AC3414"), alpha = 0.9, size = 3, shape = 17) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.09), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  #scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = "Gene function") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#3A1F46", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260722_GM_dev_MaAslin2_ARG_DOL_var_coeff_plot.pdf", plot = arg_dol_coeff, bg = "transparent", width = 7, height = 6)

arg_baseline_coeff <- arg_sign_out_num %>% filter(metadata_category == "Baseline") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = "#AC3414"), alpha = 0.9, size = 3, shape = 17) +
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.09), orientation = "y")+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = "Antibiotic Class") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#7F4B89", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260722_GM_dev_MaAslin2_ARG_baseline_var_coeff_plot.pdf", plot = arg_baseline_coeff, bg = "transparent", width = 4, height = 3)

arg_abx_coeff <- arg_sign_out_num %>% filter(metadata_category == "Antibiotic") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = "#AC3414"), alpha = 0.9, size = 3, shape = 17) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', linewidth=1.5)+
  geom_errorbar(aes(xmax = CI95upper, xmin = CI95lower, width = 0.09), orientation = "y")+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = "Antibiotic Class") +
  theme(
    strip.background = element_rect(fill = "#E3A5D6", color = "black"),
    strip.text = element_text(size = 12, face = "bold", color = "black"),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260722_GM_dev_MaAslin2_ARG_abx_var_coeff_plot.pdf", plot = arg_abx_coeff, bg = "transparent", width = 4, height = 3)
```


Random Forest Analysis to predict microbiota age (Related to Fig. 2D,E and Supplementary Fig. 3)
```{r}
library(randomForest) # v4.7.1.2
library(dplyr) #  v1.2.0
library(tidyr)
library(stringr) # v1.6.0

# 1. Apply random forests regression to regress relative abundances of all species predicted by MetaPhlAn3 in infant stool samples against their chronological age 
## Default parameters used: ntree = 10,000, importance = TRUE
# 2. Fivefold cross-validation was performed using 'rfcv' function over 100 iterations to estimate the min number of features (taxa) needed to accurately predict microbiota age. This was done using 100 iterations of the "importance" function 
### Importance determined based on the percentage increase in mean-squared error of microbiota age prediction when the relative abundance values of each taxon were randomly permuted 
# 3. Sparse model consisting of the 50 most important features constructed and trained on half of abx-naive near-term infants that were randomly selected 
# 4. Validate model in remaining half of abx-naive near-term infants 
# 5. Apply model to each preterm infant group 

# Renormalize ps object to ensure that relative abundance sums to 100
ps_filt_dol60_renorm  <- transform_sample_counts(merged_ps_filt, function(x) (x / sum(x))*100 )

# # Access taxa relative abundances, and sample metadata
# mtg_relabund.df <- data.frame(otu_table(ps_m3_mtg_long_renorm), check.names = FALSE)

# Load your taxa relative abundance dataframe and chronological age vector
relabund_meta_full.df <- psmelt(ps_filt_dol60_renorm) 
# Rename each OTU so that the name only has species names not the full phylogenetic breakdown 
relabund_meta.df <- relabund_meta_full.df %>%
    mutate(OTU = sapply(str_split(OTU, "\\|"), function(x) if(length(x) >= 7) x[7] else NA))

# Subset to only rows from ABN samples  
relabund_meta_naive.df <- relabund_meta.df %>%
  filter(abx_group_names == "ABN")

# randomForest requires that each row represent a sample and columns represent relative abundance of taxa
## we will also need a column for the response variable, age
# This means we will need to reformat the output of psmelt() which outputs a long-format dataframe where:
## each row is a specific taxon in a sample and columns include sample ID and metadata like age 

# First pivot dataframe to wide format and 
taxa_abundance <- relabund_meta_naive.df %>%
  select(Sample, OTU, Abundance) %>%
  pivot_wider(names_from = OTU, values_from = Abundance, values_fill = 0) # values_fill = 0 makes sure that missing values (e.g., taxa absent in a sample) are set to 0

# Add age as a separate column
taxa_abundance <- taxa_abundance %>%
  left_join(
    relabund_meta_naive.df %>% select(Sample, DOL) %>% distinct(),
    by = "Sample"
  )

# 2. Randomly split the data upfront into training and validation sets (60:40)
set.seed(142)  # For reproducibility
train_indices <- sample(1:nrow(taxa_abundance), size = 0.6 * nrow(taxa_abundance))
train_data <- taxa_abundance[train_indices, ]
validation_data <- taxa_abundance[-train_indices, ]

# Separate predictors and response for training set
x_train <- train_data %>% select(-Sample, -DOL)
y_train <- train_data$DOL

# Step 2: Perform 100 iterations for feature importance and RFCV
iterations <- 100
importance_list <- vector("list", iterations)
rfcv_results_list <- vector("list", iterations)

for (i in 1:iterations) {
  cat("Iteration:", i, "\n")
  
  # Fit Random Forest model on the training set
  rf_model <- randomForest(
    x = x_train,
    y = y_train,
    ntree = 10000,
    importance = TRUE
  )
  
  # Extract feature importance
  importance_df <- as.data.frame(importance(rf_model))
  importance_df <- importance_df %>%
    mutate(Taxon = rownames(importance_df),
           Iteration = i) %>%
    arrange(desc(IncNodePurity))
  
  # Append this iteration's importance data to the list
  importance_list[[i]] <- importance_df
  
  # Perform RFCV for this iteration
  rfcv_results <- rfcv(
    trainx = x_train,
    trainy = y_train,
    cv.fold = 10,
    step = 0.5
  )
  
  # Store the rfcv results with the iteration number
  rfcv_results_list[[i]] <- data.frame(
    Iteration = i,
    n_features = rfcv_results$n.var,
    error = rfcv_results$error.cv
  )
}

# Combine all feature importance results
feature_importance_df <- bind_rows(importance_list)

# Aggregate feature importance (Mean and SD for both %IncMSE and IncNodePurity)
aggregated_importance <- feature_importance_df %>%
  group_by(Taxon) %>%
  summarise(
    MeanIncNodePurity = mean(IncNodePurity, na.rm = TRUE),
    SDIncNodePurity = sd(IncNodePurity, na.rm = TRUE),
    MeanPercIncMSE = mean(`%IncMSE`, na.rm = TRUE),
    SDPercIncMSE = sd(`%IncMSE`, na.rm = TRUE)
  ) %>%
  arrange(desc(MeanPercIncMSE))
#write.csv(aggregated_importance, "260225_GM_Dev_RandomForests_aggregated_importance.csv", row.names = F)

# Combine all rfcv results
rfcv_results_df <- bind_rows(rfcv_results_list)

# Summarize rfcv results across iterations
aggregated_rfcv <- rfcv_results_df %>%
  group_by(n_features) %>%
  summarise(
    MeanError = mean(error, na.rm = TRUE),
    SDError = sd(error, na.rm = TRUE)
  )
write.csv(aggregated_rfcv, "260722_GM_Dev_RandomForests_aggregated_rfcv.csv", row.names = F)

# Plot aggregated RFCV results
rfcv_error_variables_plot <- ggplot(aggregated_rfcv, aes(x = n_features, y = MeanError)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = MeanError - SDError, ymax = MeanError + SDError), width = 5) +
  theme_minimal() +
  labs(
    x = "Number of Variables",
    y = "Cross-validation error",
    title = "RFCV Error Across Variable Subsets"
  )
ggsave("260722_GMdev_RF_RFCV_aggregated_results_plot.pdf", plot = rfcv_error_variables_plot, bg = "transparent", width = 6, height = 6)

# Based on rfcv results, minimal improvement in predictive performance was observed when including taxa beyond the top N
top_taxa_n <- 50
top_taxa_rf <- aggregated_importance %>%
  arrange(desc(MeanPercIncMSE)) %>%
  dplyr::slice(1:top_taxa_n) %>%
  pull(Taxon)  # Select top N taxa based on mean importance

top50 <- aggregated_importance %>% arrange(desc(MeanPercIncMSE)) %>% slice_head(n=50) %>% 
  mutate(Taxon = str_remove(Taxon, "^s__"),
         Taxon = str_replace_all(Taxon, "_", " "))
write.csv(top50, "260722_GM_Dev_RandomForests_top50_aggregated_importance.csv", row.names = F)

## Plot the top 50 taxa by %IncMSE
top50_taxa_plot <- ggplot(top50, aes(x = reorder(Taxon, MeanPercIncMSE), y = MeanPercIncMSE)) +
  geom_point() +
  geom_errorbar(aes(ymin = MeanPercIncMSE - SDPercIncMSE, ymax = MeanPercIncMSE + SDPercIncMSE), width = 0.3) +
  coord_flip() +
  labs(
    x = "",
    y = "Increase in Mean Squared Error (%)",
    title = "Top 50 Age-Discriminatory Taxa"
  ) +
  theme_minimal()
ggsave("260722_GMdev_RF_top50_taxa_by_MSE_plot.pdf", plot = top50_taxa_plot, bg = "transparent", width = 6, height = 8)

# Step 3: Train sparse model on training set and validate on validation set
x_train_sparse <- x_train[, top_taxa_rf, drop = FALSE]

sparse_rf_model <- randomForest(
  x = x_train_sparse,
  y = y_train,
  ntree = 10000,
  importance = TRUE
)

#############-------------Apply sparse model to Early only and Early and After infants--------------#################
ps_early_only <- ps_filt_dol60_renorm %>% subset_samples(abx_group_names == "EEO") # 801 samples; 171 infants
ps_early_and_after <- ps_filt_dol60_renorm %>% subset_samples(abx_group_names == "EAS") # 1312 samples; 183 infants
ps_abxN <- ps_filt_dol60_renorm %>% subset_samples(abx_group_names == "ABN") # 282 samples; 38 infants
ps_abxN_val <- ps_abxN %>% subset_samples(Sample_ID_R_final %in% validation_data$Sample) # 113 samples; 35 infants 

predict_early_only <- predict_microbiota_age(ps_early_only, sparse_rf_model, top_taxa_rf)
predict_early_and_after <- predict_microbiota_age(ps_early_and_after, sparse_rf_model, top_taxa_rf)
predict_abxN_val <- predict_microbiota_age(ps_abxN_val, sparse_rf_model, top_taxa_rf)

predict_early_only <- data.frame(predict_early_only)
predict_early_and_after <- data.frame(predict_early_and_after)
predict_abxN_val <- data.frame(predict_abxN_val)

# Combine all datasets into one
combined_df <- bind_rows(
  predict_early_only %>% mutate(Group = "EEO"),
  predict_early_and_after %>% mutate(Group = "EAS"),
  predict_abxN_val %>% mutate(Group = "ABN"))
#write.csv(combined_df, "260225_GM_Dev_RF_predicted_microbiota_ages.csv", row.names = F)

rf_plot <- ggplot(combined_df, aes(x = DOL, y = Predicted_Microbiota_Age, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  labs(
    x = "Chronological age (days)",
    y = "Microbiota age (days)",
    title = "Microbiota Maturation Across Groups"
  ) +
  theme_minimal() +
  scale_color_manual(values = custom_colors_names) + 
  theme(
    axis.title = element_text(size = 16, face = "bold"),      
    axis.text = element_text(size = 14),                      
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )
ggsave("260722_GMdev_RF_abxN_model_validation.pdf", plot = rf_plot, bg = "transparent", width = 10, height = 10)


## Determine how much older the ABN infants are in terms of predicted microbiota age
#### Perform lmer follower by emmeans within each DOL bin #####
lmer_emmeans_pred_age_adj <- perform_emmeans_lmer(df = combined_df, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Predicted_Microbiota_Age",
                                         fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status", "Maternal_chorioamnionitis", "Maternal_preeclampsia", "mat_inpt_antimicrobe",
                                                                "frac_days_formula", "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_pred_age_adj$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_predicted_microbiota_age_adj.csv", row.names = F)

## Plot MAZ barplot binned by DOL 
combined_df$abx_group_names <- factor(combined_df$abx_group_names, levels = c("EAS", "EEO", "ABN"))

pred_micr_age_abx_dol <- ggplot(combined_df, aes(x = abx_group_names, y = Predicted_Microbiota_Age, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Predicted Microbiota Age (days)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

# Add emmeans results to plot
pred_micr_age_abx_dol_emm_lmer <- pred_micr_age_abx_dol + stat_pvalue_manual(
  lmer_emmeans_pred_age_adj$stat.test,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin")
ggsave("260630_noBSI_v2/260807_P4_GM_Dev_predicted_microbiota_age_abx_group_DOL_bp_emmeans_ADJ.pdf", plot = pred_micr_age_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)


pred_micr_age_model_adj <- lmer(Predicted_Microbiota_Age ~ abx_group_names * DOL_bin + Gestational_age_weeks + Delivery_route + Sex + Maternal_chorioamnionitis + Maternal_preeclampsia + mat_inpt_antimicrobe + frac_days_formula + frac_days_milk + (1|Patient), data = combined_df)
summary(pred_micr_age_model_adj)

## Obtain contrasts averaged across all DOL windows 
pred_micr_age_emm_overall <- emmeans(pred_micr_age_model_adj, ~ abx_group_names)
contrasts_overall <- pairs(pred_micr_age_emm_overall)

contrasts_overall_full <- summary(contrasts_overall, infer = c(TRUE, TRUE))

write.csv(as.data.frame(contrasts_overall_full),
          "260811_noBSI_mat_inpt_antimicrobe/260811_maintext_pred_micr_age_composite_contrasts.csv", row.names = FALSE)

####----------- Revision --------------------#################
########## Calculating and comparing MAZ values ##########
# Bin DOL into age bins (month-based)
combined_df2 <- combined_df %>%
  mutate(
    Month = case_when(
      DOL <= 30 ~ "Month 1",
      DOL <= 60 ~ "Month 2",
      TRUE ~ NA_character_
    )
  )

# Merge stats back and calculate MAZ
combined_df2 <- combined_df2 %>%
  left_join(reference_stats, by = "DOL_bin") %>%
  mutate(MAZ = (Predicted_Microbiota_Age - mu_ref) / sigma_ref)

combined_df2$abx_group_names <- factor(combined_df2$abx_group_names, levels = c("EAS", "EEO", "ABN"))

#### Perform lmer follower by emmeans within each DOL bin #####
lmer_emmeans_maz_adj <- perform_emmeans_lmer(df = combined_df2, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "MAZ",
                                         fixed_effects = c("Gestational_age_weeks", "Delivery_route", "Sex",
                                                                "SIP_status", "Maternal_chorioamnionitis", "Maternal_preeclampsia", "mat_inpt_antimicrobe",
                                                                "frac_days_formula", "frac_days_milk"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_maz_adj$stat.test, "260811_noBSI_mat_inpt_antimicrobe/260811_lmer_emmeans_MAZ_adj.csv", row.names = F)


## Plot MAZ barplot binned by DOL 
maz_abx_dol <- ggplot(combined_df2, aes(x = abx_group_names, y = MAZ, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "MAZ (Microbiota-for-Age Z-score)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

# Add emmeans results to plot
maz_abx_dol_emm_lmer <- maz_abx_dol + stat_pvalue_manual(
  lmer_emmeans_maz_adj$stat.test,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin")
ggsave("260630_noBSI_v2/260722_P4_GM_Dev_MAZ_abx_group_DOL_bp_emmeans_ADJ.pdf", plot = maz_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)


save.image("260722_GM_dev_random_forests.RData")
```

Creating supplementary table 1
```{r}
rf_columns <- combined_df2 %>% select(Sample_ID_R_final, Predicted_Microbiota_Age, MAZ)
supp_metadata_table <- merge(arg_merged_meta, rf_columns, by = "Sample_ID_R_final", all.x = T)
write.csv(supp_metadata_table, "260816_supplementarytable1_metadata.csv", row.names = F)
```


Prevalence of top taxa over time (Related to Fig. 3A)
```{r}
otu_df <- as.data.frame(otu_table(merged_ps_filt))
## Extract only species names from the taxa names
rownames(otu_df) <- sapply(str_split(rownames(otu_df), "\\|"), function(x) if(length(x) >= 7) x[7] else NA)
otu_df <- t(otu_df)
otu_df <- data.frame(otu_df)
otu_df <- otu_df %>% rownames_to_column("Sample_ID_R_final")

## Get taxonomy at desired level (genus, species)
tax_df <- as.data.frame(tax_table(merged_ps_filt)) 
rownames(tax_df) <- sapply(str_split(rownames(tax_df), "\\|"), function(x) if(length(x) >= 7) x[7] else NA)
tax_df <- tax_df %>% rownames_to_column("TaxaID")

## Choose desired taxonomic rank
tax_level <- "Genus"

## Map TaxaID to taxonomic name (e.g., genus)
tax_map <- tax_df %>% select(TaxaID, !!sym(tax_level)) %>%
  mutate(Taxon = if_else(is.na(!!sym(tax_level)), TaxaID,!!sym(tax_level)))

## Reshape to long format
otu_long <- otu_df %>% pivot_longer(-Sample_ID_R_final, names_to = "TaxaID", values_to = "Abundance") %>%
  left_join(tax_map, by = "TaxaID")

## Add prevalence column 
otu_long <- otu_long %>% mutate(Prevalence = as.integer(Abundance > 0))

## Merge with sample metadata
meta_df <- as.data.frame(sample_data(merged_ps_filt))

otu_long <- otu_long %>% left_join(meta_df, by = "Sample_ID_R_final")
otu_long_earlyOnly <- otu_long %>% filter(abx_group_names == "EEO")
otu_long_earlyAfter <- otu_long %>% filter(abx_group_names == "EAS")
otu_long_naive <- otu_long %>% filter(abx_group_names == "ABN")

custom_colors_prev <- c("g__Clostridioides" = "#E69F00", "g__Clostridium" = "#F0E442", "g__Escherichia" = "#56B4E9",
                                          "g__Enterobacter" = "#0072B2", "g__Klebsiella" = "#009E73", "g__Cutibacterium" = "#D55E00",
                                          "g__Veillonella" = "#CC79A7", "g__Staphylococcus" = "#2B2D42", "g__Enterococcus" = "#FC8D62",
                                          "g__Finegoldia" = "#8DA0CB")
custom_colors_top6 <- c("g__Escherichia" = "#9F7E59","g__Enterobacter" = "#8C2B0E", "g__Klebsiella" = "#2F4B7A", "g__Staphylococcus" = "#B47E83","g__Finegoldia" = "#8DA0CB","g__Enterococcus" = "#F2A450")

## Select top taxa based on average prevalence across all timepoints
top_taxa <- otu_long %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 

prevalence <- otu_long %>%
  group_by(Taxon) %>%

per_group_prev <- otu_long %>%
  group_by(abx_group_names, Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100, 
            AvgAbund = mean(Abundance),
            number_of_stools = n(), .groups = "drop") %>%
  group_by(abx_group_names) %>%
  slice_max(AvgPrev, n = 10) %>%
  ungroup()
write.csv(per_group_prev, "260722_top_taxa_prevalence_per_abx_group.csv", row.names = F)

per_group_abund <- otu_long %>%
  group_by(abx_group_names, Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100, 
            AvgAbund = mean(Abundance),
            number_of_stools = n(), .groups = "drop") %>%
  group_by(abx_group_names) %>%
  slice_max(AvgAbund, n = 10) %>%
  ungroup()

# Get the union of top taxa across all groups
top_taxa_union <- per_group_prev %>%
  distinct(Taxon)   

top_taxa_union_abund <- per_group_abund %>%
  distinct(Taxon) 

per_group_prev_by_daybin <- otu_long %>%
  group_by(abx_group_names, DOL_bin, Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100, 
            AvgAbund = mean(Abundance),
            number_of_stools = n(), .groups = "drop") %>%
  ungroup() %>%
  arrange(abx_group_names, DOL_bin)
write.csv(per_group_prev_by_daybin, "260722_top_taxa_prevalence_by_DOL.csv", row.names = F)


## Select top 5 for plotting
top_taxa <- top_taxa_union %>% slice_head(n = 5) 

## prevalence summary over time 
prevalence_summary <- otu_long %>% filter(Taxon %in% top_taxa$Taxon) %>% group_by(Taxon, DOL, abx_group_names) %>% summarise(Prevalence = mean(Prevalence) * 100, .groups = "drop") %>% 
  mutate(Taxon = factor(Taxon, levels = top_taxa$Taxon))


## Create a custom color palette for the top 5 prevalent taxa
custom_colors_top5 <- c("g__Escherichia" = "#9F7E59","g__Enterobacter" = "#8C2B0E", "g__Klebsiella" = "#2F4B7A", "g__Enterococcus" = "#F2A450", "g__Staphylococcus" = "#B47E83")


prevalence_DOL_plot_facet <- ggplot(prevalence_summary, aes(x = DOL, y = Prevalence, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 1.2) +
  facet_wrap(~ abx_group_names, nrow = 1) +
  scale_color_manual(values = custom_colors_top5) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Postnatal age [days]",
    y = "Presence of taxon within each postnatal age bin [% of cohort]",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )
ggsave("260722_GM_Dev_top5_prevalent_taxa_DOL_faceted.pdf", plot = prevalence_DOL_plot_facet, bg = "transparent", width = 12, height = 8)
```

Stacked barplots at the genus level for preterm and near-term infants (Related to Supplementary Fig. 4)
```{r}
## Barplot of taxonomic composition (Genus level) by Enterotype #######
# Step 1: Collapse to Genus level and transform to relative abundance
ps_genus <- tax_glom(ps_infants, taxrank = "Genus")
ps_relabun <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Step 2: Extract OTU table
otu_df <- as.data.frame(otu_table(ps_relabun))
if (taxa_are_rows(ps_relabun)) {
  otu_df <- t(otu_df)
}
otu_df <- as.data.frame(otu_df) %>%
  tibble::rownames_to_column("Sample_ID_R_final")

# Step 3: Pivot to long format
otu_long <- otu_df %>%
  pivot_longer(-Sample_ID_R_final, names_to = "OTU", values_to = "Abundance")

# Step 4: Merge sample metadata
sample_df <- data.frame(sample_data(ps_relabun)) 
otu_long <- left_join(otu_long, sample_df, by = "Sample_ID_R_final")

# Step 5: Add Genus information from tax_table
tax_df <- as.data.frame(tax_table(ps_relabun)) %>%
  tibble::rownames_to_column("OTU")

otu_long <- left_join(otu_long, tax_df[, c("OTU", "Genus")], by = "OTU")

# Step 6: Summarize mean relative abundance
otu_summary <- otu_long %>%
  filter(!is.na(Genus), !is.na(abx_group_names)) %>%
  group_by(abx_group_names, Genus, DOL_bin) %>%
  summarise(Mean_RelativeAbund = mean(Abundance), .groups = "drop")

# Step 7: Identify top 15 genera across all enterotypes
top_genera <- otu_summary %>%
  group_by(Genus) %>%
  summarise(Total = sum(Mean_RelativeAbund)) %>%
  arrange(desc(Total)) %>%
  head(15) %>%
  pull(Genus)

# Step 8: Group others into "Other"
otu_summary <- otu_summary %>%
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other")) %>%
  group_by(abx_group_names, Genus, DOL_bin) %>%
  summarise(Mean_RelativeAbund = sum(Mean_RelativeAbund), .groups = "drop")


### Create a color-palette with genera belonging to the same families/order with similar hues
custom_genus_colors <- c(

  # --- Bacteroidota: blue-greens ---
  "g__Bacteroides"      = "#088096",
  "g__Parabacteroides"  = "#7AD4E4",

  # --- Actinobacteriota: warm teals ---
  "g__Bifidobacterium"  = "#9AC8A4",
  "g__Cutibacterium"    = "#6EA57B",

  # --- Firmicutes Clostridia: orange-browns ---
  "g__Clostridium"        = "#B07159",
  "g__Erysipelatoclostridium" = "#6C3322",

  # --- Firmicutes Bacilli: pink-reds ---
  "g__Enterococcus"     = "#F2A450",
  "g__Staphylococcus"   = "#B47E83",
  "g__Streptococcus"    = "#68434E",

  # --- Proteobacteria: blue-purple tones ---
  "g__Citrobacter"      = "#DFCB91",
  "g__Enterobacter"     = "#8C2B0E",
  "g__Escherichia"      = "#9F7E59",
  "g__Klebsiella"       = "#2F4B7A",
  "g__Serratia"         = "#7C97C2",

  # --- Negativicutes (Veillonellaceae): olive-greens ---
  "g__Veillonella"      = "#A8A55C",
    # --- Other: neutral gray ---
  "Other"               = "#A9A9A9"   # darkgray
)

genus_order <- c(
  # Bacteroidota
  "g__Bacteroides",
  "g__Parabacteroides",

  # Actinobacteriota
  "g__Bifidobacterium",
  "g__Cutibacterium",

  # Firmicutes: Clostridia
  "g__Clostridium",
  "g__Erysipelatoclostridium",

  # Firmicutes: Bacilli
  "g__Enterococcus",
  "g__Staphylococcus",
  "g__Streptococcus",

  # Proteobacteria (Enterobacterales)
  "g__Citrobacter",
  "g__Enterobacter",
  "g__Escherichia",
  "g__Klebsiella",
  "g__Serratia",

  # Negativicutes
  "g__Veillonella",

  # Other
  "Other"
)
otu_summary$Genus <- factor(otu_summary$Genus, levels = genus_order)

# Step 9: Plot
abx_group_DOL_bin_genus_bp <- ggplot(otu_summary, aes(x = abx_group_names, y = Mean_RelativeAbund, fill = Genus)) +
  geom_bar(stat = "identity", width = 0.85, color = "black") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = custom_genus_colors) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 16, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  labs(
    title = "Average Genus-Level Composition by Enterotype",
    x = NULL, y = "Relative Abundance"
  )
ggsave("260722_P4_GM_Dev_near-term_barplot_genus.pdf", plot = abx_group_DOL_bin_genus_bp, bg = "transparent", width = 20, height = 8)
```


Effect of receipt of breastmilk (Revision)
```{r}
## prevalence summary over time 
bmilk_genera <- c("g__Bifidobacterium", "g__Lactobacillus", "g__Enterococcus", "g__Klebsiella")

bmilk_summary <- otu_long %>% filter(Taxon %in% bmilk_genera) %>% group_by(Taxon, cum_days_milk) %>% summarise(Abundance = mean(Abundance), Prevalence = mean(Prevalence) * 100, .groups = "drop") %>% mutate(Taxon = factor(Taxon, levels = bmilk_genera)) %>% filter(!is.na(cum_days_milk)) %>%
  arrange(desc(cum_days_milk))

bmilk_prev_summary <- otu_long %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>% filter(Taxon %in% bmilk_genera) %>% arrange(desc(AvgPrev))




## Create a custom color palette for the top 5 prevalent taxa
bmilk_colors <- c("g__Bifidobacterium"  = "#9AC8A4", "g__Lactobacillus" = "#088096",  "g__Enterococcus"     = "#F2A450", "g__Klebsiella"       = "#2F4B7A")


bmilk_cumdaysmilk_abundance <- ggplot(bmilk_summary, aes(x = cum_days_milk, y = Abundance, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 1.2) +
  scale_color_manual(values = bmilk_colors) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(
    x = "Cumulative days exposed to breastmilk",
    y = "Average abundance of taxon (%)",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 16),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.position = "top"
  )
ggsave("260810_bmilk_abundance.pdf", plot = bmilk_cumdaysmilk_abundance, bg = "transparent", width = 10, height = 8)


bmilk_cumdaysmilk_prevalence <- ggplot(bmilk_summary, aes(x = cum_days_milk, y = Prevalence, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 1.2) +
  scale_color_manual(values = bmilk_colors) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Cumulative days exposed to breastmilk",
    y = "Average presence of taxon (% of cohort)",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 16),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.position = "top"
  )
ggsave("260810_bmilk_prevalence.pdf", plot = bmilk_cumdaysmilk_prevalence, bg = "transparent", width = 10, height = 8)
```


