---
title: "P4_Metagenomics_analyses_Final"
output: html_document
date: "2026-01-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)          #3.5.0
library(ggExtra)          #0.10.1
library(dplyr)            #1.1.2
library(phyloseq)         #1.46.0
library(vegan)
library(Maaslin2)         #1.12.0
library(ape)              #5.7.1
library(readxl)           #1.4.3     
library(microViz)         #0.12.0
library(ggside)           #0.3.1
library(patchwork)        #1.2
library(ComplexHeatmap)   #2.14.0
library(tidyr)            #1.3.1
library(readr)            #2.1.5
library(reshape2)         #1.4.4
library(rstatix)          #0.7.2
library(ggpubr)           #0.6.0
library(NatParksPalettes) #0.2.0
library(cowplot)          #1.1.2
library(stringr)          #1.5.1
library(Boruta)           #8.0.0 
library(caret)            #6.0-94
library(VIM)              #6.2.2 
library(corpcor)
library(corrplot)
library(Hmisc)
library(ANCOMBC)          #2_4.0   
library(MTXmodel)         #1.2.4
library(forcats)          #1.0.0
library(tidyverse)        #2.0.0  
library(scales)           #1.3.0
library(microbiome)
library(lme4)
library(broom)
library(MuMIn)
library(emmeans)
library(patchwork)
library(DirichletMultinomial)
library(lattice)
library(xtable)
library(parallel)
library(RColorBrewer)
library(e1071)
library(doParallel)
library(kableExtra)
library("anytime") 
library(tibble) # for rownames_to_column() function
library(stringr) # for str_squish()
library(flextable)
library(officer)
library(forcats)
library(tableone)
library(webshot2)
library(randomForest)


source("GM_development_functions.R")
source("~/Documents/GitHub/Preterm_Plasmidome/GM_development/customtab.R")

meta_data_path <- "~/Documents/GitHub/Preterm_Plasmidome/GM_development/260202_P4_GM_dev_meta_updated_groupings_v2.xlsx"

cat_var <- c("Sample_ID", "Patient", "Cohort", "abx_group", "BSI_exp_type", "Necrotizing_enterocolitis", "Received_antimicrobials", "Sequencing_runs","duplicate_count","Cohorts","Sex","Delivery_route","Maternal_ethnicity","Maternal_race","Site","Final_disposition",
             "Para", "Maternal_smoking",	"Maternal_alcohol",	"Maternal_substance_abuse",	"Maternal_preeclampsia",	"Maternal_diabetes",	"Maternal_chorioamnionitis","Maternal_hypertension", "mat_inpt_antimicrobe","mat_outpt_antimicrobial","Multiple_births", "PMA_bin")

num_var <- c("DOL","MOL","abx_exposure_count","Gestational_age_weeks", "Birthweight","Maternal_age_at_delivery","Gravida","antiviral_exposure", "aminoglycoside_exposure","penicillin_exposure","macrolide_exposure","firstgencephalosporin_exposure",
             "fourthgencephalosporin_exposure","thirdgencephalosporin_exposure","secondgencephalosporin_exposure","lincosamide_exposure",
             "antifungal_exposure","carbapenem_exposure","other_antimicrobials_exposure","other_non_antimicrobials_exposure","glycopeptide_exposure","med_exposure_count", "morphine_exposure", "fentanyl_exposure", "surfactant_exposure", "acetaminophen_exposure", "caffeine_exposure", "cholecalciferol_vit_d__exposure",	'vitamin_a_exposure',	'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	'bumetanide_bumax__exposure',	'other_exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	'sodium_bicarbonate_exposure','hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure','cum_days_formula','cum_days_milk','frac_days_formula', 'frac_days_milk')

custom_colors_abx <- c("None" = "#8E949F", "Early only" = "#44637D", "After only" = "#613921", "Early and after" = "#AAC9ED")
custom_colors_groups <- c("None" = "#8E949F", "Early only" = "#44637D", "Early and after" = "#AAC9ED", "Near-term" = "#613921")
custom_colors_names <- c("ABN" = "#8E949F", "EEO" = "#44637D", "EAS" = "#AAC9ED", "Near-term" = "#613921")
```

# Load data
```{r}
## Metaphlan3 tree used in Aura's paper ###
m3.tree.Jan19 <- ape::read.tree(paste0(getwd(),"/datasets/mpa_v30_CHOCOPhlAn_201901_species_tree.nwk.txt"))
m3.tree.Jan19$tip.label <- gsub('GCA_[0-9]+\\|', '', m3.tree.Jan19$tip.label)

# Read the metaphlan3 table
mphlan <- data.frame(read.table("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metaphlan/250318_P4_GM_dev_metaphlan3_merged_n2585_v2.txt", header = T, sep="\t", skip=1, check.names = FALSE))

# R adds an "X" to column names that start with a number. Steps to remove this below
#colnames(mphlan) <- gsub("^X([0-9])", "\\1", colnames(mphlan))
colnames(mphlan) <- as.character(colnames(mphlan))

metadata <- read_excel(meta_data_path)

## Processing data for phyloseq
rownames(mphlan) <- mphlan$clade_name
mphlan$clade_name <- NULL

## Create species data for phyloseq 
# keep rows with species info 
#mphlan.species <- mphlan
mphlan.species <- mphlan[grepl('s__', rownames(mphlan)), ]
# Drop rows with SBG (species level genome bins)
mphlan.species <- mphlan.species[!grepl('t__', rownames(mphlan.species)), ]
mphlan.species <- mphlan.species %>% dplyr::select_if(~sum(.) != 0)
mphlan.species <- as.matrix(mphlan.species)

## Create taxa table
species <- data.frame(Names = rownames(mphlan.species))
species <- data.frame(do.call('rbind', strsplit(as.character(species$Names), '|', fixed = TRUE)))
rownames(species) <- rownames(mphlan.species)
colnames(species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
species <- as.matrix(species)

## Metadata
rows_metadata <- rownames(metadata)[colnames(mphlan.species) %in% metadata$Sample_ID_R_final]
metadata <- data.frame(metadata)
rownames(metadata) <- metadata$Sample_ID_R_final

# Set categorical variables to factor
metadata[,cat_var] <- lapply(metadata[,cat_var], factor)
# Set numerical variables to numeric
metadata[,num_var] <- lapply(metadata[,num_var], as.numeric)
print(dim(metadata))

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(mphlan.species), rownames(metadata)))

ps <- phyloseq(otu_table(mphlan.species, taxa_are_rows = TRUE),
               sample_data(metadata),
               tax_table(species), 
               phy_tree(m3.tree.Jan19)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps.filt <- filter_taxa(ps, function(x) mean(x) > 0.01, TRUE)
ps.filt  <- transform_sample_counts(ps.filt, function(x) (x / sum(x))*100)

### Filter to DOL 0-60 and remove After only infants ########
ps_filt_dol60 <- ps.filt %>% subset_samples(DOL <= 60) %>% subset_samples(abx_group_updated != "After only")
ps_filt_dol60_meta <- data.frame(sample_data(ps_filt_dol60))

sample_data(ps_filt_dol60)$DOL_bin <- factor(sample_data(ps_filt_dol60)$DOL_bin)
sample_data(ps_filt_dol60)$abx_group_updated <- factor(sample_data(ps_filt_dol60)$abx_group_updated,
                                               levels = c("Early and after","Early only","None"))

###################### Merge with P4 cross-sectional cohort ###########
p4_mphlan <- data.frame(read.table("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metaphlan/251020_P4_merged_metaphlan_n146.txt", header = T, sep="\t", check.names = FALSE))

## Processing data for phyloseq
rownames(p4_mphlan) <- p4_mphlan$clade_name
p4_mphlan$clade_name <- NULL

new_names <- colnames(p4_mphlan)
new_names <- gsub("^X", "", new_names)
new_names <- gsub("_", ".", new_names)
new_names <- paste0("p4_", new_names)

colnames(p4_mphlan) <- new_names

## Create species data for phyloseq 
# keep rows with species info 
p4_mphlan.species <- p4_mphlan[grepl('s__', rownames(p4_mphlan)), ]
# Drop rows with SBG (species level genome bins)
p4_mphlan.species <- p4_mphlan.species[!grepl('t__', rownames(p4_mphlan.species)), ]
p4_mphlan.species <- p4_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
p4_mphlan.species <- as.matrix(p4_mphlan.species)

## Create taxa table
p4_species <- data.frame(Names = rownames(p4_mphlan.species))
p4_species <- data.frame(do.call('rbind', strsplit(as.character(p4_species$Names), '|', fixed = TRUE)))
rownames(p4_species) <- rownames(p4_mphlan.species)
colnames(p4_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
p4_species <- as.matrix(p4_species)

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(p4_mphlan.species), rownames(metadata)))

ps_p4 <- phyloseq(otu_table(p4_mphlan.species, taxa_are_rows = TRUE),
               sample_data(metadata),
               tax_table(p4_species)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps_p4.filt <- filter_taxa(ps_p4, function(x) mean(x) > 0.01, TRUE)
ps_p4.filt  <- transform_sample_counts(ps_p4.filt, function(x) (x / sum(x))*100)

ps_p4_meta <- data.frame(phyloseq::sample_data(ps_p4.filt))

ps_p4.filt <- ps_p4.filt %>% subset_samples(abx_group_updated != "After only")
ps_p4.filt_meta <- data.frame(sample_data(ps_p4.filt))

sample_data(ps_p4.filt)$DOL_bin <- factor(sample_data(ps_p4.filt)$DOL_bin)
sample_data(ps_p4.filt)$abx_group_updated <- factor(sample_data(ps_p4.filt)$abx_group_updated,
                                               levels = c("Early and after","Early only","None"))

merged_ps <- merge_phyloseq(ps_filt_dol60, ps_p4.filt)
merged_ps_meta <- data.frame(sample_data(merged_ps))

merged_ps_meta <- merged_ps_meta %>% mutate(abx_group_names = case_when(abx_group_updated == "Early only" ~ "EEO",
                                                                          abx_group_updated == "Early and after" ~ "EAS",
                                                                          abx_group_updated == "None" ~ "ABN"))

merged_ps_meta$abx_group_names <- factor(merged_ps_meta$abx_group_names, levels = c("EAS", "EEO", "ABN"))
sample_data(merged_ps) <- merged_ps_meta

#save(merged_ps, file = "260202_merged_ps.RData")

#write.csv(merged_ps_meta, "260130_P4_GM_dev_metadata.csv", row.names = F)

```

### Load near-term data
```{r}
drewg_meta <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metadata/240515_drewg_sample_metadata_v1.xlsx")

drewg_meta <- drewg_meta %>% filter(Cohort == "Gasparrini_term") 
drewg_meta <- drewg_meta[,1:7]

extra_meta <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metadata/Drewg_term_metadata.xlsx") %>%
  filter(Group == "Term") %>% select(Sample, GestationalAgeWeeks, Birthweight,
                                     Gender, DeliveryMode, TotalAntimicrobialsDays)

extra_meta <- extra_meta %>% dplyr::rename(Sample_ID = Sample, 
                                    Gestational_age_weeks = GestationalAgeWeeks,
                                    Sex = Gender,
                                    Delivery_route = DeliveryMode,
                                    abx_exposure_count = TotalAntimicrobialsDays)

drewg_meta_final <- merge(drewg_meta, extra_meta, by = "Sample_ID")

drewg_meta_dol60 <- drewg_meta_final %>% filter(DOL <= 60)

drewg_mphln <- mphlan <- data.frame(read.table("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metaphlan/251016_DrweG_term_n162_metaphlan3_merged.txt", header = T, sep="\t", skip=1, check.names = FALSE))
rownames(drewg_mphln) <- drewg_mphln$clade_name
drewg_mphln$clade_name <- NULL

drewg_mphlan.species <- drewg_mphln[grepl('s__', rownames(drewg_mphln)), ]
# Drop rows with SBG (species level genome bins)
drewg_mphlan.species <- drewg_mphlan.species[!grepl('t__', rownames(drewg_mphlan.species)), ]
drewg_mphlan.species <- drewg_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
drewg_mphlan.species <- as.matrix(drewg_mphlan.species)

## Create taxa table
drewg_species <- data.frame(Names = rownames(drewg_mphlan.species))
drewg_species <- data.frame(do.call('rbind', strsplit(as.character(drewg_species$Names), '|', fixed = TRUE)))
rownames(drewg_species) <- rownames(drewg_mphlan.species)
colnames(drewg_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
drewg_species <- as.matrix(drewg_species)

## Metadata
rows_metadata <- rownames(drewg_meta_final)[colnames(drewg_mphlan.species) %in% drewg_meta_final$Sample_ID]
drewg_meta_final <- data.frame(drewg_meta_final)
rownames(drewg_meta_final) <- drewg_meta_final$Sample_ID

drewg_cat_var <- c("Sample_ID", "Patient", "Cohort", "Sex", "Delivery_route")
drewg_num_var <- c("DOL", "DOL_months", "PostmenstrualAgeDays","PostmenstrualAgeWeeks","Gestational_age_weeks","Birthweight",
                   "abx_exposure_count")
# Set categorical variables to factor
drewg_meta_final[,drewg_cat_var] <- lapply(drewg_meta_final[,drewg_cat_var], factor)
# Set numerical variables to numeric
drewg_meta_final[,drewg_num_var] <- lapply(drewg_meta_final[,drewg_num_var], as.numeric)
print(dim(drewg_meta_final))

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(drewg_mphlan.species), rownames(drewg_meta_final)))

# 160 samples and 341 species
ps_term <- phyloseq(otu_table(drewg_mphlan.species, taxa_are_rows = TRUE),
               sample_data(drewg_meta_final),
               tax_table(drewg_species)) 

# 160 samples and 179 species
ps_term <- filter_taxa(ps_term, function(x) mean(x) > 0.01, TRUE)
ps_term  <- transform_sample_counts(ps_term, function(x) (x / sum(x))*100)

### Filter to DOL 0-60 and remove After only infants ########
ps_term_dol100 <- ps_term %>% subset_samples(DOL <= 100)

ps_term_meta <- data.frame(phyloseq::sample_data(ps_term_dol100))

# Filter out samples past DOL 100 and create DOL bin
ps_term_meta <- ps_term_meta %>%
  mutate(DOL_bin = cut(DOL,
                       breaks = seq(0,100, by = 10),
                       labels = paste("DOL", seq(1, 91, by = 10), "-", seq(10, 100, by = 10)),
                       include.lowest = TRUE,
                       right = TRUE))


sample_data(ps_term_dol100) <- ps_term_meta

ps_term_dol60 <- ps_term_dol100 %>% subset_samples(DOL <= 60) 
ps_term_dol60_meta <- data.frame(sample_data(ps_term_dol60))

sample_data(ps_term_dol60)$DOL_bin <- factor(sample_data(ps_term_dol60)$DOL_bin)
sample_data(ps_term_dol60)$abx_group_updated <- "Near-term"

ps_term_dol60_meta <- data.frame(sample_data(ps_term_dol60))

breaks <- seq(23, 48, by = 3)
labels <- paste("PMA", head(breaks, -1), "-",tail(breaks, -1))

ps_term_dol60_meta$PMA_bin <- cut(
  ps_term_dol60_meta$PostmenstrualAgeWeeks,
  breaks = breaks,
  labels = labels,
  right = FALSE,  # left-inclusive, i.e., [23,26)
  include.lowest = TRUE
)


sample_data(ps_term_dol60) <- ps_term_dol60_meta

## Merge preterm and term phyloseq objects #########
ps_term_dol60_meta <- ps_term_dol60_meta %>% dplyr::rename(Sample_ID_R_final = Sample_ID) 
pterm_meta_simple <- merged_ps_meta %>% select(any_of(colnames(ps_term_dol60_meta)))
nterm_meta_simple <- ps_term_dol60_meta %>% select(everything(), -c(DOL_months, PostmenstrualAgeDays)) 
nterm_meta_simple$Sample_ID_R_final <- as.character(nterm_meta_simple$Sample_ID_R_final)
nterm_meta_simple$Delivery_route <- as.character(nterm_meta_simple$Delivery_route)
nterm_meta_simple$Delivery_route[nterm_meta_simple$Delivery_route == "Cesarean"] <- "Cesarean Section"
nterm_meta_simple$Delivery_route <- factor(nterm_meta_simple$Delivery_route)


ps_pterm_simple <- merged_ps
sample_data(ps_pterm_simple) <- pterm_meta_simple

ps_nterm_simple <- ps_term_dol60
sample_data(ps_nterm_simple) <- nterm_meta_simple

ps_infants <- merge_phyloseq(ps_pterm_simple, ps_nterm_simple)
ps_infants_meta <- data.frame(sample_data(ps_infants))
ps_infants_meta$abx_group_updated <- factor(ps_infants_meta$abx_group_updated)
ps_infants_meta <- ps_infants_meta %>% mutate(abx_group_names = case_when(abx_group_updated == "Early only" ~ "EEO",
                                                                          abx_group_updated == "Early and after" ~ "EAS",
                                                                          abx_group_updated == "None" ~ "ABN",
                                                                          abx_group_updated == "Near-term" ~ "Near-term"))
ps_infants_meta$abx_group_names <- factor(ps_infants_meta$abx_group_names, levels = c("EAS", "EEO", "ABN", "Near-term"))
#save(ps_infants, file = "251118_Preterm_near-term_merged_object.RData")

###### Add site to metadata #########
preterm_site <- merged_ps_meta %>% select(Sample_ID_R_final, Site)
ps_infants_meta2 <- ps_infants_meta %>% left_join(preterm_site, by = "Sample_ID_R_final") %>% mutate(Site = if_else(is.na(Site), "Home", Site))
ps_infants_meta2$Site <- factor(ps_infants_meta2$Site)

### Update cohort to study ###
ps_infants_meta2 <- ps_infants_meta2 %>% mutate(Study = case_when(Cohort == "BSI" ~ "Schwartz_et_al",
                                                                  Cohort == "Gasparrini_preterm" ~ "Gasparrini_et_al",
                                                                  Cohort == "Gasparrini_term" ~ "Gasparrini_et_al",
                                                                  Cohort == "Molly" ~ "Gibson_et_al",
                                                                  Cohort == "NEC_control" ~ "Thanert_et_al",
                                                                  Cohort == "No_abx_preterm" ~ "This_study",
                                                                  Cohort == "P4_cross_sectional" ~ "This_study"))
ps_infants_meta2$Study <- factor(ps_infants_meta2$Study)
rownames(ps_infants_meta2) <- ps_infants_meta2$Sample_ID_R_final
sample_data(ps_infants) <- ps_infants_meta2

ps_infants_meta <- data.frame(sample_data(ps_infants))
```

Table 1
```{r}
library(kableExtra)
library("anytime") 
library(tibble) # for rownames_to_column() function
library(stringr) # for str_squish()
library(flextable)
library(officer)
library(forcats)
library(tableone)
library(webshot2)

source("~/Documents/GitHub/Preterm_Plasmidome/GM_development/customtab.R")


meta_pt <- data.frame(sample_data(merged_ps))

# Set categorical variables as factors !!!!!!!!!!! Modify these as needed later !!!!!!!!!!!!!!!!!!!!
categorical_vars <- c('Patient', 'Site','abx_group', 'Sex', 'Delivery_route', 'Necrotizing_enterocolitis', 'Cohort')

meta_pt[,categorical_vars] <- lapply(meta_pt[,categorical_vars], factor)

numerical_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks")
meta_pt[,numerical_vars] <- lapply(meta_pt[,numerical_vars], as.numeric)

## All numerical variables are non-normal so we'll use median (IQR) instead of mean (SD)
normality_results <- lapply(numerical_vars, function(v) {
  p <- shapiro.test(meta_pt[[v]])$p.value
  data.frame(variable = v, p_value = p)
}) %>% bind_rows()


# Generate metadata comparison table 
comp_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks", 'Site','Cohort', 'Sex', 'Delivery_route', 'Necrotizing_enterocolitis')


tab_pt <- CreateTableOne(vars=comp_vars,
                           strata='abx_group_names', 
                           data=meta_pt)
tab_plot_pt <- print(tab_pt, 
  showAllLevels = TRUE,
  quote        = FALSE,
  noSpaces     = TRUE,
  test         = TRUE,
  contDigits   = 1,
  nonnormal    = numerical_vars,  
  printToggle  = FALSE,
  dropEqual    = TRUE,
  explain      = FALSE)

# Convert to dataframe
tab_pt_df <- as.data.frame(tab_plot_pt) %>% rownames_to_column(var = "Variable")

# Set Table header
header <- str_squish(str_remove("Table 1. Infant demographics", "\n"))
# Set Table footer
footer <- str_squish(str_remove("Numbers are either No. or mean values", "\n"))
# Set custom_tab() defaults
customtab_defaults()
# Create the flextable object
flextable_processed_samples <- custom_tab(tab_pt_df, header, footer)

# Save as word .docx
save_as_docx(flextable_processed_samples, path = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/Metadata/Table1.docx",
pr_section = prop_section(page_size = page_size(orient = "landscape"), type = "continuous"))

unique_infants_pt <- meta_pt %>%
  group_by(abx_group_updated, Site) %>%
  summarise(N = n_distinct(Patient), .groups = "drop_last") %>%
  mutate(
    total_site = sum(N),                 # total unique patients at this site
    prop = N / total_site                # proportion per group
  ) %>%
  ungroup()
write.csv(unique_infants_pt, "unique_infants_site_dist.csv", row.names = F)

abxN_meta <- meta_pt %>% filter(abx_group_updated == "None")
write.csv(abxN_meta, "251126_abxN_samples_meta.csv", row.names = F)

# 1. Count number of unique group labels per infant
group_changes <- ps_infants_meta %>%
  group_by(Patient) %>%
  dplyr::summarise(
    groups = n_distinct(abx_group_updated),
    unique_groups = paste(sort(unique(abx_group_updated)), collapse = ", ")
  ) %>%
  ungroup()
write.csv(group_changes, "260202_infants_that_change_abx_groups.csv", row.names = F)

###########-- Including near-term infants in the table ----############
meta_pt <- ps_infants_meta

eeo <- ps_infants_meta %>% filter(abx_group_names == "EEO")
abn <- ps_infants_meta %>% filter(abx_group_names == "ABN")
eas <- ps_infants_meta %>% filter(abx_group_names == "EAS")

meta_pt$Delivery_route[meta_pt$Delivery_route == "Cesarean"] <- "Cesarean Section"
meta_pt$Cohort[meta_pt$Cohort == "BSI_new"] <- "BSI"

# Set categorical variables as factors !!!!!!!!!!! Modify these as needed later !!!!!!!!!!!!!!!!!!!!
categorical_vars <- c('Patient','abx_group_names', 'Sex', 'Delivery_route', 'Cohort')

meta_pt[,categorical_vars] <- lapply(meta_pt[,categorical_vars], factor)
meta_pt$Delivery_route <- factor(meta_pt$Delivery_route, levels = c("Cesarean Section", "Vaginal"))
meta_pt$Cohort <- factor(meta_pt$Cohort, levels = c("Molly", "Gasparrini_preterm", "Gasparrini_term", "BSI", "NEC_control", "No_abx_preterm", "P4_cross_sectional"))

numerical_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks")
meta_pt[,numerical_vars] <- lapply(meta_pt[,numerical_vars], as.numeric)

## All numerical variables are non-normal so we'll use median (IQR) instead of mean (SD)

# Generate metadata comparison table 
comp_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks",'Cohort', 'Sex', 'Delivery_route')


tab_pt <- CreateTableOne(vars=comp_vars,
                           strata='abx_group_names', 
                           data=meta_pt)
tab_plot_pt <- print(tab_pt, 
  showAllLevels = TRUE,
  quote        = FALSE,
  noSpaces     = TRUE,
  test         = TRUE,
  contDigits   = 1,
  nonnormal    = numerical_vars,  
  printToggle  = FALSE,
  dropEqual    = TRUE,
  explain      = FALSE)

# Convert to dataframe
tab_pt_df <- as.data.frame(tab_plot_pt) %>% rownames_to_column(var = "Variable")

# Set Table header
header <- str_squish(str_remove("Table 1. Infant demographics", "\n"))
# Set Table footer
footer <- str_squish(str_remove("Numbers are either No. or mean values", "\n"))
# Set custom_tab() defaults
customtab_defaults()
# Create the flextable object
flextable_processed_samples <- custom_tab(tab_pt_df, header, footer)

# Save as word .docx
save_as_docx(flextable_processed_samples, path = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/Metadata/Table1_near-term.docx",
pr_section = prop_section(page_size = page_size(orient = "landscape"), type = "continuous"))


# 2. Filter infants who have more than one group label
changed_infants <- group_changes %>%
  filter(groups > 1)

# 3. View how many and which infants changed groups
n_changed <- nrow(changed_infants)
print(paste("Number of infants who changed group:", n_changed))
print(changed_infants)

```


## Alpha diversity
```{r}
ps.filt.trans <- transform_sample_counts(ps_infants, function(x) trunc(x*100000))
adiv <- estimate_richness(ps.filt.trans, measures=c('Observed', 'Shannon'))

sample_data(ps_infants)$Richness <- adiv$Observed
sample_data(ps_infants)$Shannon <- adiv$Shannon

ps.filt.meta <- data.frame(phyloseq::sample_data(ps_infants))
#write.csv(ps.filt.meta, "260201_ps_filt_meta.csv", row.names = F)

counts_by_bin <- ps.filt.meta %>%
  filter(!is.na(Richness), !is.na(abx_group_names)) %>%
  mutate(abx_group_names = droplevels(abx_group_names)) %>%
  count(DOL_bin, abx_group_names, name = "n") %>%
  group_by(DOL_bin) %>% mutate(groups_present = n_distinct(abx_group_names)) %>% ungroup()
write.csv(counts_by_bin, "260202_abx_exposure_n_by_group_names.csv", row.names = F)

## There are repeat samples and differences in DOL so we need to use linear models to account for these confounders
patient_repeats_per_DOL_bin <- ps.filt.meta %>%
  group_by(DOL_bin, Patient) %>% summarise(n_samples = n(), .groups = "drop")

########### Richness ##################
### Perform stat test and plot #######
lmer_emmeans_richness <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness, "260202_lmer_emmeans_richness.csv", row.names = F)
## Plot alpha diversity barplot binned by DOL 
richness_abx_dol <- ggplot(ps.filt.meta, aes(x = abx_group_names, y = Richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Species Richness", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )


richness_abx_dol_emm_lmer <- richness_abx_dol + stat_pvalue_manual(
  lmer_emmeans_richness,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Richness_abx_group_DOL_bp_emmeans.pdf", plot = richness_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)


########### Shannon ##################
### Perform stat test and plot #######
lmer_emmeans_shannon <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon, "260202_lmer_emmeans_shannon.csv", row.names = F)

shannon_abx_dol <- ggplot(ps.filt.meta, aes(x = abx_group_names, y = Shannon, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Shannon Diversity", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )


shannon_abx_dol_emm_lmer <- shannon_abx_dol + stat_pvalue_manual(
  lmer_emmeans_shannon,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Shannon_abx_group_DOL_bp_emmeans.pdf", plot = shannon_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)

###### Facet by abx group #######
lmer_emmeans_shannon_DOL <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon_DOL, "260202_lmer_emmeans_shannon_by_DOL.csv", row.names = F)

shannon_abx_dol2 <- ggplot(ps.filt.meta, aes(x = DOL_bin, y = Shannon, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ abx_group_names, nrow = 2, strip.position = "top") +
  labs(
    x = "DOL bin", 
    y = "Shannon Diversity", 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260202_P4_GM_Dev_Shannon_abx_group_facet_DOL_bp.pdf", plot = shannon_abx_dol2, bg = "transparent", width = 20, height = 10)

lmer_emmeans_shannon_DOL <- lmer_emmeans_shannon_DOL %>% separate(contrast, into = c("group1", "group2"), sep = "\\) - \\(") %>%
  mutate(
    group1 = gsub("^\\(|\\)$", "", group1),
    group2 = gsub("^\\(|\\)$", "", group2)
  )

shannon_abx_dol2_emm_lmer <- shannon_abx_dol2 + stat_pvalue_manual(
  lmer_emmeans_shannon_DOL,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.4, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Shannon_abx_group_facet_DOL_bp_emmeans.pdf", plot = shannon_abx_dol2_emm_lmer, bg = "transparent", width = 20, height = 10)


lmer_emmeans_richness_DOL <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness_DOL, "260202_lmer_emmeans_richness_by_DOL.csv", row.names = F)

richness_abx_dol2 <- ggplot(ps.filt.meta, aes(x = DOL_bin, y = Richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ abx_group_names, nrow = 2, strip.position = "top") +
  labs(
    x = "DOL bin", 
    y = "Species Richness", 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260202_P4_GM_Dev_Richness_abx_group_facet_DOL_bp.pdf", plot = richness_abx_dol2, bg = "transparent", width = 20, height = 10)

lmer_emmeans_richness_DOL <- lmer_emmeans_richness_DOL %>% separate(contrast, into = c("group1", "group2"), sep = "\\) - \\(") %>%
  mutate(
    group1 = gsub("^\\(|\\)$", "", group1),
    group2 = gsub("^\\(|\\)$", "", group2)
  )

richness_abx_dol2_emm_lmer <- richness_abx_dol2 + stat_pvalue_manual(
  lmer_emmeans_richness_DOL,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.4, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Richness_abx_group_facet_DOL_bp_emmeans.pdf", plot = richness_abx_dol2_emm_lmer, bg = "transparent", width = 20, height = 10)

###--- Run lmer and emmeans separately on term dataset ######
nterm_meta <- ps.filt.meta %>% filter(abx_group_names == "Near-term")
lmer_emmeans_richness_DOL_nterm <- perform_emmeans_lmer(df = nterm_meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness_DOL_nterm, "260203_lmer_emmeans_richness_by_DOL_nterm.csv", row.names = F)

lmer_emmeans_shannon_DOL_nterm <- perform_emmeans_lmer(df = nterm_meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon_DOL_nterm, "260203_lmer_emmeans_shannon_by_DOL_nterm.csv", row.names = F)

```

ARG Comparison 
```{r}
#look at overall ARG burden between groups
##load shortbred output
arg.full <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/250905_P4_GM_Dev_ShortBRED_merged_output_n2585.txt"))
arg.full <- arg.full %>% mutate(Sample_ID_R_final = gsub("-", ".", paste0("p4_", Sample)))
arg.p4 <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251022_GM_Dev_P4_shortBRED_joint_output.txt"))
arg.p4 <- arg.p4 %>% mutate(Sample_ID_R_final = paste0("p4_", gsub("_", ".", Sample)))
arg_df <- rbind(arg.full, arg.p4)

# starts_with_number <- grepl("^[0-9]", arg.full$Sample)
# modified_column <- arg.full$Sample
# modified_column[starts_with_number] <- paste0("p4_", gsub("-", ".", arg.full$Sample[starts_with_number]))
# arg.full$Sample_ID_R_final <- modified_column

arg.wide <- as.data.frame(dcast(arg_df, Sample_ID_R_final ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg.wide) <- arg.wide$Sample_ID_R_final
arg.wide <- arg.wide[, c(-1)]

##sum counts to get RPKM per sample
arg.wide$RPKM <- rowSums(arg.wide)

##get ARG richness
arg.wide$Richness <- specnumber(arg.wide[, 1:ncol(arg.wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg.wide$Sample_ID_R_final <- row.names(arg.wide)
# Subset arg.wide to columns we want
arg.want <- arg.wide %>%
  select(c(Sample_ID_R_final, RPKM, Richness)) %>% dplyr::rename(ShortBRED_richness = Richness)

# Merge with metadata
arg.df <- merge(ps_infants_meta, arg.want, by = "Sample_ID_R_final")

arg.df <- arg.df %>% filter(abx_group_updated != "After only") 

##### Read in term ARG data ######
arg_term <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251016_DrewG_term_n162_shortbred_merged.txt"))

arg_term_wide <- as.data.frame(dcast(arg_term, Sample ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg_term_wide) <- arg_term_wide$Sample
arg_term_wide <- arg_term_wide[, c(-1)]

##sum counts to get RPKM per sample
arg_term_wide$RPKM <- rowSums(arg_term_wide)

##get ARG richness
arg_term_wide$Richness <- specnumber(arg_term_wide[, 1:ncol(arg_term_wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg_term_wide$Sample <- row.names(arg_term_wide)
# Subset arg.wide to columns we want
arg_term_want <- arg_term_wide %>%
  select(c(Sample, RPKM, Richness)) %>% dplyr::rename(Sample_ID_R_final = Sample, ShortBRED_richness = Richness)

# Merge with metadata
term_arg.df <- merge(ps_infants_meta, arg_term_want, by = "Sample_ID_R_final") 

#### Merge with preterm ARG data ###
arg.df2 <- arg.df %>% 
  select(Sample_ID_R_final, Patient, Study, Site, DOL_bin, Gestational_age_weeks, abx_group_updated, abx_group_names, RPKM, ShortBRED_richness)
term_arg.df2 <- term_arg.df %>% select(Sample_ID_R_final, Patient, Study, Site, DOL_bin, Gestational_age_weeks, abx_group_updated, abx_group_names, RPKM, ShortBRED_richness)

arg_full <- rbind(arg.df2, term_arg.df2)
write.csv(arg_full, "260130_P4_GM_Dev_ARG_data.csv", row.names = F)


###### Plot and compare using linear mixed models ###########
lmer_emmeans_arg_rpkm <- perform_emmeans_lmer(df = arg_full, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "RPKM",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_arg_rpkm, "260202_lmer_emmeans_ARG_rpkm.csv", row.names = F)

counts_by_bin <- arg_full %>%
  filter(!is.na(RPKM), !is.na(abx_group_names)) %>%
  mutate(abx_group_names = droplevels(abx_group_names)) %>%
  count(DOL_bin, abx_group_names, name = "n") %>%
  group_by(DOL_bin) %>% mutate(groups_present = n_distinct(abx_group_names)) %>% ungroup()
write.csv(counts_by_bin, "260202_abx_exposure_n_by_group_shortbred_comparison.csv", row.names = F)

shortbred_hits_rpkm <- ggplot(arg_full, aes(x = abx_group_names, y = RPKM, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Total ShortBRED hits (RPKM)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

shortbred_hits_rpkm_stat  <- shortbred_hits_rpkm + stat_pvalue_manual(
  lmer_emmeans_arg_rpkm,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_ARG_RPKM_abx_group_DOL_bp_emmeans.pdf", plot = shortbred_hits_rpkm_stat, bg = "transparent", width = 20, height = 10)

########### Total ShortBRED Hits (Richness) #########
lmer_emmeans_arg_richness <- perform_emmeans_lmer(df = arg_full, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "ShortBRED_richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_arg_richness, "260202_lmer_emmeans_ARG_richness.csv", row.names = F)

shortbred_hits_richness <- ggplot(arg_full, aes(x = abx_group_names, y = ShortBRED_richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Total ShortBRED hits (Unique ARGs)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

shortbred_hits_richness_stat <- shortbred_hits_richness + stat_pvalue_manual(
  lmer_emmeans_arg_richness,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_ARG_Richness_abx_group_DOL_bp_emmeans.pdf", plot = shortbred_hits_richness_stat, bg = "transparent", width = 20, height = 10)

###################-------------Richness normalized ARGs------------#######################
### Merge ARG data with phyloseq metadata ###
arg_full_want <- arg_full %>% select(Sample_ID_R_final, RPKM, ShortBRED_richness) %>% 
  dplyr::rename(ShortBRED_RPKM = RPKM)
ps.filt.meta2 <- merge(ps.filt.meta, arg_full_want, by = "Sample_ID_R_final")

ps.filt.meta2$norm_arg_abund <- ps.filt.meta2$ShortBRED_RPKM / ps.filt.meta2$Richness
#write.csv(ps.filt.meta2, "260202_P4_GM_Dev_all_infants_meta_alpha_ARGs.csv", row.names = F)

lmer_emmeans_norm_arg_richness <- perform_emmeans_lmer(df = ps.filt.meta2, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "norm_arg_abund",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_norm_arg_richness, "260206_lmer_emmeans_richness_norm_ARG_rpkm.csv", row.names = F)

norm_arg_abund <- ggplot(ps.filt.meta2, aes(x = abx_group_names, y = norm_arg_abund, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Richness Normalized Total ShortBRED Hits (RPKM)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260206_P4_GM_Dev_Richness_Norm_ARG_rpkm_abx_group_.pdf", plot = norm_arg_abund, bg = "transparent", width = 20, height = 10)

norm_arg_abund_stat <- norm_arg_abund + stat_pvalue_manual(
  lmer_emmeans_norm_arg_richness,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)

```


## Beta diversity comparisons
```{r}
######## Plot ordination plot #####
gm_dev_ordination <- phyloseq::ordinate(merged_ps, method='PCoA', distance='bray')

pcoa_gm_dev <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                         type="samples", color= "DOL_bin")+
  geom_point(shape=19, size=3)+
  #scale_color_manual(values = dol_bin_color_vector) + 
  scale_color_manual(values=natparks.pals("Arches"))+ 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev <- ggMarginal(pcoa_gm_dev, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_DOL_bin_P4_joined.pdf", plot = pcoa_gm_dev, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_abx <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                             type="samples", color= "abx_group_updated")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_abx) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev_abx <- ggMarginal(pcoa_gm_dev_abx, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_abx_exposure_group_P4_joined.pdf", plot = pcoa_gm_dev_abx, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_abx_count <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                             type="samples", color= "abx_exposure_count")+
  geom_point(shape=19, size=3)+
  scale_color_gradientn(colors = natparks.pals("Arches")) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 16),
        legend.text = element_text(size=14),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))
ggsave("260202_GM_Dev_PCoA_abx_exposure_count_P4_joined.pdf", plot = pcoa_gm_dev_abx_count, bg = "transparent", width = 9, height = 6)


### Including term infants #####
gm_dev_ordination2 <- phyloseq::ordinate(ps_infants, method='PCoA', distance='bray')

pcoa_gm_dev_abx_term <- phyloseq::plot_ordination(ps_infants, gm_dev_ordination2,
                                             type="samples", color= "abx_group_names")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_names) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev_abx_term <- ggMarginal(pcoa_gm_dev_abx_term, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_abx_exposure_group_tern_joined.pdf", plot = pcoa_gm_dev_abx_term, bg = "transparent", width = 9, height = 6)


## Pairwise comparisons by DOL bin ##
group_levels <- levels(factor(ps_infants_meta$abx_group_names))
group_pairs <- combn(group_levels, 2, simplify = FALSE)

# Loop through each pair
for (pair in group_pairs) {
  # Subset the phyloseq object
  print(pair)
  ps_subset <- subset_samples(ps_infants, abx_group_names %in% pair)
  ps_subset <- prune_samples(sample_sums(ps_subset) > 0, ps_subset)  # avoid empty samples
  ps_subset
  
  # Perform ordination
  ord <- ordinate(ps_subset, method = "PCoA", distance = "bray")
  
  # Base PCoA plot
  p <- plot_ordination(ps_subset, ord, type = "samples", color = "abx_group_names") +
    geom_point(shape = 19, size = 3) +
    stat_ellipse() +
    scale_color_manual(values = custom_colors_names[pair]) +
    theme_classic() +
    theme(
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 16),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      legend.position = "left",
      plot.title = element_text(size = 16, hjust = 0.5),
      panel.border = element_rect(fill = NA, colour = "black", size = 1)
    ) +
    labs(title = paste(pair[1], "vs", pair[2]))
  
  # Add marginal boxplots
  p_with_marginals <- ggMarginal(p, type = "boxplot", groupFill = TRUE, size = 10)
  
  # Save the plot
  file_name <- paste0("PCoA_", pair[1], "_vs_", pair[2], ".pdf")
  ggsave(file.path(file_name), plot = p_with_marginals, bg = "transparent", width = 9, height = 6)
  
  message("Saved: ", file_name)
}


ps_infants1 <- ps_infants %>% subset_samples(DOL_bin == "DOL 1 - 10")
plot_pcoa(ps_obj = ps_infants1, group_var = "abx_group_names", custom_colors = custom_colors_names, 
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants2 <- ps_infants %>% subset_samples(DOL_bin == "DOL 11 - 20")
plot_pcoa(ps_obj = ps_infants2, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants3 <- ps_infants %>% subset_samples(DOL_bin == "DOL 21 - 30")
plot_pcoa(ps_obj = ps_infants3, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants4 <- ps_infants %>% subset_samples(DOL_bin == "DOL 31 - 40")
plot_pcoa(ps_obj = ps_infants4, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants5 <- ps_infants %>% subset_samples(DOL_bin == "DOL 41 - 50")
plot_pcoa(ps_obj = ps_infants5, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants6 <- ps_infants %>% subset_samples(DOL_bin == "DOL 51 - 60")
plot_pcoa(ps_obj = ps_infants6, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")


```

Repeat Measure PERMANOVA Summary Figure
```{r}
DOL <- c("DOL")
baseline_var <- c('Gestational_age_weeks',	'Birthweight','Sex',	'Delivery_route',	'Cohort', 'Necrotizing_enterocolitis', 'Site')
diet_var <- c('cum_days_formula', 'cum_days_milk', 'frac_days_formula','frac_days_milk')
mat_var <- c('Maternal_race','Maternal_age_at_delivery',	'Gravida',	'Para',	'Maternal_smoking',	'Maternal_alcohol',	'Maternal_substance_abuse',	'Maternal_preeclampsia',
'Maternal_diabetes',	'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe',	'mat_outpt_antimicrobial',	'Multiple_births')
abx_var <- c('abx_group_updated','abx_exposure_count','aminoglycoside_exposure',	'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	
'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure','lincosamide_exposure',		'carbapenem_exposure',
	'other_antimicrobials_exposure', 'glycopeptide_exposure','acute_abx_exposure_count', 'acute_aminoglycoside_exposure', 'acute_penicillin_exposure','acute_macrolide_exposure', 'acute_firstgencephalosporin_exposure','acute_fourthgencephalosporin_exposure',
'acute_thirdgencephalosporin_exposure','acute_secondgencephalosporin_exposure','acute_lincosamide_exposure','acute_carbapenem_exposure','acute_other_antimicrobials_exposure',
'acute_other_non_antimicrobials_exposure','acute_glycopeptide_exposure')
medications_var <- c('med_exposure_count', 'other_non_antimicrobials_exposure',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure', 'antifungal_exposure', 'acute_med_exposure_count','acute_morphine_exposure','acute_fentanyl_exposure','acute_surfactant_exposure',
'acute_acetaminophen_exposure','acute_caffeine_exposure','acute_cholecalciferol_vit_d__exposure',
'acute_vitamin_a_exposure','acute_famotidine_pepcid__exposure','acute_iron_exposure',
'acute_calcium_gluconate_exposure','acute_dopamine_exposure','acute_bumetanide_bumax__exposure',
'acute_other_exposure','acute_indomethacin_exposure','acute_multi_vitamin_with_iron_exposure',
'acute_lasix_exposure','acute_atropine_exposure','acute_dobutamine_exposure',
'acute_sodium_chloride_exposure','acute_levothyroxine_exposure','acute_dexamethasone_exposure',
'acute_phenobarbitol_exposure','acute_hydrocortisone_exposure','acute_ibuprofen_exposure',
'acute_chlorathiazide_diuril__exposure','acute_midazolam_versed__exposure',
'acute_potassium_chloride_exposure','acute_citric_acid_sodium_citrate_bicitra__exposure',
'acute_pancuronium_or_vecurionium_exposure','acute_insulin_exposure', "acute_multi_vitamin_no_iron_exposure", "acute_levocarnitine_exposure", "acute_digoxin_exposure", "acute_lorazapam_ativan__exposure", "acute_sodium_bicarbonate_exposure",
'acute_antifungal_exposure', 'acute_antiviral_exposure', 'antiviral_exposure')

results <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/Beta_diversity/260203_GM_Dev_repeat_permanova_results_combined.xlsx")
results$p_adj_BH_redo <- p.adjust(results[["Pr(>F)"]], method = "BH")
results$p_adj_BH_redo <- as.numeric(results$p_adj_BH_redo)
write.csv(results, "260209_Repeat_measures_PERMANOVA_results_p_adj_redo.csv", row.names = F)

results <- results %>%
  mutate(significant = p_adj_BH_redo < 0.05,
         Category = case_when(Variable %in% DOL ~ "DOL",
                              Variable %in% baseline_var ~ "Baseline",
                              Variable %in% diet_var ~ "Diet",
                              Variable %in% mat_var ~ "Maternal",
                              Variable %in% abx_var ~ "Antibiotic",
                              Variable %in% medications_var ~ "Non-antibiotic medication"))
var_order <- c("DOL","Baseline", "Diet", "Maternal","Antibiotic", "Non-antibiotic medication")

results$Category <- factor(results$Category, levels = var_order)

results_plot <- results %>% mutate(R2_percent = R2 * 100,
                                   sig_label = case_when(p_adj_BH_redo <= 0.001 ~ "***",
                                                         p_adj_BH_redo <= 0.01 ~ "**",
                                                         p_adj_BH_redo <= 0.05 ~ "*")) %>%
  filter(p_adj_BH_redo < 0.05) %>%
  arrange(desc(R2_percent)) %>%
  mutate(Label = factor(Label, levels = unique(Label))) %>%
  ungroup()

perm_cat_vector = c("DOL" = "#3A1F46","Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

permanova_sign_var_p <- ggplot(results_plot, aes(x = Label, y = R2_percent, fill = Category)) +
  geom_col(width = 0.8, color = "black") + 
  geom_text(aes(label = sig_label, y = R2_percent + 0.15), size = 5, vjust = 0) +
  scale_fill_manual(values=perm_cat_vector) + 
  labs(x = NULL, y = "Variance explained (%)", fill = NULL) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    axis.line.x = element_line(color = "black", linewidth = 0.6),
    axis.ticks.x = element_line(color = "black", linewidth = 0.4),
    plot.margin = ggplot2::margin(t = 10, r = 10, b = 40, l = 10))  # extra space for category bar
ggsave("260205_GM_Dev_repeat_measure_PERMANOVA_sign_var_acute_combined.pdf", plot = permanova_sign_var_p, bg = "transparent", width = 12, height = 6)
```


Run Linear Mixed Models (MaAslin2) to test association between species and metadata
```{r}
############### ----------------------------- Taxa ----------------------##################################################

repeat_permanova_sign_var <- results_plot$Variable

pterm_infants_otu <- data.frame(otu_table(merged_ps), check.names = FALSE) 
pterm_infants_meta <- data.frame(sample_data(merged_ps), check.names = FALSE)

pterm_infants_meta <- pterm_infants_meta %>% mutate(Study = case_when(Cohort == "BSI" ~ "Schwartz_et_al",
                                                                  Cohort == "Gasparrini_preterm" ~ "Gasparrini_et_al",
                                                                  Cohort == "Gasparrini_term" ~ "Gasparrini_et_al",
                                                                  Cohort == "Molly" ~ "Gibson_et_al",
                                                                  Cohort == "NEC_control" ~ "Thanert_et_al",
                                                                  Cohort == "No_abx_preterm" ~ "This_study",
                                                                  Cohort == "P4_cross_sectional" ~ "This_study"))

cat_var <- c("Patient",'Sex','Delivery_route', 'Necrotizing_enterocolitis', 'Site','Maternal_substance_abuse','Maternal_preeclampsia','Maternal_diabetes', 'Maternal_chorioamnionitis', 'Maternal_hypertension','mat_inpt_antimicrobe', 'mat_outpt_antimicrobial','Multiple_births','Cohort', 'Site', 'abx_group_updated',
             'abx_group_names', 'Study')

num_var <- c('Gestational_age_weeks', 'Birthweight','DOL','abx_exposure_count', 'antiviral_exposure', 'antiviral_exposure',	'aminoglycoside_exposure',	
'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure',	
'lincosamide_exposure',	'antifungal_exposure',	'carbapenem_exposure',	'other_antimicrobials_exposure',	'other_non_antimicrobials_exposure','glycopeptide_exposure', 'med_exposure_count',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure', 
  'cum_days_formula', 'cum_days_milk', 'frac_days_formula','frac_days_milk', "acute_fentanyl_exposure", "acute_cholecalciferol_vit_d__exposure", "acute_dopamine_exposure", "acute_levothyroxine_exposure", "acute_hydrocortisone_exposure", "acute_insulin_exposure", "acute_surfactant_exposure", "acute_iron_exposure", "acute_indomethacin_exposure", "acute_dexamethasone_exposure",
"acute_citric_acid_sodium_citrate_bicitra__exposure", "acute_lorazapam_ativan__exposure", )

pterm_infants_meta <- pterm_infants_meta %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(results_plot$Variable, names(.))), as.numeric))

pterm_infants_meta$abx_group_updated <- factor(pterm_infants_meta$abx_group_updated, levels = c("None", "Early only", "Early and after"))

maaslin_taxa_results <- Maaslin2(input_data = pterm_infants_otu,
                                input_metadata = pterm_infants_meta,
                                transform = "LOG",
                                normalization = "NONE",
                                min_abundance = 0.0001,
                                min_prevalence = 0.1,
                                output = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/MaAslin2/Taxa/all_sign_var",
                                fixed_effects = c("abx_group_updated",repeat_permanova_sign_var),
                                random_effects = c("Patient", "Site", "Study"))

############ --------------- Coefficient Plots ------------------------------#############################################
taxa_sign_out <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/MaAslin2/Taxa/all_sign_var/significant_results.xlsx")

num_samples <- 2406

taxa_sign_out_filt <- taxa_sign_out %>%
  filter(qval < 0.05) %>%
  mutate(
    # grab text AFTER "g__" up to the next dot
    genus  = str_extract(feature, "(?<=g__)[^.]+"),
    genus  = if_else(is.na(genus) | genus == "", "Unknown", genus),
    value  = as.factor(value),
    feature = gsub(".*s__", "", feature)
  )
write.csv(taxa_sign_out_filt, "260205_MaAslin2_all_sign_var_results.csv", row.names = F)

unique_taxa <- unique(taxa_sign_out_filt$feature)
unique_genera <- unique(taxa_sign_out_filt$genus)

maaslin_taxa_abxGroup <- taxa_sign_out %>% filter(metadata == "abx_group_updated") %>% 
  #filter(qval < 0.05) %>%
  mutate(
    # grab text AFTER "g__" up to the next dot
    genus  = str_extract(feature, "(?<=g__)[^.]+"),
    genus  = if_else(is.na(genus) | genus == "", "Unknown", genus),
    value  = as.factor(value),
    feature = gsub(".*s__", "", feature)
  ) %>%
  arrange(desc(coef))

#obtain min and max coefficient values (needed for plotting later)
min <- min(maaslin_taxa_abxGroup$coef)
max <- max(maaslin_taxa_abxGroup$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
maaslin_taxa_abxGroup <- maaslin_taxa_abxGroup %>% mutate(species = fct_reorder(feature, coef))

maaslin_taxa_abxGroup$coefmax <- maaslin_taxa_abxGroup$coef + maaslin_taxa_abxGroup$stderr
maaslin_taxa_abxGroup$coefmin <- maaslin_taxa_abxGroup$coef - maaslin_taxa_abxGroup$stderr

maaslin_taxa_abxGroup$CI95upper <- maaslin_taxa_abxGroup$coef + 1.96*maaslin_taxa_abxGroup$stderr
maaslin_taxa_abxGroup$CI95lower <- maaslin_taxa_abxGroup$coef - 1.96*maaslin_taxa_abxGroup$stderr

maaslin_taxa_abxGroup$value <- factor(maaslin_taxa_abxGroup$value, levels = c("Early only", "Early and after"))
write.csv(maaslin_taxa_abxGroup, "260205_MaAslin2_sign_taxa_abx_group.csv", row.names = F)

# Plot coefficients
x_min_lim <- (floor(min / 5) * 6) 
x_max_lim <- (ceiling(max / 5) * 6) 
colors_list <- c("#44637D","#AAC9ED")

abxgroup_coeff_p <- ggplot(maaslin_taxa_abxGroup, aes(x=coef, y=species,color=value))+
  geom_point(size = 3)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.5)+
  scale_color_manual(values=colors_list)+
  theme_linedraw()+
  theme(axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=17),
        axis.title.x = element_text(size=17),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.position = "bottom",
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=1))+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  xlab('Coefficient') + ggtitle("Antibiotic Exposure Group")
ggsave("260205_GM_dev_abx_group_MaAslin2_taxa_coeff.pdf", plot = abxgroup_coeff_p, bg = "transparent", width = 11, height = 12)

#### Generate prevalence plot ###
### Access species abundance data 
abund_gmdev <- psmelt(merged_ps)

abund_gmdev <- abund_gmdev %>%  mutate(
    genus = str_extract(OTU, "g__[^|]+"),  # extract 'g__Genus'
    genus = str_remove(genus, "g__"),          # remove 'g__' prefix
    genus = if_else(is.na(genus), "Unknown", genus),
    OTU=gsub(".*s__", "", OTU))

abund_abx_group <- subset(abund_gmdev, OTU %in% maaslin_taxa_abxGroup$feature)

# Re-order feature abs abundances according to coefficient value
# This order is preserved in filename, which was previously ordered: 
order_abx_group <- data.frame(OTU = unique(maaslin_taxa_abxGroup$feature))
# sort = F makes sure that the order is based on order_index df which is ordered by coefficient value
abund_abx_group.df <- merge(order_abx_group, abund_abx_group, by="OTU", sort = F)
abund_abx_group.df$OTU <- as.factor(abund_abx_group.df$OTU)
abund_abx_group.df <- abund_abx_group.df %>% mutate(OTU = fct_relevel(OTU, order_abx_group$OTU))

# For each significant feature in each sample, set AbunNotZero == 1 if its 
# abundance is greater than 0. 
abund_abx_group.df$AbunNotZero <- 1*(abund_abx_group.df$Abundance > 0)

# Summarize pathway prevalence by variable of interest
abund_abx_group.df.n0 <- abund_abx_group.df %>% 
  group_by(OTU, abx_group) %>%
  dplyr::summarise(N = n(), SumNotZero = sum(AbunNotZero)) %>%
  mutate(PercentNotZero = 100*(SumNotZero / N)) %>%
  ungroup()
write.csv(abund_abx_group.df.n0, "260205_GM_dev_abx_group_maaslin2_species_prevalence.csv", 
          row.names = F)

prev_p_abx_group <- ggplot(abund_abx_group.df.n0, aes(x = abx_group, y=OTU, fill=PercentNotZero))+
  geom_tile()+
  scale_fill_distiller(palette ='RdBu', type='seq', breaks = c(0,50,100), 'Prevalence (%)')+
  theme_classic()+
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(size=17, angle=45, hjust=1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=17),
        legend.text = element_text(size = 18),
        panel.border = element_rect(fill=NA, colour="black", size=1))+
  scale_y_discrete(limits=rev)
ggsave("260205_GM_dev_abx_group_MaAslin2_taxa_prev.pdf", plot = prev_p_abx_group, bg = "transparent", width = 11, height = 12)

combined_abx_group <- abxgroup_coeff_p | prev_p_abx_group 
ggsave("260205_GM_dev_MaAslin2_taxa_plot.pdf", plot = combined_abx_group, bg = "transparent", width = 12, height = 6)

############---------------- Other numerical variables ------------------------------------------#########
library(tidytext)
library(plyr)
library(stringr)

perm_cat_vector = c("DOL" = "#3A1F46", "Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

taxa_sign_out_num <- taxa_sign_out_filt %>% filter(metadata != "abx_group_updated") %>% 
  mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 
write.csv(taxa_sign_out_num, "260205_sign_MaAslin2_taxa.csv", row.names = F)

taxa_sign_out_num <- read_excel("260205_sign_MaAslin2_taxa.xlsx")
taxa_sign_out_num <- taxa_sign_out_num %>% mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 


taxa_sign_out_num$metadata_category <- factor(taxa_sign_out_num$metadata_category, levels = names(perm_cat_vector))

#obtain min and max coefficient values (needed for plotting later)
min <- min(taxa_sign_out_num$coef)
max <- max(taxa_sign_out_num$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
#maaslin_taxa <- taxa_sign_out_num %>% mutate(species = fct_reorder(feature, coef))

taxa_sign_out_num$coefmax <- taxa_sign_out_num$coef + taxa_sign_out_num$stderr
taxa_sign_out_num$coefmin <- taxa_sign_out_num$coef - taxa_sign_out_num$stderr

taxa_sign_out_num$CI95upper <- taxa_sign_out_num$coef + 1.96*taxa_sign_out_num$stderr
taxa_sign_out_num$CI95lower <- taxa_sign_out_num$coef - 1.96*taxa_sign_out_num$stderr

taxa_sign_out_num$direction <- factor(taxa_sign_out_num$direction, levels = c("Positive", "Negative"))

library(ggh4x)

genus_color_vector <- c(
  "Veillonella"      = "#A8A55C",  # blue
  "Streptococcus"      = "#68434E",  # green
  "Clostridium"    = "#B07159",  # red
  "Klebsiella"     = "#2F4B7A",  # orange
  "Enterobacter"      = "#8C2B0E",  
  "Staphylococcus"      = "#B47E83",  # brown
  "Enterococcus"  = "#F2A450",  # light blue
  "Escherichia"    = "#9F7E59",  # pink
  "Cutibacterium"          = "#6EA57B",  # lavender
  "Finegoldia"       = "#8DA0CB",  # light yellow
  "Clostridioides"          = "#088096"  
)

custom_genus_colors <- c(

  # --- Bacteroidota: blue-greens ---
  "g__Bacteroides"      = "#088096",
  "g__Parabacteroides"  = "#7AD4E4",

  # --- Actinobacteriota: warm teals ---
  "g__Bifidobacterium"  = "#9AC8A4",
  "g__Cutibacterium"    = "#6EA57B",

  # --- Firmicutes Clostridia: orange-browns ---
  "g__Clostridium"        = "#B07159",
  "g__Erysipelatoclostridium" = "#6C3322",

  # --- Firmicutes Bacilli: pink-reds ---
  "g__Enterococcus"     = "#F2A450",
  "g__Staphylococcus"   = "#B47E83",
  "g__Streptococcus"    = "#68434E",

  # --- Proteobacteria: blue-purple tones ---
  "g__Citrobacter"      = "#DFCB91",
  "g__Enterobacter"     = "#8C2B0E",
  "g__Escherichia"      = "#9F7E59",
  "g__Klebsiella"       = "#2F4B7A",
  "g__Serratia"         = "#7C97C2",

  # --- Negativicutes (Veillonellaceae): olive-greens ---
  "g__Veillonella"      = "#A8A55C",
    # --- Other: neutral gray ---
  "Other"               = "#A9A9A9"   # darkgray
)

custom_colors_top6 <- c("g__Escherichia" = "#9F7E59","g__Enterobacter" = "#8C2B0E", "g__Klebsiella" = "#2F4B7A", "g__Staphylococcus" = "#B47E83","g__Finegoldia" = "#8DA0CB","g__Enterococcus" = "#F2A450")

taxa_dol_coeff <- taxa_sign_out_num %>% filter(metadata_category == "DOL") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#3A1F46", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260205_GM_dev_MaAslin2_taxa_DOL_var_coeff_plot.pdf", plot = taxa_dol_coeff, bg = "transparent", width = 5, height = 4)

taxa_baseline_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Baseline") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#7F4B89", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260205_GM_dev_MaAslin2_taxa_baseline_var_coeff_plot.pdf", plot = taxa_baseline_coeff, bg = "transparent", width = 4, height = 3)

taxa_diet_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Diet") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#F3DAE4", color = "black"),
    strip.text = element_text(size = 11, face = "bold", color = "black"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.box = "horizontal")
ggsave("260205_GM_dev_MaAslin2_taxa_diet_var_coeff_plot.pdf", plot = taxa_diet_coeff, bg = "transparent", width = 9, height = 4)

taxa_abx_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Antibiotic") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#E3A5D6", color = "black"),
    strip.text = element_text(size = 12, face = "bold", color = "black"),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.box = "horizontal")
ggsave("260205_GM_dev_MaAslin2_taxa_abx_var_coeff_plot.pdf", plot = taxa_abx_coeff, bg = "transparent", width = 18, height = 6)

taxa_med_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Non-antibiotic medication") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#7C5467", color = "black"),
    strip.text = element_text(size = 9, face = "bold", color = "white"),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 12, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260205_GM_dev_MaAslin2_taxa_med_var_coeff_plot.pdf", plot = taxa_med_coeff, bg = "transparent", width = 22, height = 8)

###################### ---------------------------- ShortBRED-------------------###########################################
## Merge term infant ARGs with class
arg.map <- read.delim('~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/ShortBRED_mapping.txt',header = F)

## Merge preterm infant ARGs with class
arg.byclass <- merge(arg_df, arg.map, by.x = 'Family', by.y = 'V1') %>% 
  dplyr::rename(Class = V2)

arg.byclass_preterm <- arg.byclass %>% 
  dplyr::group_by(Sample_ID_R_final, Class) %>% 
  dplyr::summarise(Count = sum(Count))

arg.byclass_preterm.wide <- dcast(arg.byclass_preterm, Sample_ID_R_final ~ Class, value.var = 'Count')
arg.byclass_preterm.wide[is.na(arg.byclass_preterm.wide)] <- 0
row.names(arg.byclass_preterm.wide) <- arg.byclass_preterm.wide$Sample_ID_R_final
arg.byclass_preterm.wide$Sample_ID_R_final <- NULL


# Create metadata file
merged_ps_meta <- data.frame(sample_data(merged_ps))

merged_ps_meta_filt <- merged_ps_meta %>% select(Sample_ID_R_final, Patient, Cohort, Site, DOL, DOL_bin, Gestational_age_weeks, Birthweight, Sex, Delivery_route, abx_exposure_count, abx_group_updated, abx_group_names)

merged_ps_meta_filt$abx_group_updated <- factor(merged_ps_meta_filt$abx_group_updated, levels = c("None", "Early only", "Early and after"))

############--------- Run MaAslin2 -------------------------------############################

shortbred_preterm <- Maaslin2(input_data = arg.byclass_preterm.wide,
                                input_metadata = merged_ps_meta_filt,
                                transform = "LOG",
                                normalization = "NONE",
                                min_abundance = 0.0001,
                                min_prevalence = 0.1,
                                output = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/MaAslin2/ShortBRED/Preterm_site_cohort_random_effect_v2",
                                fixed_effects = c("Gestational_age_weeks","abx_group_updated, None"),
                                random_effects = c("Patient", "Site", "Cohort"))


```


Determine the hospitalized site proportion in the abx-exposed infants
```{r}
abx_infants <- merged_ps_meta %>% filter(abx_group_updated != "None")
table(abx_infants$Site)

## Get proportions per site
hosp_site_prop <- abx_infants %>% count(Site) %>% mutate(proportion = n / sum(n))
```

Random Forest Analysis to predict microbiota age
```{r}
library(randomForest)
library(dplyr)
library(stringr)
library(tidyr)
library(caret)

load("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/260109_Preterm_GA_matched_PS_object.RData")
load("my_environment.RData")
# 1. Apply random forests regression to regress relative abundances of all species predicted by MetaPhlAn3 in infant stool samples against their chronological age 
## Default parameters used: ntree = 10,000, importance = TRUE
# 2. Fivefold cross-validation was performed using 'rfcv' function over 100 iterations to estimate the min number of features (taxa) needed to accurately predict microbiota age. This was done using 100 iterations of the "importance" function 
### Importance determined based on the percentage increase in mean-squared error of microbiota age prediction when the relative abundance values of each taxon were randomly permuted 
# 3. Sparse model consisting of the 50 most important features constructed and trained on half of abx-naive near-term infants that were randomly selected 
# 4. Validate model in remaining half of abx-naive near-term infants 
# 5. Apply model to each preterm infant group 

# Renormalize ps object to ensure that relative abundance sums to 100
ps_filt_dol60_renorm  <- transform_sample_counts(merged_ps, function(x) (x / sum(x))*100 )

# # Access taxa relative abundances, and sample metadata
# mtg_relabund.df <- data.frame(otu_table(ps_m3_mtg_long_renorm), check.names = FALSE)

# Load your taxa relative abundance dataframe and chronological age vector
relabund_meta_full.df <- psmelt(ps_filt_dol60_renorm) 
# Rename each OTU so that the name only has species names not the full phylogenetic breakdown 
relabund_meta.df <- relabund_meta_full.df %>%
    mutate(OTU = sapply(str_split(OTU, "\\|"), function(x) if(length(x) >= 7) x[7] else NA))

# Subset to only rows from ABN samples  
relabund_meta_naive.df <- relabund_meta.df %>%
  filter(abx_group_updated == "None")

# randomForest requires that each row represent a sample and columns represent relative abundance of taxa
## we will also need a column for the response variable, age
# This means we will need to reformat the output of psmelt() which outputs a long-format dataframe where:
## each row is a specific taxon in a sample and columns include sample ID and metadata like age 

# First pivot dataframe to wide format and 
taxa_abundance <- relabund_meta_naive.df %>%
  select(Sample, OTU, Abundance) %>%
  pivot_wider(names_from = OTU, values_from = Abundance, values_fill = 0) # values_fill = 0 makes sure that missing values (e.g., taxa absent in a sample) are set to 0

# Add age as a separate column
taxa_abundance <- taxa_abundance %>%
  left_join(
    relabund_meta_naive.df %>% select(Sample, DOL, Site) %>% distinct(),
    by = "Sample"
  )

# 2. Randomly split the data upfront into training and validation sets
set.seed(142)  # For reproducibility
train_indices <- sample(1:nrow(taxa_abundance), size = 0.6 * nrow(taxa_abundance))
train_data <- taxa_abundance[train_indices, ]
validation_data <- taxa_abundance[-train_indices, ]

# Separate predictors and response for training set
x_train <- train_data %>% select(-Sample, -DOL)
y_train <- train_data$DOL

# Step 2: Perform 100 iterations for feature importance and RFCV
iterations <- 100
importance_list <- vector("list", iterations)
rfcv_results_list <- vector("list", iterations)

for (i in 1:iterations) {
  cat("Iteration:", i, "\n")
  
  # Fit Random Forest model on the training set
  rf_model <- randomForest(
    x = x_train,
    y = y_train,
    ntree = 10000,
    importance = TRUE
  )
  
  # Extract feature importance
  importance_df <- as.data.frame(importance(rf_model))
  importance_df <- importance_df %>%
    mutate(Taxon = rownames(importance_df),
           Iteration = i) %>%
    arrange(desc(IncNodePurity))
  
  # Append this iteration's importance data to the list
  importance_list[[i]] <- importance_df
  
  # Perform RFCV for this iteration
  rfcv_results <- rfcv(
    trainx = x_train,
    trainy = y_train,
    cv.fold = 10,
    step = 0.5
  )
  
  # Store the rfcv results with the iteration number
  rfcv_results_list[[i]] <- data.frame(
    Iteration = i,
    n_features = rfcv_results$n.var,
    error = rfcv_results$error.cv
  )
}

# Combine all feature importance results
feature_importance_df <- bind_rows(importance_list)

# Aggregate feature importance (Mean and SD for both %IncMSE and IncNodePurity)
aggregated_importance <- feature_importance_df %>%
  group_by(Taxon) %>%
  summarise(
    MeanIncNodePurity = mean(IncNodePurity, na.rm = TRUE),
    SDIncNodePurity = sd(IncNodePurity, na.rm = TRUE),
    MeanPercIncMSE = mean(`%IncMSE`, na.rm = TRUE),
    SDPercIncMSE = sd(`%IncMSE`, na.rm = TRUE)
  ) %>%
  arrange(desc(MeanPercIncMSE))


# Combine all rfcv results
rfcv_results_df <- bind_rows(rfcv_results_list)

# Summarize rfcv results across iterations
aggregated_rfcv <- rfcv_results_df %>%
  group_by(n_features) %>%
  summarise(
    MeanError = mean(error, na.rm = TRUE),
    SDError = sd(error, na.rm = TRUE)
  )

# Plot aggregated RFCV results
rfcv_error_variables_plot <- ggplot(aggregated_rfcv, aes(x = n_features, y = MeanError)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = MeanError - SDError, ymax = MeanError + SDError), width = 5) +
  theme_minimal() +
  labs(
    x = "Number of Variables",
    y = "Cross-validation error",
    title = "RFCV Error Across Variable Subsets"
  )
ggsave("260202_GMdev_RF_RFCV_aggregated_results_plot.pdf", plot = rfcv_error_variables_plot, bg = "transparent", width = 6, height = 6)

# Based on rfcv results, minimal improvement in predictive performance was observed when including taxa beyond the top N
top_taxa_n <- 50
top_taxa_rf <- aggregated_importance %>%
  arrange(desc(MeanPercIncMSE)) %>%
  dplyr::slice(1:top_taxa_n) %>%
  pull(Taxon)  # Select top N taxa based on mean importance

top50 <- aggregated_importance %>% arrange(desc(MeanPercIncMSE)) %>% slice_head(n=50)

## Plot the top 50 taxa by %IncMSE
top50_taxa_plot <- ggplot(top50, aes(x = reorder(Taxon, MeanPercIncMSE), y = MeanPercIncMSE)) +
  geom_point() +
  geom_errorbar(aes(ymin = MeanPercIncMSE - SDPercIncMSE, ymax = MeanPercIncMSE + SDPercIncMSE), width = 0.3) +
  coord_flip() +
  labs(
    x = "",
    y = "Increase in Mean Squared Error (%)",
    title = "Top 50 Age-Discriminatory Taxa"
  ) +
  theme_minimal()
ggsave("260202_GMdev_RF_top50_taxa_by_MSE_plot.pdf", plot = top50_taxa_plot, bg = "transparent", width = 6, height = 8)

# Step 3: Train sparse model on training set and validate on validation set
x_train_sparse <- x_train[, top_taxa_rf, drop = FALSE]

sparse_rf_model <- randomForest(
  x = x_train_sparse,
  y = y_train,
  ntree = 10000,
  importance = TRUE
)


#############-------------Apply sparse model to Early only and Early and After infants--------------#################
ps_early_only <- ps_filt_dol60_renorm %>% subset_samples(abx_group_updated == "Early only") # 801 samples; 171 infants
ps_early_and_after <- ps_filt_dol60_renorm %>% subset_samples(abx_group_updated == "Early and after") # 1312 samples; 183 infants
ps_abxN <- ps_filt_dol60_renorm %>% subset_samples(abx_group_updated == "None") # 282 samples; 38 infants
ps_abxN_val <- ps_abxN %>% subset_samples(Sample_ID_R_final %in% validation_data$Sample) # 113 samples; 35 infants 
# ps_abxN_df <- data.frame(sample_data(ps_abxN))
# ps_abxN_val_df <- ps_abxN_df %>% subset(Sample_ID_R_final %in% validation_data$Sample)

predict_early_only <- predict_microbiota_age(ps_early_only, sparse_rf_model, top_taxa_rf)
predict_early_and_after <- predict_microbiota_age(ps_early_and_after, sparse_rf_model, top_taxa_rf)
predict_abxN_val <- predict_microbiota_age(ps_abxN_val, sparse_rf_model, top_taxa_rf)

predict_early_only <- data.frame(predict_early_only)
predict_early_and_after <- data.frame(predict_early_and_after)
predict_abxN_val <- data.frame(predict_abxN_val)

# Combine all datasets into one
combined_df <- bind_rows(
  predict_early_only %>% mutate(Group = "Early only"),
  predict_early_and_after %>% mutate(Group = "Early and after"),
  predict_abxN_val %>% mutate(Group = "None"))

rf_plot <- ggplot(combined_df, aes(x = DOL, y = Predicted_Microbiota_Age, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  labs(
    x = "Chronological age (days)",
    y = "Microbiota age (days)",
    title = "Microbiota Maturation Across Groups"
  ) +
  theme_minimal() +
  scale_color_manual(values = custom_colors_abx) + 
  theme(
    axis.title = element_text(size = 16, face = "bold"),      
    axis.text = element_text(size = 14),                      
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )
ggsave("260202_GMdev_RF_abxN_model_validation.pdf", plot = rf_plot, bg = "transparent", width = 10, height = 10)

########## Calculating and comparing MAZ values ##########
# Bin DOL into age bins (month-based)
combined_df2 <- combined_df %>%
  mutate(
    Month = case_when(
      DOL <= 30 ~ "Month 1",
      DOL <= 60 ~ "Month 2",
      TRUE ~ NA_character_
    )
  )

# Calculate mean and sd of predicted microbiota age for each bin in reference group
reference_group <- "None"
reference_stats <- combined_df2 %>%
  filter(abx_group == reference_group) %>%
  group_by(DOL_bin) %>%
  summarise(
    mu_ref = mean(Predicted_Microbiota_Age, na.rm = TRUE),
    sigma_ref = sd(Predicted_Microbiota_Age, na.rm = TRUE),
    .groups = "drop"
  )

# Merge stats back and calculate MAZ
combined_df2 <- combined_df2 %>%
  left_join(reference_stats, by = "DOL_bin") %>%
  mutate(MAZ = (Predicted_Microbiota_Age - mu_ref) / sigma_ref)

combined_df2$abx_group <- factor(combined_df2$abx_group, levels = c("Early and after", "Early only", "None"))

### Plot MAZ per group
MAZ_p <- ggplot(combined_df2, aes(x = abx_group, y = MAZ, fill = abx_group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  scale_color_manual(values = custom_colors_abx) + 
  scale_fill_manual(values = custom_colors_abx) + 
  stat_compare_means(method = "wilcox.test", label = "p.signif", ref.group = reference_group) +
  facet_wrap(~ Month) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 16, face = "bold"),
    legend.position = "none"
  ) +
  ylab("MAZ (Microbiota-for-Age Z-score)") +
  xlab(NULL)
ggsave("251211_GMdev_predicted_MAZ_Comp.pdf", plot = MAZ_p, bg = "transparent", width = 10, height = 8)


#save.image("251020_GM_dev_wP4_random_forests.RData")
```

Prevalence of top taxa over time 
```{r}
otu_df <- as.data.frame(otu_table(merged_ps))
## Extract only species names from the taxa names
rownames(otu_df) <- sapply(str_split(rownames(otu_df), "\\|"), function(x) if(length(x) >= 7) x[7] else NA)
otu_df <- t(otu_df)
otu_df <- data.frame(otu_df)
otu_df <- otu_df %>% rownames_to_column("Sample_ID_R_final")

## Get taxonomy at desired level (genus, species)
tax_df <- as.data.frame(tax_table(ps_filt_dol60)) 
rownames(tax_df) <- sapply(str_split(rownames(tax_df), "\\|"), function(x) if(length(x) >= 7) x[7] else NA)
tax_df <- tax_df %>% rownames_to_column("TaxaID")

## Choose desired taxonomic rank
tax_level <- "Genus"

## Map TaxaID to taxonomic name (e.g., genus)
tax_map <- tax_df %>% select(TaxaID, !!sym(tax_level)) %>%
  mutate(Taxon = if_else(is.na(!!sym(tax_level)), TaxaID,!!sym(tax_level)))

## Reshape to long format
otu_long <- otu_df %>% pivot_longer(-Sample_ID_R_final, names_to = "TaxaID", values_to = "Abundance") %>%
  left_join(tax_map, by = "TaxaID")

## Add prevalence column 
otu_long <- otu_long %>% mutate(Prevalence = as.integer(Abundance > 0))

## Merge with sample metadata
meta_df <- as.data.frame(sample_data(merged_ps))

otu_long <- otu_long %>% left_join(meta_df, by = "Sample_ID_R_final")
otu_long_earlyOnly <- otu_long %>% filter(abx_group_updated == "Early only")
otu_long_earlyAfter <- otu_long %>% filter(abx_group_updated == "Early and after")
otu_long_naive <- otu_long %>% filter(abx_group_updated == "None")

custom_colors_prev <- c("g__Clostridioides" = "#E69F00", "g__Clostridium" = "#F0E442", "g__Escherichia" = "#56B4E9",
                                          "g__Enterobacter" = "#0072B2", "g__Klebsiella" = "#009E73", "g__Cutibacterium" = "#D55E00",
                                          "g__Veillonella" = "#CC79A7", "g__Staphylococcus" = "#2B2D42", "g__Enterococcus" = "#FC8D62",
                                          "g__Finegoldia" = "#8DA0CB")
custom_colors_top6 <- c("g__Escherichia" = "#9F7E59","g__Enterobacter" = "#8C2B0E", "g__Klebsiella" = "#2F4B7A", "g__Staphylococcus" = "#B47E83","g__Finegoldia" = "#8DA0CB","g__Enterococcus" = "#F2A450")

## Select top taxa based on average prevalence across all timepoints
top_taxa <- otu_long %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa, "260203_GM_Dev_top10_prevalent_taxa.csv", row.names = F)

top_taxa_eeo <- otu_long_earlyOnly %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa_eeo, "260203_GM_Dev_top10_prevalent_taxa_EEO.csv", row.names = F)

top_taxa_eas <- otu_long_earlyAfter %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa_eas, "260203_GM_Dev_top10_prevalent_taxa_EAS.csv", row.names = F)

top_taxa_naive <- otu_long_naive %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa_naive, "260203_GM_Dev_top10_prevalent_taxa_ABN.csv", row.names = F)

## Select top 6 for plotting
top_taxa <- top_taxa %>% slice_head(n = 6) 

## prevalence summary over time 
prevalence_summary <- otu_long %>% filter(Taxon %in% top_taxa$Taxon) %>% group_by(Taxon, DOL, abx_group_names) %>% summarise(Prevalence = mean(Prevalence) * 100, mean(Prevalence), .groups = "drop") %>% 
  mutate(Taxon = factor(Taxon, levels = c("g__Escherichia", "g__Klebsiella", "g__Enterobacter", "g__Staphylococcus","g__Finegoldia","g__Enterococcus")))

prevalence_DOL_plot <- ggplot(prevalence_summary, aes(x = DOL, y = Prevalence, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  scale_color_manual(values = custom_colors_top6) + 
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Postnatal age [days]",
    y = "Prevalence [% of cohort]",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )
ggsave("260203_GM_Dev_NoAfterOnly_top6_prevalent_taxa_DOL.pdf", plot = prevalence_DOL_plot, bg = "transparent", width = 10, height = 8)


prevalence_DOL_plot_facet <- ggplot(prevalence_summary, aes(x = DOL, y = Prevalence, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, size = 1.2) +
  facet_wrap(~ abx_group_names, nrow = 1) +
  scale_color_manual(values = custom_colors_top6) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Postnatal age [days]",
    y = "Prevalence [% of cohort]",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )
ggsave("260203_GM_Dev_NoAfterOnly_top6_prevalent_taxa_DOL_faceted.pdf", plot = prevalence_DOL_plot_facet, bg = "transparent", width = 10, height = 6)
```

Stacked barplots at the genus level for preterm and near-term infants
```{r}
## Barplot of taxonomic composition (Genus level) by Enterotype #######
# Step 1: Collapse to Genus level and transform to relative abundance
ps_genus <- tax_glom(ps_infants, taxrank = "Genus")
ps_relabun <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Step 2: Extract OTU table
otu_df <- as.data.frame(otu_table(ps_relabun))
if (taxa_are_rows(ps_relabun)) {
  otu_df <- t(otu_df)
}
otu_df <- as.data.frame(otu_df) %>%
  tibble::rownames_to_column("Sample_ID_R_final")

# Step 3: Pivot to long format
otu_long <- otu_df %>%
  pivot_longer(-Sample_ID_R_final, names_to = "OTU", values_to = "Abundance")

# Step 4: Merge sample metadata
sample_df <- data.frame(sample_data(ps_relabun)) 
otu_long <- left_join(otu_long, sample_df, by = "Sample_ID_R_final")

# Step 5: Add Genus information from tax_table
tax_df <- as.data.frame(tax_table(ps_relabun)) %>%
  tibble::rownames_to_column("OTU")

otu_long <- left_join(otu_long, tax_df[, c("OTU", "Genus")], by = "OTU")

# Step 6: Summarize mean relative abundance
otu_summary <- otu_long %>%
  filter(!is.na(Genus), !is.na(abx_group_names)) %>%
  group_by(abx_group_names, Genus, DOL_bin) %>%
  summarise(Mean_RelativeAbund = mean(Abundance), .groups = "drop")

# Step 7: Identify top 15 genera across all enterotypes
top_genera <- otu_summary %>%
  group_by(Genus) %>%
  summarise(Total = sum(Mean_RelativeAbund)) %>%
  arrange(desc(Total)) %>%
  head(15) %>%
  pull(Genus)

# Step 8: Group others into "Other"
otu_summary <- otu_summary %>%
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other")) %>%
  group_by(abx_group_names, Genus, DOL_bin) %>%
  summarise(Mean_RelativeAbund = sum(Mean_RelativeAbund), .groups = "drop")


### Create a color-palette with genera belonging to the same families/order with similar hues
custom_genus_colors <- c(

  # --- Bacteroidota: blue-greens ---
  "g__Bacteroides"      = "#088096",
  "g__Parabacteroides"  = "#7AD4E4",

  # --- Actinobacteriota: warm teals ---
  "g__Bifidobacterium"  = "#9AC8A4",
  "g__Cutibacterium"    = "#6EA57B",

  # --- Firmicutes Clostridia: orange-browns ---
  "g__Clostridium"        = "#B07159",
  "g__Erysipelatoclostridium" = "#6C3322",

  # --- Firmicutes Bacilli: pink-reds ---
  "g__Enterococcus"     = "#F2A450",
  "g__Staphylococcus"   = "#B47E83",
  "g__Streptococcus"    = "#68434E",

  # --- Proteobacteria: blue-purple tones ---
  "g__Citrobacter"      = "#DFCB91",
  "g__Enterobacter"     = "#8C2B0E",
  "g__Escherichia"      = "#9F7E59",
  "g__Klebsiella"       = "#2F4B7A",
  "g__Serratia"         = "#7C97C2",

  # --- Negativicutes (Veillonellaceae): olive-greens ---
  "g__Veillonella"      = "#A8A55C",
    # --- Other: neutral gray ---
  "Other"               = "#A9A9A9"   # darkgray
)

genus_order <- c(
  # Bacteroidota
  "g__Bacteroides",
  "g__Parabacteroides",

  # Actinobacteriota
  "g__Bifidobacterium",
  "g__Cutibacterium",

  # Firmicutes: Clostridia
  "g__Clostridium",
  "g__Erysipelatoclostridium",

  # Firmicutes: Bacilli
  "g__Enterococcus",
  "g__Staphylococcus",
  "g__Streptococcus",

  # Proteobacteria (Enterobacterales)
  "g__Citrobacter",
  "g__Enterobacter",
  "g__Escherichia",
  "g__Klebsiella",
  "g__Serratia",

  # Negativicutes
  "g__Veillonella",

  # Other
  "Other"
)
otu_summary$Genus <- factor(otu_summary$Genus, levels = genus_order)

# Step 9: Plot
abx_group_DOL_bin_genus_bp <- ggplot(otu_summary, aes(x = abx_group_names, y = Mean_RelativeAbund, fill = Genus)) +
  geom_bar(stat = "identity", width = 0.85, color = "black") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = custom_genus_colors) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 16, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  labs(
    title = "Average Genus-Level Composition by Enterotype",
    x = NULL, y = "Relative Abundance"
  )
ggsave("260204_P4_GM_Dev_near-term_barplot_genus.pdf", plot = abx_group_DOL_bin_genus_bp, bg = "transparent", width = 20, height = 8)

```



Random Forest analysis to predict last time point preterm samples
```{r}
library(dplyr)
library(tidyr)
library(caret)
library(ranger)
library(Boruta)
library(doParallel)
library(ggplot2)
library(patchwork)

load("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/251020_GM_dev_wP4_random_forests.RData")
load("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260114_P4_GM_Dev_RF_last_tp.RData")

# Metadata
merged_ps_meta

# Get species abundance, richness and shannon values
metaphlan_mat <- data.frame(otu_table(merged_ps, taxa_are_rows = TRUE))
metaphlan_mat_t <- t(metaphlan_mat)

### Get species richness and shannon diversity values
ps.filt.trans <- transform_sample_counts(merged_ps, function(x) trunc(x*100000))
adiv <- estimate_richness(ps.filt.trans, measures=c('Observed', 'Shannon'))
adiv <- adiv %>% dplyr::rename(Richness = Observed)

metaphlan_mat_t2 <- merge(metaphlan_mat_t, adiv, by = "row.names")
rownames(metaphlan_mat_t2) <- metaphlan_mat_t2$Row.names
metaphlan_mat_t2$Row.names <- NULL

metaphlan_mat <- data.frame(t(metaphlan_mat_t2))
metaphlan_mat$temp <- rownames(metaphlan_mat)

metaphlan_mat2 <- metaphlan_mat %>%
  mutate(temp = ifelse(grepl("\\|", temp), sapply(str_split(temp, "\\|"), function(x) if(length(x) >= 7) x[7] else NA),
                       temp),
         temp = gsub("s__","", temp))
rownames(metaphlan_mat2) <- metaphlan_mat2$temp
metaphlan_mat2$temp <- NULL

# Transpose taxonomic data so rows = samples, cols = taxa features
metaphlan_samp <- t(metaphlan_mat2)

##load preterm shortbred output
arg.full <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/250905_P4_GM_Dev_ShortBRED_merged_output_n2585.txt"))
arg.full <- arg.full %>% mutate(Sample_ID_R_final = gsub("-", ".", paste0("p4_", Sample)))
arg.p4 <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251022_GM_Dev_P4_shortBRED_joint_output.txt"))
arg.p4 <- arg.p4 %>% mutate(Sample_ID_R_final = paste0("p4_", gsub("_", ".", Sample)))
arg_df <- rbind(arg.full, arg.p4)

arg.wide <- as.data.frame(dcast(arg_df, Sample_ID_R_final ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg.wide) <- arg.wide$Sample_ID_R_final
arg.wide <- arg.wide[, c(-1)]

##sum counts to get RPKM per sample
arg.wide$RPKM <- rowSums(arg.wide)

##get ARG richness
arg.wide$ARG_Richness <- specnumber(arg.wide[, 1:ncol(arg.wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg.wide$Sample_ID_R_final <- row.names(arg.wide)

#####--Align samples ----------------#################
common_samples <- Reduce(intersect, list(rownames(metaphlan_samp), rownames(arg.wide), merged_ps_meta$Sample_ID_R_final))

metaphlan_mat_sub <- metaphlan_samp[common_samples, , drop = FALSE]
shortbred_mat_sub <- arg.wide[common_samples, , drop = FALSE]

meta_sub <- merged_ps_meta %>% filter(Sample_ID_R_final %in% common_samples) %>% arrange(match(Sample_ID_R_final, common_samples))

## ---- Build feature matrix ---- ############
X <- cbind(metaphlan_mat_sub, shortbred_mat_sub)
X$Sample_ID_R_final <- NULL

## ---- Log-transform and filter constant features ----
pseudocount <- 1e-6
X_log <- log10(X + pseudocount)

global_sd <- apply(X_log, 2, sd, na.rm = TRUE)
X_log <- X_log[, global_sd > 0, drop = FALSE]

## ---- Outcome ----
meta_sub$Group <- factor(make.names(meta_sub$abx_group))   # avoid caret probability naming issues

rf_df <- data.frame(Group = meta_sub$Group, DOL = meta_sub$DOL, Patient = meta_sub$Patient, X_log, check.names = FALSE)


set.seed(123)

rf_res <- run_boruta_rf_parallel(
  data = rf_df,          # Group + predictors
  patient_col = "Patient",
  time_col = "DOL",
  train_p = 0.6,         # define training cohort once
  n_iter = 100,
  inner_p = 0.8,
  n_cores = 6,
  use_tentative = FALSE
)

perf_long <- rf_res$inner_perf_long
imp_avg   <- rf_res$inner_importance_summary
imp_avg$Feature <- gsub("^`|`$", "", imp_avg$Feature)

######----- Plot performance metrics ----------######
gg_box <-ggplot(perf_long, aes(x = Class, y = Value)) +
  geom_boxplot(outlier.size = 0.8) +
  facet_wrap(~ Metric, nrow = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
gg_box
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260119_P4_GM_dev_RF_train_data_performance_metrics.pdf", plot = gg_box, bg = "transparent", width = 10, height = 6)


cm <- rf_res$final_confusion
tab <- as.matrix(cm$table)
classes <- colnames(tab)
N <- sum(tab)

cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

gg_cm <- ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "black", linewidth = 0.6) +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradientn(
    colours = c("white", "#9ecae1", "#08306b"),
    values  = scales::rescale(c(0, max(cm_df$Freq), max(cm_df$Freq))),
    name = "Count"
  ) +
  scale_x_discrete(position = "top") +
  labs(x = "Actual", y = "Predicted") +
  coord_equal() +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold")
  )
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260114_P4_GM_dev_RF_last_tp_samples.pdf", plot = gg_cm, bg = "transparent", width = 10, height = 8)

outer_perf <- dplyr::bind_rows(lapply(classes, function(cls) {
  TP <- tab[cls, cls]
  FN <- sum(tab[, cls]) - TP
  FP <- sum(tab[cls, ]) - TP
  TN <- N - TP - FN - FP

  data.frame(
    Class = cls, TP = TP, TN = TN, FP = FP, FN = FN,
    Accuracy = (TP + TN) / N,
    Sensitivity = TP / (TP + FN),
    Specificity = TN / (TN + FP)
  )
}))
outer_perf

top20 <- imp_avg %>%
  arrange(desc(Mean_Importance)) %>%
  dplyr::slice(1:21)
write.csv(top20, "260119_top20_features_boruta.csv", row.names = F)

top20_mod <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260119_top20_features_boruta_labeled.xlsx") %>% arrange(desc(Mean_Importance))

p_top20_features <- ggplot(top20_mod, aes(x = Mean_Importance, y = reorder(Label, Mean_Importance))) +
  geom_col() +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank()) +
  labs(x = "Mean permutation importance (inner resamples)",
       y = "Feature",
       title = "Top 20 features (averaged over 100 inner splits)")
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260114_P4_GM_dev_RF_top20_boruta_features.pdf", plot = p_top20_features, bg = "transparent", width = 10, height = 8)
#save.image("260114_P4_GM_Dev_RF_last_tp.RData")

outer_perf_long <- outer_perf %>%
  pivot_longer(
    cols = c(Accuracy, Sensitivity, Specificity),
    names_to = "Metric",
    values_to = "Value"
  )

gg_perf_last_tp <- ggplot(outer_perf_long, aes(x = Class, y = Value)) +
  geom_col(fill = "#3182bd", width = 0.7) +
  facet_wrap(~ Metric, nrow = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Performance",
    title = "Random forest performance on held-out test patients"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260114_P4_GM_dev_RF_perf_metrics_last_tp.pdf", plot = gg_perf_last_tp, bg = "transparent", width = 10, height = 8)


train_idx <- createDataPartition(rf_df$Group, p = 0.7, list = FALSE)
train_df <- rf_df[train_idx, ]
test_df  <- rf_df[-train_idx, ]

## ---- Remove NZV (near-zero variance features) based on training only ----
train_x <- train_df[, -1, drop = FALSE]
test_x  <- test_df[,  -1, drop = FALSE]

## Need to remove NZV features from the test dataset as well since they must have the exact same features space as the training set
nzv_idx <- nearZeroVar(train_x)
if (length(nzv_idx) > 0) {
  train_x <- train_x[, -nzv_idx, drop = FALSE]
  test_x  <- test_x[,  -nzv_idx, drop = FALSE]
}

train_y <- train_df$Group
test_y  <- test_df$Group

########################################-------- Boruta Feature Selection --------------------###################################
## ---- Parallel backend (Boruta can benefit from faster RF backend; caret CV will also use it later) ----
n_cores <- max(1, parallel::detectCores() - 1)
cl <- makeCluster(n_cores)
registerDoParallel(cl)

set.seed(123)

## ---- Boruta ----
# doTrace = 1 prints progress; maxRuns controls iterations (increase for stability)
bor <- Boruta(x = train_x, y = train_y, doTrace = 1, maxRuns = 200, getImp = function(x, y, ...) {
    # Custom importance using ranger permutation importance
    # Boruta expects a numeric importance vector per feature
    rf <- ranger(
      dependent.variable.name = "y",
      data = data.frame(y = y, x),
      classification = TRUE,
      importance = "permutation",
      num.trees = 800,
      ...
    )
    rf$variable.importance
  }
)

## ---- Resolve tentative features ----
bor_fixed <- TentativeRoughFix(bor)

## Confirmed important features
confirmed_feats <- getSelectedAttributes(bor_fixed, withTentative = FALSE)

length(confirmed_feats)
head(confirmed_feats)

############################----Train sparse Random Forest using Boruta-selected features---------------###################################
# Build sparse training/testing datasets
train_sparse <- data.frame(Group = train_y, train_x[, confirmed_feats, drop = FALSE], check.names = FALSE)
test_sparse  <- data.frame(Group = test_y,  test_x[,  confirmed_feats, drop = FALSE], check.names = FALSE)

## ---- caret CV control ----
ctrl_all <- trainControl(
  method = "cv",
  number = 10,              # change folds here
  classProbs = TRUE,
  summaryFunction = defaultSummary,
  allowParallel = TRUE
)

## ---- Train sparse RF (ranger) ----
set.seed(23)
rf_sparse <- train(
  Group ~ .,
  data = train_sparse,
  method = "ranger",
  trControl = ctrl_all,
  tuneLength = 7,
  importance = "permutation",
  num.trees = 1500,
  metric = "Accuracy"
)

stopCluster(cl)
registerDoSEQ()

rf_sparse
rf_sparse$bestTune

############################----Predict test set + confusion matrix---------------###################################

pred_test <- predict(rf_sparse, newdata = test_sparse)
cm <- confusionMatrix(pred_test, test_sparse$Group)
cm

cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

gg_cm <- ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 4) +
  scale_x_discrete(position = "top") +
  labs(x = "Actual", y = "Predicted", fill = "Count") +
  coord_equal() +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260113_P4_GM_dev_RF_all_samples.pdf", plot = gg_cm, bg = "transparent", width = 10, height = 8)

imp_sparse <- varImp(rf_sparse, scale = TRUE)$importance
imp_sparse$Feature <- rownames(imp_sparse)
imp_sparse$Feature <- gsub("^`|`$", "", imp_sparse$Feature)

if ("Overall" %in% colnames(imp_sparse)) {
  imp_sparse$Weight <- imp_sparse$Overall
} else {
  imp_sparse$Weight <- rowMeans(imp_sparse[, setdiff(colnames(imp_sparse), "Feature"), drop = FALSE])
}
############################----top feature importances (15 strongest)---------------###################################

imp_sparse <- imp_sparse %>% arrange(desc(Weight))
write.csv(imp_sparse, "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260113_P4_GM_dev_RF_important_features_all_samples.csv", row.names = F)
n_top_plot <- min(20, nrow(imp_sparse))
imp_top <- imp_sparse[1:n_top_plot, ]

gg_weights <- ggplot(imp_top, aes(x = Weight, y = reorder(Feature, Weight))) +
  geom_col() +
  labs(x = "Importance", y = "Feature") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/RandomForest/Boruta/260113_P4_GM_dev_RF_important_features_all_samples.pdf", plot = gg_weights, bg = "transparent", width = 20, height = 8)

#################----Per-class TP/TN/FP/FN + Accuracy/Sensitivity/Specificity on test set---------------#########################
tab <- as.matrix(cm$table)  # rows=Pred, cols=Actual
classes <- colnames(tab)
N <- sum(tab)

perf_per_class <- lapply(classes, function(cls) {
  TP <- tab[cls, cls]
  FN <- sum(tab[, cls]) - TP
  FP <- sum(tab[cls, ]) - TP
  TN <- N - TP - FN - FP

  data.frame(
    Group = cls,
    TP = TP, TN = TN, FP = FP, FN = FN,
    Accuracy = (TP + TN) / N,
    Sensitivity = ifelse((TP + FN) == 0, NA_real_, TP / (TP + FN)),
    Specificity = ifelse((TN + FP) == 0, NA_real_, TN / (TN + FP))
  )
}) %>% bind_rows()

perf_long <- perf_per_class %>%
  select(Group, Accuracy, Sensitivity, Specificity) %>%
  pivot_longer(cols = c(Accuracy, Sensitivity, Specificity),
               names_to = "Metric", values_to = "Value")

gg_perf <- ggplot(perf_long, aes(x = Group, y = Value, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", Value)),
            position = position_dodge(width = 0.8),
            vjust = -0.25, size = 3) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Group", y = "Performance (test set)",
       title = "Sparse RF (Boruta-selected) performance by group") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank())

perf_per_class
gg_perf

#save.image("260112_RandomForest_P4.RData")

```


GA-matched (by MatchIt) SVM
```{r}
library(kernlab)
load("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/260109_Preterm_GA_matched_PS_object.RData")

ps_infants_ga_meta <- data.frame(sample_data(ps_infants_ga))

#### Summarize Metadata #########
meta_pt <- ps_infants_ga_meta
# Set categorical variables as factors !!!!!!!!!!! Modify these as needed later !!!!!!!!!!!!!!!!!!!!
categorical_vars <- c('Patient','abx_group', 'Sex', 'Delivery_route', 'Cohort')

meta_pt[,categorical_vars] <- lapply(meta_pt[,categorical_vars], factor)

numerical_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks", "PostmenstrualAgeWeeks")
meta_pt[,numerical_vars] <- lapply(meta_pt[,numerical_vars], as.numeric)

## All numerical variables are non-normal so we'll use median (IQR) instead of mean (SD)
normality_results <- lapply(numerical_vars, function(v) {
  p <- shapiro.test(meta_pt[[v]])$p.value
  data.frame(variable = v, p_value = p)
}) %>% bind_rows()


# Generate metadata comparison table 
comp_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks", "PostmenstrualAgeWeeks",'Cohort', 'Sex', 'Delivery_route')


tab_pt <- CreateTableOne(vars=comp_vars,
                         strata='abx_group', 
                         data=meta_pt)
tab_plot_pt <- print(tab_pt, 
                     showAllLevels = TRUE,
                     quote        = FALSE,
                     noSpaces     = TRUE,
                     test         = TRUE,
                     contDigits   = 1,
                     nonnormal    = numerical_vars,  
                     printToggle  = FALSE,
                     dropEqual    = TRUE,
                     explain      = FALSE)

# Convert to dataframe
tab_pt_df <- as.data.frame(tab_plot_pt) %>% rownames_to_column(var = "Variable")

# Set Table header
header <- str_squish(str_remove("Table 1. Infant demographics", "\n"))
# Set Table footer
footer <- str_squish(str_remove("Numbers are either No. or mean values", "\n"))
# Set custom_tab() defaults
customtab_defaults()
# Create the flextable object
flextable_processed_samples <- custom_tab(tab_pt_df, header, footer)

# Save as word .docx
save_as_docx(flextable_processed_samples, path = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/260110_metadata.docx",
             pr_section = prop_section(page_size = page_size(orient = "landscape"), type = "continuous"))

#### Pairwise wilcoxon tests to confirm matching PMA and GA ####
##### There are still significant differences between the None and abx-exposed groups but the differences are less than before
PMA_stat <- ps_infants_ga_meta %>% wilcox_test(PostmenstrualAgeWeeks ~ abx_group) %>% adjust_pvalue(method = "BH") %>% add_significance()
GA_stat <-  ps_infants_ga_meta %>% wilcox_test(Gestational_age_weeks ~ abx_group) %>% adjust_pvalue(method = "BH") %>% add_significance()

####################################

metaphlan_mat <- data.frame(otu_table(ps_infants_ga, taxa_are_rows = TRUE))
metaphlan_mat_t <- t(metaphlan_mat)

### Get species richness and shannon diversity values
ps.filt.trans <- transform_sample_counts(ps_infants_ga, function(x) trunc(x*100000))
adiv <- estimate_richness(ps.filt.trans, measures=c('Observed', 'Shannon'))
adiv <- adiv %>% dplyr::rename(Richness = Observed)

metaphlan_mat_t2 <- merge(metaphlan_mat_t, adiv, by = "row.names")
rownames(metaphlan_mat_t2) <- metaphlan_mat_t2$Row.names
metaphlan_mat_t2$Row.names <- NULL

metaphlan_mat <- data.frame(t(metaphlan_mat_t2))
metaphlan_mat$temp <- rownames(metaphlan_mat)

metaphlan_mat2 <- metaphlan_mat %>%
  mutate(temp = ifelse(grepl("\\|", temp), sapply(str_split(temp, "\\|"), function(x) if(length(x) >= 7) x[7] else NA),
                       temp),
         temp = gsub("s__","", temp))
rownames(metaphlan_mat2) <- metaphlan_mat2$temp
metaphlan_mat2$temp <- NULL

# Transpose taxonomic data so rows = samples, cols = taxa features
metaphlan_samp <- t(metaphlan_mat2)

##load preterm shortbred output
arg.full <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/250905_P4_GM_Dev_ShortBRED_merged_output_n2585.txt"))
arg.full <- arg.full %>% mutate(Sample_ID_R_final = gsub("-", ".", paste0("p4_", Sample)))
arg.p4 <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251022_GM_Dev_P4_shortBRED_joint_output.txt"))
arg.p4 <- arg.p4 %>% mutate(Sample_ID_R_final = paste0("p4_", gsub("_", ".", Sample)))
arg_df <- rbind(arg.full, arg.p4)

arg.wide <- as.data.frame(dcast(arg_df, Sample_ID_R_final ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg.wide) <- arg.wide$Sample_ID_R_final
arg.wide <- arg.wide[, c(-1)]

##sum counts to get RPKM per sample
arg.wide$RPKM <- rowSums(arg.wide)

##get ARG richness
arg.wide$Richness <- specnumber(arg.wide[, 1:ncol(arg.wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg.wide$Sample_ID_R_final <- row.names(arg.wide)

## load near-term shortbred output
arg_term <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251016_DrewG_term_n162_shortbred_merged.txt"))

arg_term_wide <- as.data.frame(dcast(arg_term, Sample ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg_term_wide) <- arg_term_wide$Sample
arg_term_wide <- arg_term_wide[, c(-1)]

##sum counts to get RPKM per sample
arg_term_wide$RPKM <- rowSums(arg_term_wide)

##get ARG richness
arg_term_wide$Richness <- specnumber(arg_term_wide[, 1:ncol(arg_term_wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg_term_wide$Sample <- row.names(arg_term_wide)
arg_term_wide <- arg_term_wide %>% dplyr::rename(Sample_ID_R_final = Sample) %>% dplyr::rename(ShortBRED_richness = Richness)

#### Merge preterm and near-term shortbred outputs
arg_full <- rbind(arg.wide, arg_term_wide)

# Merge with metadata
ps_infants_ga_meta_ids <- ps_infants_ga_meta %>% select(Sample_ID_R_final)
arg_full.df <- merge(ps_infants_ga_meta_ids, arg_full, by = "Sample_ID_R_final") 
rownames(arg_full.df) <- arg_full.df$Sample_ID_R_final

### Build combined feature matrix (samples x features)
X <- cbind(metaphlan_samp, arg_full.df)
X$Sample_ID_R_final <- NULL
#X <- merge(metaphlan_samp, arg_full, by = "row.names")
dim(X)

## Match order with metadata
meta_sub <- ps_infants_ga_meta %>%
  arrange(match(Sample_ID_R_final, rownames(X)))

## --------------------------------------------------
## 3. Log transform & remove globally zero-variance features
## --------------------------------------------------
pseudocount <- 1e-6
X_log <- log10(X + pseudocount)

global_sd <- apply(X_log, 2, sd, na.rm = TRUE)
X_log <- X_log[, global_sd > 0, drop = FALSE]

## --------------------------------------------------
## 4. Build modeling data frame & train/test split
## --------------------------------------------------
# Create antibiotic exposed vs unexposed vs term class label
meta_sub <- meta_sub %>% mutate(abx_group_simple = case_when(abx_group == "Early only" ~ "Preterm_abx_exposed",
                                                             abx_group == "Early and after" ~ "Preterm_abx_exposed",
                                                             abx_group == "None" ~ "Preterm_abx_naive",
                                                             abx_group == "None Near-term" ~ "Near-term_abx_naive")) %>%
  mutate(abx_group_simple = factor(abx_group_simple, levels = c("Preterm_abx_exposed", "Preterm_abx_naive", "Near-term_abx_naive"))) %>%
  mutate(term_group = case_when(abx_group == "Early only" ~ "Preterm",
                                 abx_group == "Early and after" ~ "Preterm",
                                 abx_group == "None" ~ "Preterm",
                                 abx_group == "None Near-term" ~ "Near-term")) %>%
  mutate(term_group = factor(term_group, levels = c("Preterm", "Near-term")))
# Make class labels a factor (ensure they’re syntactically valid)

#meta_sub$Group <- factor(make.names(meta_sub$abx_group))
#meta_sub$Group <- factor(make.names(meta_sub$abx_group_simple))
meta_sub$Group <- factor(make.names(meta_sub$term_group))

# Combine the response (Group) with the scaled features into a single data frame
# Group goes in the first column, all features after
svm_df <- data.frame(
  Group = meta_sub$Group,
  X_log,
  check.names = FALSE
)

set.seed(123)  # For reproducibility of the split

# Stratified split: create training indices, keeping class proportions
# p = 0.6 means 60% of samples go to the training set
train_idx <- createDataPartition(svm_df$Group, p = 0.6, list = FALSE)
train_df <- svm_df[train_idx, ]
test_df  <- svm_df[-train_idx, ]

train_x <- train_df[, -1, drop = FALSE]
train_y <- train_df$Group
test_x  <- test_df[,  -1, drop = FALSE]
test_y  <- test_df$Group

## --------------------------------------------------
## 5. Remove near-zero-variance predictors (training-based)
## --------------------------------------------------
nzv_idx <- nearZeroVar(train_x)
if (length(nzv_idx) > 0) {
  train_x <- train_x[, -nzv_idx, drop = FALSE]
  test_x  <- test_x[,  -nzv_idx, drop = FALSE]
}

train_df_clean <- data.frame(Group = train_y, train_x, check.names = FALSE)
test_df_clean  <- data.frame(Group = test_y,  test_x,  check.names = FALSE)
write.csv(train_df_clean, 
          "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/SVM/term_group/260110_term_group_SVM_train_dataset.csv", row.names = T)
write.csv(test_df_clean, "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/SVM/term_group/260110_term_group_SVM_test_dataset.csv", row.names = T)

## --------------------------------------------------
## 6. Parallel backend for caret
## --------------------------------------------------
n_cores <- max(1, parallel::detectCores() - 2)
cl <- makeCluster(n_cores)
registerDoParallel(cl)

cat("Using", n_cores, "cores\n")

## --------------------------------------------------
## 7. Initial linear SVM (for feature importance)
##    (linear kernel so we can interpret “weights”)
## --------------------------------------------------
ctrl_all <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = defaultSummary,
  allowParallel = TRUE
)

set.seed(123)
svm_full <- train(
  Group ~ .,
  data = train_df_clean,
  method = "svmLinear",
  trControl = ctrl_all,
  preProcess = c("center", "scale"),
  tuneLength = 5,
  metric = "Accuracy"
)

## --------------------------------------------------
## 8. Rank features by importance from full linear SVM
## --------------------------------------------------
imp_full <- varImp(svm_full, scale = TRUE)$importance
imp_full$Feature <- rownames(imp_full)

# For binary classification, caret usually gives an "Overall" column
if ("Overall" %in% colnames(imp_full)) {
  imp_full$Weight <- imp_full$Overall
} else {
  # in case of multiple class-specific columns, average them
  imp_full$Weight <- rowMeans(imp_full[, setdiff(colnames(imp_full), "Feature"), drop = FALSE])
}

imp_full <- imp_full %>%
  arrange(desc(Weight))

## --------------------------------------------------
## 9. Feature-selection: smallest top-K set with CV accuracy ≥ 0.90
##    (search over top N candidates to keep runtime reasonable)
## --------------------------------------------------
max_candidates <- min(200, nrow(imp_full))  # train on the top 200 most important features
candidate_feats <- imp_full$Feature[1:max_candidates] # Extract names of these top features

k_values <- seq(5, max_candidates, by = 5) # test k = 5, 10, 15, ... up to max_candidates to reduce runtime

# Initialize variables to store
best_k <- NA # the final optimal number of features
best_acc <- -Inf # accuracy achieved at best_k
acc_by_k <- data.frame(k = integer(), Accuracy = numeric()) # table storing accuracy results for each tested K

for (k in k_values) {
  feats_k <- candidate_feats[1:k]
  train_k <- train_df_clean[, c("Group", feats_k), drop = FALSE] # Build a training dataframe containing the outcome variable and only the top k predictors
  
  set.seed(123 + k) # Each K gets a slightly different RNG seed to avoid fold repetition bias.
  svm_k <- train(
    Group ~ .,
    data = train_k,
    method = "svmLinear",
    trControl = ctrl_all,
    preProcess = c("center", "scale"),
    tuneLength = 3,
    metric = "Accuracy"
  )
  
  acc_k <- max(svm_k$results$Accuracy)
  acc_by_k <- rbind(acc_by_k, data.frame(k = k, Accuracy = acc_k))
  
  if (acc_k > best_acc) {
    best_acc <- acc_k
    best_k <- k
  }
  
  if (acc_k >= 0.90) {
    best_k <- k
    best_acc <- acc_k
    break
  }
}

cat("Chosen K (sparse model) =", best_k, "with CV Accuracy =", round(best_acc, 3), "\n") ## best_k = 190, accuracy = 0.586

## --------------------------------------------------
## 10. Train final sparse SVM on top-K features
## --------------------------------------------------
top_feats_sparse <- candidate_feats[1:best_k]

train_sparse <- train_df_clean[, c("Group", top_feats_sparse), drop = FALSE]
test_sparse  <- test_df_clean[, c("Group", top_feats_sparse), drop = FALSE]

set.seed(999)
svm_sparse <- train(
  Group ~ .,
  data = train_sparse,
  method = "svmLinear",
  trControl = ctrl_all,
  preProcess = c("center", "scale"),
  tuneLength = 5,
  metric = "Accuracy"
)

stopCluster(cl)
registerDoSEQ()

## --------------------------------------------------
## 11. Predict held-out test set with sparse SVM
## --------------------------------------------------
pred_test <- predict(svm_sparse, newdata = test_sparse)
cm <- confusionMatrix(pred_test, test_sparse$Group)
cm

## --------------------------------------------------
## 12. Plot (a) Confusion matrix heatmap
## --------------------------------------------------
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

gg_cm <- ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradient(low = "#d9f0d3", high = "#1a9850") +
  scale_x_discrete(position = "top") +
  labs(x = "Actual", y = "Predicted", fill = "Count") +
  coord_equal() +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.margin = margin(5, 10, 5, 5)
  )
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/SVM/term_group/260110_GMdev_SVM_term_group_confusion_matrix_validation.pdf", plot = gg_cm, bg = "transparent", width = 10, height = 8)

## --------------------------------------------------
## 13. Plot (b) Bar plot of sparse-feature “weights”
## --------------------------------------------------
imp_sparse <- varImp(svm_sparse, scale = TRUE)$importance
imp_sparse$Feature <- rownames(imp_sparse)

if ("Overall" %in% colnames(imp_sparse)) {
  imp_sparse$Weight <- imp_sparse$Overall
} else {
  imp_sparse$Weight <- rowMeans(imp_sparse[, setdiff(colnames(imp_sparse), "Feature"), drop = FALSE])
}

imp_sparse <- imp_sparse %>%
  arrange(desc(Weight))

# Take all sparse features, or e.g. top 15 as in the figure
n_top_plot <- min(20, nrow(imp_sparse))
imp_top20 <- imp_sparse[1:n_top_plot, ]

gg_weights <- ggplot(imp_top20,
                     aes(x = Weight,
                         y = reorder(Feature, Weight))) +
  geom_col() +
  labs(x = "Weight", y = "Feature") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5, 5, 5, 10)
  )
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/SVM/term_group/260110_GMdev_term_group_SVM_feature_importance_top20.pdf", plot = gg_weights, bg = "transparent", width = 20, height = 8)

## ---- Per-group metrics from confusion matrix ----

# Confusion matrix as a plain matrix
tab <- as.matrix(cm$table)
classes <- colnames(tab)  # rows = true classes (reference)
# Total number of test samples
N <- sum(tab)


############ Change################
perf_per_class <- lapply(classes, function(cls) {
  # TP = predicted cls AND actual cls (diagonal element for cls)
  TP <- tab[cls, cls]
  
  # FN = actual cls but predicted not cls (rest of the column)
  FN <- sum(tab[, cls]) - TP
  
  # FP = predicted cls but actual not cls (rest of the row)
  FP <- sum(tab[cls, ]) - TP
  
  # TN = everything else
  TN <- N - TP - FN - FP
  
  Accuracy    <- (TP + TN) / N
  Sensitivity <- ifelse((TP + FN) == 0, NA_real_, TP / (TP + FN))
  Specificity <- ifelse((TN + FP) == 0, NA_real_, TN / (TN + FP))
  
  data.frame(
    Group = cls,
    TP = TP, TN = TN, FP = FP, FN = FN,
    Accuracy = Accuracy,
    Sensitivity = Sensitivity,
    Specificity = Specificity,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows()
##################################

# Long format: one row per (Group, Metric)
perf_long <- perf_per_class %>%
  pivot_longer(
    cols = c(Accuracy, Sensitivity, Specificity),
    names_to = "Metric",
    values_to = "Value"
  )

gg_perf <- ggplot(
  perf_long,
  aes(x = Group, y = Value, fill = Metric)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    aes(label = sprintf("%.2f", Value)),
    position = position_dodge(width = 0.8),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Group",
    y = "Performance (proportion)",
    fill = "Metric",
    title = "Sparse SVM performance on held-out test set"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

gg_perf_faceted <- ggplot(
  perf_long,
  aes(x = Group, y = Value, fill = Group)
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = sprintf("%.2f", Value)),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  facet_wrap(~ Metric, nrow = 1) +
  labs(
    x = "Group",
    y = "Performance (proportion)",
    title = "Sparse SVM performance by metric and group"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
ggsave("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/SVM/term_group/260110_GMdev_term_group_SVM_sparse_model_performance.pdf", plot = gg_perf_faceted, bg = "transparent", width = 10, height = 6)


save.image("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/SVM/PMA_25to32weeks/abx_group_simple/260107_GM_dev_SVM.RData")


################ Old code #############


# By-class metrics from caret (Sensitivity, Specificity, etc.)
byc <- as.data.frame(cm$byClass)

# If caret returns a single row (binary), coerce properly
if (nrow(byc) == 1 && length(classes) == 2) {
  rownames(byc) <- classes
}



# Compute per-class accuracy manually:
# accuracy_k = (TP + TN) / N for each class k
acc_vec <- numeric(length(classes))

for (i in seq_along(classes)) {
  print(i)
  TP <- tab[i, i]
  print(paste("TP:", TP))
  FN <- sum(tab[i, ]) - TP
  print(paste("FN:", FN))
  FP <- sum(tab[, i]) - TP
  TN <- N - TP - FP - FN
  acc_vec[i] <- (TP + TN) / N
}

# Combine into a summary data frame
perf_per_class <- data.frame(
  Group       = classes,
  Accuracy    = acc_vec,
  Sensitivity = byc$Sensitivity,
  Specificity = byc$Specificity,
  row.names   = NULL
)
```

GM Shift Analysis
```{r}
library(ggpubr)

samples_per_infant <- merged_ps_meta %>% group_by(Patient) %>% dplyr::summarise(n_samples = n())

n_single_sample <- sum(samples_per_infant$n_samples == 1)

pt_long_meta <- merged_ps_meta %>% group_by(Patient) %>% filter(n() > 1) %>% ungroup()

long_pt_ps <- merged_ps %>% subset_samples(Sample_ID_R_final %in% pt_long_meta$Sample_ID_R_final)
 
pt_bc_matrix <- phyloseq::distance(long_pt_ps, method="bray")
pt_bc_matrix <- as.matrix(pt_bc_matrix)
pt_bc_matrix <- melt(pt_bc_matrix)

colnames(pt_bc_matrix)[1] <- "Sample_ID1"
colnames(pt_bc_matrix)[2] <- "Sample_ID2"
colnames(pt_bc_matrix)[3] <- "bray"

# Removing duplicated rows
pt_bc_matrix_uniq <- pt_bc_matrix[!duplicated(apply(pt_bc_matrix, 1, function(x) paste(sort(x), collapse = ""))),]
# Removing self comparisons with values of zero
pt_bc_matrix_uniq <- pt_bc_matrix_uniq[which(pt_bc_matrix_uniq$Sample_ID1 != pt_bc_matrix_uniq$Sample_ID2),] 
# Subset metadata to columns I want
pt_long_want <- pt_long_meta %>% select(Sample_ID_R_final, Patient, DOL, abx_group_updated, abx_group_names)

# Merge final distance matrix with metadata
pt_bc_matrix_uniq <- merge(pt_bc_matrix_uniq, pt_long_want, by.x = "Sample_ID1", by.y = "Sample_ID_R_final")
pt_bc_matrix_uniq <- merge(pt_bc_matrix_uniq, pt_long_want, by.x = "Sample_ID2", by.y = "Sample_ID_R_final")

### Comparing within and between participant distances 
# Adding comparison type column 
pt_bc_matrix_uniq$Comparison[pt_bc_matrix_uniq$Patient.x == pt_bc_matrix_uniq$Patient.y] <- "Within participant"
pt_bc_matrix_uniq$Comparison[pt_bc_matrix_uniq$Patient.x != pt_bc_matrix_uniq$Patient.y] <- "Between participant"

pt_bc_matrix_uniq$Comparison <- factor(pt_bc_matrix_uniq$Comparison, levels = c("Between participant","Within participant"))

## Create df_sample
df_sample <- pt_long_meta %>% arrange(Patient, DOL)

## Create df_within and df_between
df_within = filter(pt_bc_matrix_uniq, Comparison=="Within participant")
df_between = filter(pt_bc_matrix_uniq, Comparison=="Between participant")

### Plot saturation curve with all sample pairs (not just consecutive)
## Calculate DOL differences between intra-patient samples
df_within <- df_within %>% mutate(DOL_diff = abs(DOL.x - DOL.y))

sat_curve_data_all <- df_within %>%
  group_by(DOL_diff) %>%
  summarise(mean_BC = mean(bray, na.rm = TRUE),
            median_BC = median(bray, na.rm = TRUE),
            .groups = 'drop') %>%
  arrange(DOL_diff)

sat_curve_all <- ggplot(sat_curve_data_all, aes(x = DOL_diff, y = mean_BC)) +
  geom_point(size = 2) +
  geom_line() +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
  labs(x = "Time difference (DOL)",
       y = "Bray-Curtis Dissimilarity") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        axis.line = element_line(colour = "black"),
        panel.background = element_blank())
ggsave("260211_GMdev_pterm_mean_bray_bw_all_within_infant_samples_across_DOL_diff.pdf", plot = sat_curve_all, bg = "transparent", width = 10, height = 6)


## Identify GM shifts
### First identify samples that are consecutive from each infant
df_within_consecutive = data.frame()
for (patient in unique(df_sample$Patient)) {
  # Get all samples for this patient (already sorted by DOL)
  sample_info <- filter(df_sample, Patient == patient)
  
  # Loop through consecutive pairs
  for (i in 2:nrow(sample_info)) {
    sample_prev <- sample_info$Sample_ID_R_final[i-1]
    sample_curr <- sample_info$Sample_ID_R_final[i]
    # Find this pair in df_within (check both orders)
    pair_data <- df_within %>%
      filter((Sample_ID1 == sample_prev & Sample_ID2 == sample_curr) |
             (Sample_ID2 == sample_prev & Sample_ID1 == sample_curr))
    
    if (nrow(pair_data) > 0) {
      df_within_consecutive <- rbind(df_within_consecutive, pair_data)
    }
  }
}
### 2069 consecutive pairs of samples found ##

##### Plot bray-curtis dissimilarity by DOL diff for consecutive sample pairs
# First, calculate mean BC for each DOL_diff
bc_sat_data <- df_within_consecutive %>%
  group_by(DOL_diff) %>%
  summarise(
    mean_BC = mean(bray, na.rm = TRUE),
    median_BC = median(bray, na.rm = TRUE),
    n = n(),
    .groups = 'drop'
  ) %>%
  arrange(DOL_diff)

sat_curve <- ggplot(bc_sat_data, aes(x = DOL_diff, y = mean_BC)) +
  geom_point(size = 2) +
  geom_line() +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
  labs(x = "Time difference (DOL)",
       y = "Bray-Curtis Dissimilarity") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        axis.line = element_line(colour = "black"),
        panel.background = element_blank())
ggsave("260211_GMdev_pterm_mean_bray_bw_consecutive_samples_across_DOL_diff.pdf", plot = sat_curve, bg = "transparent", width = 10, height = 6)

#### Calculate DOL difference for between-patient data
df_between <- df_between %>% mutate(DOL_diff = abs(DOL.x - DOL.y)) %>% select(Sample_ID1, everything())
## Check distribution of DOL differences
table(df_within_consecutive$DOL_diff)

############-----------Determine shift thresholds for each DOL difference------------#################
## Reorder df_within_consecutive just for ease of keeping track of samples and DOL
df_within_consecutive <- df_within_consecutive %>% select(Sample_ID1, everything())

# Determine the range of DOL differences to analyze
# Use only DOL_diffs that have sufficient data
dol_diff_counts <- table(df_within_consecutive$DOL_diff)
print(dol_diff_counts)

# Choose range (e.g., 0-10 days, or based on your data)
# Make sure each DOL_diff has enough samples (e.g., >5 pairs)
DOL_diff_range <- 0:10  # Adjust based on your data

# Initialize results dataframe
shift_thresholds <- data.frame(
  DOL_diff = DOL_diff_range,
  Cutoff = NA
)

# Calculate intersection point for each DOL difference
for (i in 1:nrow(shift_thresholds)) {
  dol_val <- shift_thresholds$DOL_diff[i]
  
  # Combine data for DOL_diff = i and i+1 (for smoothing)
  if (dol_val == max(DOL_diff_range)) {
    # For the last value, only use that value
    intra_BC <- df_within_consecutive$bray[
      df_within_consecutive$DOL_diff == dol_val]
    inter_BC <- df_between$bray[
      df_between$DOL_diff == dol_val]
  } else {
    intra_BC <- df_within_consecutive$bray[
      df_within_consecutive$DOL_diff == dol_val | 
      df_within_consecutive$DOL_diff == dol_val + 1]
    inter_BC <- df_between$bray[
      df_between$DOL_diff == dol_val | 
      df_between$DOL_diff == dol_val + 1]
  }
  
  # Skip if insufficient data
  if (length(intra_BC) < 5 | length(inter_BC) < 5) {
    warning(paste("Insufficient data for DOL_diff =", dol_val))
    next
  }
  
  # Calculate density distributions
  all_BC <- c(intra_BC, inter_BC)
  lower.limit <- min(all_BC)
  upper.limit <- max(all_BC)
  
  density_intra <- density(intra_BC, from = lower.limit, to = upper.limit)
  density_inter <- density(inter_BC, from = lower.limit, to = upper.limit, adjust = 10)
  
  # Find intersection point
  density_diff <- density_intra$y - density_inter$y
  crossings <- which(diff(density_diff > 0) != 0) + 1
  
  if (length(crossings) > 0) {
    shift_thresholds$Cutoff[i] <- min(density_intra$x[crossings])
  } else {
    warning(paste("No intersection found for DOL_diff =", dol_val))
  }
}

####### Visualize threshold vs DOL difference
# Plot cutoff vs DOL difference with regression
cutoff_plot <- ggplot(shift_thresholds, aes(x = DOL_diff, y = Cutoff)) +
  geom_point(size = 3) +
  geom_smooth(method = lm, color = "red", fill = "#69b3a2", se = TRUE) +
  stat_cor(label.y = max(shift_thresholds$Cutoff, na.rm = TRUE) * 0.95) + 
  stat_regline_equation(label.y = max(shift_thresholds$Cutoff, na.rm = TRUE) * 0.90) +
  labs(x = "DOL Difference", y = "Bray-Curtis Cutoff") +
  theme_classic() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
ggsave("shift_cutoff_plot.pdf", cutoff_plot, width = 5, height = 4)

###### Fit regression model to predict thresholds ##########
# Fit linear model
cutoff_model <- lm(Cutoff ~ DOL_diff, data = shift_thresholds)
summary(cutoff_model)

# Extract coefficients
intercept <- coef(cutoff_model)[1]
slope <- coef(cutoff_model)[2]

print(paste("Threshold formula: Cutoff =", round(intercept, 4), "+", 
            round(slope, 4), "* DOL_diff"))

####### Apply thresholds to identify shift events##############
# Calculate expected threshold for each consecutive pair based on its DOL_diff
df_within_consecutive$Expected_Cutoff <- intercept + slope * df_within_consecutive$DOL_diff

# Determine if it's a shift event
df_within_consecutive$Shift <- ifelse(
  df_within_consecutive$bray >= df_within_consecutive$Expected_Cutoff,
  "Yes", "No")

# Count shift events
table(df_within_consecutive$Shift)

# View some shift events
shift_events <- df_within_consecutive %>% filter(Shift == "Yes")
print(paste("Total shift events:", nrow(shift_events)))
head(shift_events)

# Determine which sample is the "later" one (higher DOL)
df_within_consecutive$Sample_later <- if_else(
  df_within_consecutive$DOL.x > df_within_consecutive$DOL.y, df_within_consecutive$Sample_ID1, df_within_consecutive$Sample_ID2)
df_within_consecutive$DOL_later <- pmax(df_within_consecutive$DOL.x, df_within_consecutive$DOL.y)

df_within_consecutive$Sample_earlier <- if_else(
  df_within_consecutive$DOL.x < df_within_consecutive$DOL.y, df_within_consecutive$Sample_ID1, df_within_consecutive$Sample_ID2)
df_within_consecutive$DOL_earlier <- pmin(df_within_consecutive$DOL.x, df_within_consecutive$DOL.y)

df_within_consecutive$abx_group_earlier <- if_else(
  df_within_consecutive$DOL.x < df_within_consecutive$DOL.y, df_within_consecutive$abx_group_names.x, df_within_consecutive$abx_group_names.y)
df_within_consecutive$abx_group_later <- if_else(
  df_within_consecutive$DOL.x > df_within_consecutive$DOL.y, df_within_consecutive$abx_group_names.x, df_within_consecutive$abx_group_names.y)

write.csv(df_within_consecutive, "260210_within_infant_consecutive_with_shifts.csv", row.names = F)

### Create a transition label for shifts that occur between abx exposure groups
df_within_consecutive$abx_group_transition <- paste0(df_within_consecutive$abx_group_earlier, " -> ", df_within_consecutive$abx_group_later)

## Categorize as within-group or between-group change in abx_group
df_within_consecutive$transition_type <- if_else(
  df_within_consecutive$abx_group_earlier == df_within_consecutive$abx_group_later, "Within-group", "Between-group")


######### Visualize shift events #############
library(ggridges)
library(ggalluvial)
library(ggpubr)

# Filter for shift events only
shift_events <- df_within_consecutive %>% filter(Shift == "Yes")

# Count total shifts by transition type
transition_summary <- shift_events %>%
  group_by(transition_type, abx_group_transition) %>%
  summarise(
    n_shifts = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_shifts))

# Count within-group shifts
within_group_shifts <- shift_events %>%
  filter(transition_type == "Within-group") %>%
  group_by(abx_group_earlier) %>%
  summarise(
    n_shifts = n(),
    mean_later_DOL = mean(DOL_later, na.rm = TRUE),
    median_later_DOL = median(DOL_later, na.rm = TRUE),
    .groups = 'drop')

# Count between-group shifts
between_group_shifts <- shift_events %>%
  filter(transition_type == "Between-group") %>%
  group_by(abx_group_earlier, abx_group_later) %>%
  summarise(
    n_shifts = n(),
    mean_later_DOL = mean(DOL_later, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_shifts))

# Calculate total consecutive pairs (shifts + non-shifts) for each transition
all_pairs_summary <- df_within_consecutive %>%
  group_by(transition_type, abx_group_transition) %>%
  summarise(
    total_pairs = n(),
    n_shifts = sum(Shift == "Yes"),
    n_no_shifts = sum(Shift == "No"),
    shift_proportion = n_shifts / total_pairs,
    .groups = 'drop'
  ) %>%
  arrange(desc(shift_proportion))

## Bar plot of shift counts ##
all_pairs_summary$abx_group_transition <- factor(all_pairs_summary$abx_group_transition, 
                                                 levels = c("EAS -> EAS", "EEO -> EEO", "ABN -> ABN", "EEO -> EAS"))
plot1 <- ggplot(all_pairs_summary, aes(x = abx_group_transition, y = n_shifts, 
                                        fill = transition_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n_shifts), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Within-group" = "#3291a4", "Between-group" = "#f28a5f")) +
  labs(x = "Antibiotic Exposure Group",
    y = "Number of Shift Events",
    fill = "Transition Type"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 14))
ggsave("260211_GM_shift_counts_by_abx_exposure_group.pdf", plot1, width = 8, height = 6)

# Stacked bar showing shift vs no-shift proportions
plot2 <- ggplot(all_pairs_summary, aes(x = abx_group_transition, y = shift_proportion, 
                                        fill = transition_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(shift_proportion * 100, 1), "%")), 
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Within-group" = "#3291a4", "Between-group" = "#f28a5f")) +
  labs(
    title = "Proportion of Consecutive Pairs that are Shifts",
    x = "Antibiotic Exposure Group",
    y = "Shift Proportion",
    fill = "Transition Type"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 14))+
  ylim(0, max(all_pairs_summary$shift_proportion) * 1.1)
ggsave("260211_GM_shift_proportions_by_abx_exposure_group.pdf", plot2, width = 8, height = 6)

#### Chi-square test for shift proportions
## Test if shift proportions differ among within-group transitions
# Create contingency table
within_contingency <- df_within_consecutive %>% filter(transition_type == "Within-group") %>%
  select(abx_group_earlier, Shift) %>% table()

# Chi-square test
chisq_within <- chisq.test(within_contingency) # X-squared = 7.0556, df = 2, p-value = 0.02937

# Pairwise comparison since the overall chisq test was significant
pairwise_prop_test_results <- pairwise.prop.test(x = within_contingency[, "Yes"],
                                       n = rowSums(within_contingency),
                                       p.adjust.method = "bonferroni") ## Within stable pairs, the EAS group has significantly higher number of shifts than the EEO group (q = 0.034)

################ Test shift rates between all four groups
df_within_consecutive <- df_within_consecutive %>% mutate(comparison_type = paste0(transition_type,"_",abx_group_later))
transition_contingency <- df_within_consecutive %>% select(comparison_type, Shift) %>% table()

# Chi-square test
chisq_bw <- chisq.test(transition_contingency) # X-squared = 9.381, df = 3, p-value = 0.025

# Pairwise comparison since the overall chisq test was significant
pairwise_results_bw <- pairwise.prop.test(x = transition_contingency[, "Yes"],
                                       n = rowSums(transition_contingency),
                                       p.adjust.method = "BH") ## Within stable pairs, the EAS group has significantly higher number of shifts than the EEO group (q = 0.034)

################ Test if between-group transitions have different shift rates than within-group
transition_contingency2 <- df_within_consecutive %>% select(transition_type, Shift) %>% table()

chisq_transition <- chisq.test(transition_contingency2) # Not significant; X-squared = 1.8215, df = 1, p-value = 0.1771 (Pearson's Chi-squared test with Yates' continuity correction)

################ Test if shift timing differs among abx exposure groups
kruskal_timing <- kruskal.test(DOL_later ~ abx_group_later, data = shift_events) #Kruskal-Wallis chi-squared = 90.729, df = 2, p-value < 2.2e-16


# Post-hoc pairwise comparisons since significant
library(FSA)
posthoc_timing <- dunnTest(DOL_later ~ abx_group_later, 
                             data = shift_events, 
                             method = "bonferroni")

# Compare timing between within-group and between-group shifts
wilcox_transition_timing <- wilcox.test(DOL_later ~ transition_type, data = shift_events)

### Check DOL distributions by abx group for all samples in this shift analysis
# Get all unique samples involved in shift events
# (both the earlier and later samples from each shift pair)
samples_in_shifts <- unique(c(shift_events$Sample_earlier, shift_events$Sample_later)) #465 samples

# Filter df_sample to only those samples
samples_shift_analysis <- df_sample %>%
  filter(Sample_ID_R_final %in% samples_in_shifts)

table(samples_shift_analysis$abx_group_names)

# Calculate summary statistics for DOL by treatment group
dol_summary <- samples_shift_analysis %>%
  group_by(abx_group_names) %>%
  summarise(
    n_samples = n(),
    mean_DOL = mean(DOL, na.rm = TRUE),
    median_DOL = median(DOL, na.rm = TRUE),
    sd_DOL = sd(DOL, na.rm = TRUE),
    min_DOL = min(DOL, na.rm = TRUE),
    max_DOL = max(DOL, na.rm = TRUE),
    q25_DOL = quantile(DOL, 0.25, na.rm = TRUE),
    q75_DOL = quantile(DOL, 0.75, na.rm = TRUE),
    .groups = 'drop')

# Ridge plot showing DOL distribution for each treatment group
plot_ridge <- ggplot(samples_shift_analysis, aes(x = DOL, y = abx_group_names, fill = abx_group_names)) +
  geom_density_ridges(alpha = 0.6, scale = 0.9, quantile_lines = TRUE, quantiles = 2) +
  scale_fill_manual(values = custom_colors_names) +
  labs(
    title = "Distribution of Sample Collection DOL by Antibiotic Exposure Group",
    subtitle = "Samples included in shift event analysis",
    x = "Day of Life (DOL)",
    y = "Treatment Group"
  ) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none")
ggsave("260211_DOL_distribution_of_GM_shift_samples_by_group_ridge.pdf", plot_ridge, width = 8, height = 6)

# Violin plot with statistical comparisons
plot_violin <- ggplot(samples_shift_analysis, aes(x = abx_group_names, y = DOL, fill = abx_group_names)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = custom_colors_names) +
  stat_compare_means(method = "kruskal.test", label.y = max(samples_shift_analysis$DOL) * 1.05) +
  stat_compare_means(comparisons = list(c("EAS", "EEO"), 
                                        c("EAS", "ABN"), 
                                        c("EEO", "ABN")),
                     method = "wilcox.test",
                     label = "p.signif") +
  labs(
    title = "DOL distribution of samples involved in shift events",
    x = "Antibiotic Exposure Group",
    y = "Day of Life (DOL)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 14),
    legend.position = "none")
ggsave("260211_shift_samples_DOL_distribution_by_group_violin.pdf", plot_violin, width = 8, height = 6)


# Ridge plot showing WHEN shifts occur for each abx exposure group
plot4 <- ggplot(shift_events, aes(x = DOL_later, y = abx_group_later, 
                                   fill = abx_group_later)) +
  geom_density_ridges(alpha = 0.6, scale = 0.9, quantile_lines = TRUE, quantiles = 2) +
  scale_fill_manual(values = custom_colors_names) +
  labs(
    title = "Timing of Shift Events by Antibiotic Exposure Group (After Shift)",
    x = "DOL Later (When Shift Occurred)",
    y = "Antibiotic Exposure Group (After Shift)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 14),
    legend.position = "none"
  ) +
  xlim(0, max(shift_events$DOL_later, na.rm = TRUE))
ggsave("260211_shift_timing_by_abx_exposure_group.pdf", plot4, width = 8, height = 6)

# Show timing separately for within-group vs between-group shifts
plot5 <- ggplot(shift_events, aes(x = DOL_later, y = abx_group_later, 
                                   fill = transition_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.9, quantile_lines = TRUE, quantiles = 2) +
  scale_fill_manual(values = c("Within-group" = "#3291a4", "Between-group" = "#f28a5f")) +
  facet_wrap(~transition_type, ncol = 1) +
  labs(
    title = "Shift Timing: Within-Group vs Between-Group Transitions",
    x = "DOL Later (When Shift Occurred)",
    y = "Treatment Group (After Shift)",
    fill = "Transition Type"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 14))
ggsave("260211_shift_timing_by_transition_type_and_abx_exposure_group.pdf", plot5, width = 8, height = 8)

## Compare shift timings between groups in plot5
shift_events <- shift_events %>% mutate(Comparison_group = paste0(transition_type, "_", abx_group_later))
table(shift_events$Comparison_group)
shift_events$Comparison_group <- factor(shift_events$Comparison_group)

# Pairwise wilcoxon tests
shift_time_wilcox <- pairwise.wilcox.test(x = shift_events$DOL_later,
                                          g = shift_events$Comparison_group,
                                          p.adjust.method = "BH", paired = FALSE)

## Pairwise permutation tests
library(coin)

# 1. Permutation test (best for many ties)
# Overall test
overall_perm <- independence_test(
  DOL_later ~ factor(Comparison_group),
  data = shift_events,
  distribution = approximate(nresample = 10000)
)

# Pairwise with correction
groups <- unique(shift_events$Comparison_group)
pairwise_results <- data.frame()

for (i in 1:(length(groups)-1)) {
  for (j in (i+1):length(groups)) {
    subset_data <- shift_events %>%
      filter(Comparison_group %in% c(groups[i], groups[j]))
    
    perm_test <- independence_test(
      DOL_later ~ factor(Comparison_group),
      data = subset_data,
      distribution = approximate(nresample = 10000)
    )
    
    pairwise_results <- rbind(pairwise_results, data.frame(
      Comparison = paste(groups[i], "vs", groups[j]),
      p_value = pvalue(perm_test)
    ))
  }
}

pairwise_results$p_adjusted <- p.adjust(pairwise_results$p_value, "bonferroni")
pairwise_results$significant <- pairwise_results$p_adjusted < 0.05


# 2. Report summary statistics
summary_stats <- shift_events %>%
  group_by(Comparison_group) %>%
  summarise(
    n = n(),
    median = median(DOL_later),
    mean = mean(DOL_later),
    sd = sd(DOL_later),
    IQR = IQR(DOL_later)
  )


# Box plot with permutation results
custom_colors_shift <- c("Between-group_EAS" = "#f28a5f", "Within-group_ABN" = "#8E949F", "Within-group_EAS" = "#AAC9ED",
                         "Within-group_EEO" = "#44637D")
shift_events$Comparison_group <- factor(shift_events$Comparison_group, levels = c("Within-group_ABN", "Within-group_EEO", "Within-group_EAS", "Between-group_EAS"))
plot_timing_comparison <- ggplot(shift_events, aes(x = Comparison_group, y = DOL_later, fill = Comparison_group)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = custom_colors_shift) +
  labs(
    title = "Shift timing comparisons",
    x = "Antibiotic Exposure Group",
    y = "DOL Later (When Shift Occurred)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 14), 
    legend.position = "none")
ggsave("260212_shift_timing_distribution_by_group_violin.pdf", plot_timing_comparison, width = 10, height = 6)


##################### Visualize density distributions for a specific DOL difference ###########
# Example: Visualize densities for DOL_diff = 3
dol_example <- 10

intra_BC <- df_within_consecutive$bray[
  df_within_consecutive$DOL_diff == dol_example]
inter_BC <- df_between$bray[
  df_between$DOL_diff == dol_example]

plot_data <- data.frame(
  BC = c(intra_BC, inter_BC),
  Type = c(rep("Intra-patient", length(intra_BC)), 
           rep("Inter-patient", length(inter_BC)))
)

threshold_val <- shift_thresholds$Cutoff[shift_thresholds$DOL_diff == dol_example]

DOL10_density_plot <- ggplot(plot_data, aes(x = BC, fill = Type)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = threshold_val, linetype = "dashed", color = "#AD3C36", size = 1) +
  annotate("text", x = threshold_val, y = Inf, 
           label = paste("Threshold =", round(threshold_val, 3)), 
           hjust = -0.1, vjust = 2) +
  scale_fill_manual(values = c("Intra-patient" = "#AA6B77", "Inter-patient" = "#EACACF")) +
  labs(title = paste("Density Distributions for DOL_diff =", dol_example),
       x = "Bray-Curtis Dissimilarity",
       y = "Density") +
  theme_classic() + 
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14)) 
ggsave("260211_GMdev_pterm_DOL_diff10_threshold_plot.pdf", plot = DOL10_density_plot, bg = "transparent", width = 10, height = 6)


```

