---
title: "P4_Metagenomics_analyses_Final"
output: html_document
date: "2026-01-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq) # v1.50.0
library(vegan) # v.2.7.2
library(tibble) # v3.3.1 
library(stringr) # v1.6.0 
library(flextable) # v0.9.11
library(officer) # v0.7.3
library(tableone) # v0.13.2
library(rlang) # v1.1.7
library(dplyr) # v1.2.0
library(lme4) # v1.1.38
library(emmeans) # v2.0.1
library(rstatix) # v0.7.3
library(ggplot2) # v4.0.2
library(ggpubr) # v0.6.3
library(doParallel) # v1.0.17
library(foreach) # v1.5.2
library(permute) # v0.9.10
library(tidytext) # v0.4.3
library(ggh4x) # v0.3.1
library(Maaslin2) # v1.20.0
library(randomForest) # v4.7.1.2
library(NatParksPalettes) #v0.2.0

source("GM_development_functions.R")
source("~/Documents/GitHub/Preterm_Plasmidome/GM_development/customtab.R")

meta_data_path <- "~/Documents/GitHub/Preterm_Plasmidome/GM_development/260202_P4_GM_dev_meta_updated_groupings_v2.xlsx"

cat_var <- c("Sample_ID", "Patient", "Cohort", "abx_group", "BSI_exp_type", "Necrotizing_enterocolitis", "Received_antimicrobials", "Sequencing_runs","duplicate_count","Cohorts","Sex","Delivery_route","Maternal_ethnicity","Maternal_race","Site","Final_disposition",
             "Para", "Maternal_smoking",	"Maternal_alcohol",	"Maternal_substance_abuse",	"Maternal_preeclampsia",	"Maternal_diabetes",	"Maternal_chorioamnionitis","Maternal_hypertension", "mat_inpt_antimicrobe","mat_outpt_antimicrobial","Multiple_births", "PMA_bin")

num_var <- c("DOL","MOL","abx_exposure_count","Gestational_age_weeks", "Birthweight","Maternal_age_at_delivery","Gravida","antiviral_exposure", "aminoglycoside_exposure","penicillin_exposure","macrolide_exposure","firstgencephalosporin_exposure",
             "fourthgencephalosporin_exposure","thirdgencephalosporin_exposure","secondgencephalosporin_exposure","lincosamide_exposure",
             "antifungal_exposure","carbapenem_exposure","other_antimicrobials_exposure","other_non_antimicrobials_exposure","glycopeptide_exposure","med_exposure_count", "morphine_exposure", "fentanyl_exposure", "surfactant_exposure", "acetaminophen_exposure", "caffeine_exposure", "cholecalciferol_vit_d__exposure",	'vitamin_a_exposure',	'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	'bumetanide_bumax__exposure',	'other_exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	'sodium_bicarbonate_exposure','hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure','cum_days_formula','cum_days_milk','frac_days_formula', 'frac_days_milk')

custom_colors_abx <- c("None" = "#8E949F", "Early only" = "#44637D", "After only" = "#613921", "Early and after" = "#AAC9ED")
custom_colors_groups <- c("None" = "#8E949F", "Early only" = "#44637D", "Early and after" = "#AAC9ED", "Near-term" = "#613921")
custom_colors_names <- c("ABN" = "#8E949F", "EEO" = "#44637D", "EAS" = "#AAC9ED", "Near-term" = "#613921")
```

# Load data
```{r}
# Read the metaphlan3 table
mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/250318_P4_GM_dev_metaphlan3_merged_n2585_v2.txt", header = T, sep="\t", skip=1, check.names = FALSE))

# R adds an "X" to column names that start with a number. Steps to remove this below
#colnames(mphlan) <- gsub("^X([0-9])", "\\1", colnames(mphlan))
colnames(mphlan) <- as.character(colnames(mphlan))

metadata <- read_excel(meta_data_path)

## Processing data for phyloseq
rownames(mphlan) <- mphlan$clade_name
mphlan$clade_name <- NULL

## Create species data for phyloseq 
# keep rows with species info 
#mphlan.species <- mphlan
mphlan.species <- mphlan[grepl('s__', rownames(mphlan)), ]
# Drop rows with SBG (species level genome bins)
mphlan.species <- mphlan.species[!grepl('t__', rownames(mphlan.species)), ]
mphlan.species <- mphlan.species %>% dplyr::select_if(~sum(.) != 0)
mphlan.species <- as.matrix(mphlan.species)

## Create taxa table
species <- data.frame(Names = rownames(mphlan.species))
species <- data.frame(do.call('rbind', strsplit(as.character(species$Names), '|', fixed = TRUE)))
rownames(species) <- rownames(mphlan.species)
colnames(species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
species <- as.matrix(species)

## Metadata
rows_metadata <- rownames(metadata)[colnames(mphlan.species) %in% metadata$Sample_ID_R_final]
metadata <- data.frame(metadata)
rownames(metadata) <- metadata$Sample_ID_R_final

# Set categorical variables to factor
metadata[,cat_var] <- lapply(metadata[,cat_var], factor)
# Set numerical variables to numeric
metadata[,num_var] <- lapply(metadata[,num_var], as.numeric)
print(dim(metadata))

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(mphlan.species), rownames(metadata)))

ps <- phyloseq(otu_table(mphlan.species, taxa_are_rows = TRUE),
               sample_data(metadata),
               tax_table(species)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps.filt <- filter_taxa(ps, function(x) mean(x) > 0.01, TRUE)
ps.filt  <- transform_sample_counts(ps.filt, function(x) (x / sum(x))*100)

### Filter to DOL 0-60 and remove After only infants ########
ps_filt_dol60 <- ps.filt %>% subset_samples(DOL <= 60) %>% subset_samples(abx_group_updated != "After only")
ps_filt_dol60_meta <- data.frame(sample_data(ps_filt_dol60))

sample_data(ps_filt_dol60)$DOL_bin <- factor(sample_data(ps_filt_dol60)$DOL_bin)
sample_data(ps_filt_dol60)$abx_group_updated <- factor(sample_data(ps_filt_dol60)$abx_group_updated,
                                               levels = c("Early and after","Early only","None"))

###################### Merge with P4 cross-sectional cohort ###########
p4_mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/251020_P4_merged_metaphlan_n146.txt", header = T, sep="\t", check.names = FALSE))

## Processing data for phyloseq
rownames(p4_mphlan) <- p4_mphlan$clade_name
p4_mphlan$clade_name <- NULL

new_names <- colnames(p4_mphlan)
new_names <- gsub("^X", "", new_names)
new_names <- gsub("_", ".", new_names)
new_names <- paste0("p4_", new_names)

colnames(p4_mphlan) <- new_names

## Create species data for phyloseq 
# keep rows with species info 
p4_mphlan.species <- p4_mphlan[grepl('s__', rownames(p4_mphlan)), ]
# Drop rows with SBG (species level genome bins)
p4_mphlan.species <- p4_mphlan.species[!grepl('t__', rownames(p4_mphlan.species)), ]
p4_mphlan.species <- p4_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
p4_mphlan.species <- as.matrix(p4_mphlan.species)

## Create taxa table
p4_species <- data.frame(Names = rownames(p4_mphlan.species))
p4_species <- data.frame(do.call('rbind', strsplit(as.character(p4_species$Names), '|', fixed = TRUE)))
rownames(p4_species) <- rownames(p4_mphlan.species)
colnames(p4_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
p4_species <- as.matrix(p4_species)

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(p4_mphlan.species), rownames(metadata)))

ps_p4 <- phyloseq(otu_table(p4_mphlan.species, taxa_are_rows = TRUE),
               sample_data(metadata),
               tax_table(p4_species)) 

#ps.filt <- prune_taxa(taxa_sums(ps)>0.01, ps)
ps_p4.filt <- filter_taxa(ps_p4, function(x) mean(x) > 0.01, TRUE)
ps_p4.filt  <- transform_sample_counts(ps_p4.filt, function(x) (x / sum(x))*100)

ps_p4_meta <- data.frame(phyloseq::sample_data(ps_p4.filt))

ps_p4.filt <- ps_p4.filt %>% subset_samples(abx_group_updated != "After only")
ps_p4.filt_meta <- data.frame(sample_data(ps_p4.filt))

sample_data(ps_p4.filt)$DOL_bin <- factor(sample_data(ps_p4.filt)$DOL_bin)
sample_data(ps_p4.filt)$abx_group_updated <- factor(sample_data(ps_p4.filt)$abx_group_updated,
                                               levels = c("Early and after","Early only","None"))

merged_ps <- merge_phyloseq(ps_filt_dol60, ps_p4.filt)
merged_ps_meta <- data.frame(sample_data(merged_ps))

merged_ps_meta <- merged_ps_meta %>% mutate(abx_group_names = case_when(abx_group_updated == "Early only" ~ "EEO",
                                                                          abx_group_updated == "Early and after" ~ "EAS",
                                                                          abx_group_updated == "None" ~ "ABN"))

merged_ps_meta$abx_group_names <- factor(merged_ps_meta$abx_group_names, levels = c("EAS", "EEO", "ABN"))
sample_data(merged_ps) <- merged_ps_meta

#save(merged_ps, file = "260225_merged_ps.RData")


```

### Load near-term data
```{r}
drewg_meta <- read_excel("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/240515_near-term_sample_metadata_v1.xlsx")

drewg_meta <- drewg_meta %>% filter(Cohort == "Gasparrini_term") 
drewg_meta <- drewg_meta[,1:7]

extra_meta <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Metadata/240515_near-term_sample_metadata2.xlsx") %>%
  filter(Group == "Term") %>% select(Sample, GestationalAgeWeeks, Birthweight,
                                     Gender, DeliveryMode, TotalAntimicrobialsDays)

extra_meta <- extra_meta %>% dplyr::rename(Sample_ID = Sample, 
                                    Gestational_age_weeks = GestationalAgeWeeks,
                                    Sex = Gender,
                                    Delivery_route = DeliveryMode,
                                    abx_exposure_count = TotalAntimicrobialsDays)

drewg_meta_final <- merge(drewg_meta, extra_meta, by = "Sample_ID")

drewg_meta_dol60 <- drewg_meta_final %>% filter(DOL <= 60)

drewg_mphln <- mphlan <- data.frame(read.table("~/Documents/GitHub/PretermPlasmidPhageProject/R/metagenomics/251016_near-term_n162_metaphlan3_merged.txt", header = T, sep="\t", skip=1, check.names = FALSE))
rownames(drewg_mphln) <- drewg_mphln$clade_name
drewg_mphln$clade_name <- NULL

drewg_mphlan.species <- drewg_mphln[grepl('s__', rownames(drewg_mphln)), ]
# Drop rows with SBG (species level genome bins)
drewg_mphlan.species <- drewg_mphlan.species[!grepl('t__', rownames(drewg_mphlan.species)), ]
drewg_mphlan.species <- drewg_mphlan.species %>% dplyr::select_if(~sum(.) != 0)
drewg_mphlan.species <- as.matrix(drewg_mphlan.species)

## Create taxa table
drewg_species <- data.frame(Names = rownames(drewg_mphlan.species))
drewg_species <- data.frame(do.call('rbind', strsplit(as.character(drewg_species$Names), '|', fixed = TRUE)))
rownames(drewg_species) <- rownames(drewg_mphlan.species)
colnames(drewg_species) <- c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species')
drewg_species <- as.matrix(drewg_species)

## Metadata
rows_metadata <- rownames(drewg_meta_final)[colnames(drewg_mphlan.species) %in% drewg_meta_final$Sample_ID]
drewg_meta_final <- data.frame(drewg_meta_final)
rownames(drewg_meta_final) <- drewg_meta_final$Sample_ID

drewg_cat_var <- c("Sample_ID", "Patient", "Cohort", "Sex", "Delivery_route")
drewg_num_var <- c("DOL", "DOL_months", "PostmenstrualAgeDays","PostmenstrualAgeWeeks","Gestational_age_weeks","Birthweight",
                   "abx_exposure_count")
# Set categorical variables to factor
drewg_meta_final[,drewg_cat_var] <- lapply(drewg_meta_final[,drewg_cat_var], factor)
# Set numerical variables to numeric
drewg_meta_final[,drewg_num_var] <- lapply(drewg_meta_final[,drewg_num_var], as.numeric)
print(dim(drewg_meta_final))

# Differences between metadata rows and metaphlan column names 
## These are samples that we have dropped because we have other samples from the same DOL  
print(setdiff(colnames(drewg_mphlan.species), rownames(drewg_meta_final)))

# 160 samples and 341 species
ps_term <- phyloseq(otu_table(drewg_mphlan.species, taxa_are_rows = TRUE),
               sample_data(drewg_meta_final),
               tax_table(drewg_species)) 

# 160 samples and 179 species
ps_term <- filter_taxa(ps_term, function(x) mean(x) > 0.01, TRUE)
ps_term  <- transform_sample_counts(ps_term, function(x) (x / sum(x))*100)

### Filter to DOL 0-60 and remove After only infants ########
ps_term_dol100 <- ps_term %>% subset_samples(DOL <= 100)

ps_term_meta <- data.frame(phyloseq::sample_data(ps_term_dol100))

# Filter out samples past DOL 100 and create DOL bin
ps_term_meta <- ps_term_meta %>%
  mutate(DOL_bin = cut(DOL,
                       breaks = seq(0,100, by = 10),
                       labels = paste("DOL", seq(1, 91, by = 10), "-", seq(10, 100, by = 10)),
                       include.lowest = TRUE,
                       right = TRUE))


sample_data(ps_term_dol100) <- ps_term_meta

ps_term_dol60 <- ps_term_dol100 %>% subset_samples(DOL <= 60) 
ps_term_dol60_meta <- data.frame(sample_data(ps_term_dol60))

sample_data(ps_term_dol60)$DOL_bin <- factor(sample_data(ps_term_dol60)$DOL_bin)
sample_data(ps_term_dol60)$abx_group_updated <- "Near-term"

ps_term_dol60_meta <- data.frame(sample_data(ps_term_dol60))

breaks <- seq(23, 48, by = 3)
labels <- paste("PMA", head(breaks, -1), "-",tail(breaks, -1))

ps_term_dol60_meta$PMA_bin <- cut(
  ps_term_dol60_meta$PostmenstrualAgeWeeks,
  breaks = breaks,
  labels = labels,
  right = FALSE,  # left-inclusive, i.e., [23,26)
  include.lowest = TRUE
)


sample_data(ps_term_dol60) <- ps_term_dol60_meta

## Merge preterm and term phyloseq objects #########
ps_term_dol60_meta <- ps_term_dol60_meta %>% dplyr::rename(Sample_ID_R_final = Sample_ID) 
pterm_meta_simple <- merged_ps_meta %>% select(any_of(colnames(ps_term_dol60_meta)))
nterm_meta_simple <- ps_term_dol60_meta %>% select(everything(), -c(DOL_months, PostmenstrualAgeDays)) 
nterm_meta_simple$Sample_ID_R_final <- as.character(nterm_meta_simple$Sample_ID_R_final)
nterm_meta_simple$Delivery_route <- as.character(nterm_meta_simple$Delivery_route)
nterm_meta_simple$Delivery_route[nterm_meta_simple$Delivery_route == "Cesarean"] <- "Cesarean Section"
nterm_meta_simple$Delivery_route <- factor(nterm_meta_simple$Delivery_route)


ps_pterm_simple <- merged_ps
sample_data(ps_pterm_simple) <- pterm_meta_simple

ps_nterm_simple <- ps_term_dol60
sample_data(ps_nterm_simple) <- nterm_meta_simple

ps_infants <- merge_phyloseq(ps_pterm_simple, ps_nterm_simple)
ps_infants_meta <- data.frame(sample_data(ps_infants))
ps_infants_meta$abx_group_updated <- factor(ps_infants_meta$abx_group_updated)
ps_infants_meta <- ps_infants_meta %>% mutate(abx_group_names = case_when(abx_group_updated == "Early only" ~ "EEO",
                                                                          abx_group_updated == "Early and after" ~ "EAS",
                                                                          abx_group_updated == "None" ~ "ABN",
                                                                          abx_group_updated == "Near-term" ~ "Near-term"))
ps_infants_meta$abx_group_names <- factor(ps_infants_meta$abx_group_names, levels = c("EAS", "EEO", "ABN", "Near-term"))
#save(ps_infants, file = "251118_Preterm_near-term_merged_object.RData")

###### Add site to metadata #########
preterm_site <- merged_ps_meta %>% select(Sample_ID_R_final, Site)
ps_infants_meta2 <- ps_infants_meta %>% left_join(preterm_site, by = "Sample_ID_R_final") %>% mutate(Site = if_else(is.na(Site), "Home", Site))
ps_infants_meta2$Site <- factor(ps_infants_meta2$Site)

### Update cohort to study ###
ps_infants_meta2 <- ps_infants_meta2 %>% mutate(Study = case_when(Cohort == "BSI" ~ "Schwartz_et_al",
                                                                  Cohort == "Gasparrini_preterm" ~ "Gasparrini_et_al",
                                                                  Cohort == "Gasparrini_term" ~ "Gasparrini_et_al",
                                                                  Cohort == "Molly" ~ "Gibson_et_al",
                                                                  Cohort == "NEC_control" ~ "Thanert_et_al",
                                                                  Cohort == "No_abx_preterm" ~ "This_study",
                                                                  Cohort == "P4_cross_sectional" ~ "This_study"))
ps_infants_meta2$Study <- factor(ps_infants_meta2$Study)
rownames(ps_infants_meta2) <- ps_infants_meta2$Sample_ID_R_final
sample_data(ps_infants) <- ps_infants_meta2

ps_infants_meta <- data.frame(sample_data(ps_infants))

#save(ps_infants, file = "260225_ps_infants.RData")

#### Create metadata file for Supplementary Table 1 ######
ps_infants_meta_all <- merge(ps_infants_meta, merged_ps_meta, by = "Sample_ID_R_final", all = TRUE)
write.csv(ps_infants_meta_all, "260225_GM_Dev_Metadata_Supplementary_Table1_Final.csv", row.names = F)

```

Table 1
```{r}
library(tibble) # v3.3.1 for rownames_to_column() function
library(stringr) # v1.6.0 for str_squish()
library(flextable) # v0.9.11
library(officer) # v0.7.3
library(tableone) # v0.13.2

source("~/Documents/GitHub/Preterm_Plasmidome/GM_development/customtab.R")

meta_pt <- ps_infants_meta

eeo <- ps_infants_meta %>% filter(abx_group_names == "EEO")
abn <- ps_infants_meta %>% filter(abx_group_names == "ABN")
eas <- ps_infants_meta %>% filter(abx_group_names == "EAS")

meta_pt$Delivery_route[meta_pt$Delivery_route == "Cesarean"] <- "Cesarean Section"
meta_pt$Cohort[meta_pt$Cohort == "BSI_new"] <- "BSI"

# Set categorical variables as factors !!!!!!!!!!! Modify these as needed later !!!!!!!!!!!!!!!!!!!!
categorical_vars <- c('Patient','abx_group_names', 'Sex', 'Delivery_route', 'Cohort')

meta_pt[,categorical_vars] <- lapply(meta_pt[,categorical_vars], factor)
meta_pt$Delivery_route <- factor(meta_pt$Delivery_route, levels = c("Cesarean Section", "Vaginal"))
meta_pt$Cohort <- factor(meta_pt$Cohort, levels = c("Molly", "Gasparrini_preterm", "Gasparrini_term", "BSI", "NEC_control", "No_abx_preterm", "P4_cross_sectional"))

numerical_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks")
meta_pt[,numerical_vars] <- lapply(meta_pt[,numerical_vars], as.numeric)

meta_pt[,numerical_vars] <- lapply(meta_pt[,numerical_vars], as.numeric)

## All numerical variables are non-normal so we'll use median (IQR) instead of mean (SD)
normality_results <- lapply(numerical_vars, function(v) {
  p <- shapiro.test(meta_pt[[v]])$p.value
  data.frame(variable = v, p_value = p)
}) %>% bind_rows()

# Generate metadata comparison table 
comp_vars <- c("Birthweight", "abx_exposure_count", "Gestational_age_weeks",'Cohort', 'Sex', 'Delivery_route')


tab_pt <- CreateTableOne(vars=comp_vars,
                           strata='abx_group_names', 
                           data=meta_pt)
tab_plot_pt <- print(tab_pt, 
  showAllLevels = TRUE,
  quote        = FALSE,
  noSpaces     = TRUE,
  test         = TRUE,
  contDigits   = 1,
  nonnormal    = numerical_vars,  
  printToggle  = FALSE,
  dropEqual    = TRUE,
  explain      = FALSE)

# Convert to dataframe
tab_pt_df <- as.data.frame(tab_plot_pt) %>% rownames_to_column(var = "Variable")

# Set Table header
header <- str_squish(str_remove("Table 1. Infant demographics", "\n"))
# Set Table footer
footer <- str_squish(str_remove("Numbers are either No. or mean values", "\n"))
# Set custom_tab() defaults
customtab_defaults()
# Create the flextable object
flextable_processed_samples <- custom_tab(tab_pt_df, header, footer)

# Save as word .docx
save_as_docx(flextable_processed_samples, path = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/Metadata/Table1_near-term.docx",
pr_section = prop_section(page_size = page_size(orient = "landscape"), type = "continuous"))


# 2. Filter infants who have more than one group label
changed_infants <- group_changes %>%
  filter(groups > 1)

# 3. View how many and which infants changed groups
n_changed <- nrow(changed_infants)
print(paste("Number of infants who changed group:", n_changed))
print(changed_infants)

```


## Alpha diversity (Related to Fig. 1A, Supplementary Figs 1B-D)
```{r}
library(rlang) # v1.1.7
library(dplyr) # v1.2.0
library(lme4) # v1.1.38
library(emmeans) # v2.0.1
library(rstatix) # v0.7.3
library(ggplot2) # v4.0.2
library(ggpubr) # v0.6.3

ps.filt.trans <- transform_sample_counts(ps_infants, function(x) trunc(x*100000))
adiv <- estimate_richness(ps.filt.trans, measures=c('Observed', 'Shannon'))

sample_data(ps_infants)$Richness <- adiv$Observed
sample_data(ps_infants)$Shannon <- adiv$Shannon

#### Merge alpha diversity with metadata file for supplementary table 1 ################################
current_meta <- read_excel("~/Documents/GitHub/Preterm_Plasmidome/GM_development/260225_GM_Dev_Metadata_Supplementary_Table1_Final.xlsx")
alpha_div_meta <- data.frame(phyloseq::sample_data(ps_infants)) %>% select(Sample_ID_R_final, Richness, Shannon)
alpha_merged_meta <- merge(current_meta, alpha_div_meta, by = "Sample_ID_R_final", all = TRUE)
########################################## ------------------------- #################################

counts_by_bin <- ps.filt.meta %>%
  filter(!is.na(Richness), !is.na(abx_group_names)) %>%
  mutate(abx_group_names = droplevels(abx_group_names)) %>%
  count(DOL_bin, abx_group_names, name = "n") %>%
  group_by(DOL_bin) %>% mutate(groups_present = n_distinct(abx_group_names)) %>% ungroup()
write.csv(counts_by_bin, "260202_abx_exposure_n_by_group_names.csv", row.names = F)

## There are repeat samples and differences in DOL so we need to use linear models to account for these confounders
patient_repeats_per_DOL_bin <- ps.filt.meta %>%
  group_by(DOL_bin, Patient) %>% summarise(n_samples = n(), .groups = "drop")

########### Richness ##################
### Perform stat test and plot #######
lmer_emmeans_richness <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness, "260202_lmer_emmeans_richness.csv", row.names = F)
## Plot alpha diversity barplot binned by DOL 
richness_abx_dol <- ggplot(ps.filt.meta, aes(x = abx_group_names, y = Richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Species Richness", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )


richness_abx_dol_emm_lmer <- richness_abx_dol + stat_pvalue_manual(
  lmer_emmeans_richness,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Richness_abx_group_DOL_bp_emmeans.pdf", plot = richness_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)


########### Shannon ##################
### Perform stat test and plot #######
lmer_emmeans_shannon <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon, "260202_lmer_emmeans_shannon.csv", row.names = F)

shannon_abx_dol <- ggplot(ps.filt.meta, aes(x = abx_group_names, y = Shannon, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Shannon Diversity", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )


shannon_abx_dol_emm_lmer <- shannon_abx_dol + stat_pvalue_manual(
  lmer_emmeans_shannon,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Shannon_abx_group_DOL_bp_emmeans.pdf", plot = shannon_abx_dol_emm_lmer, bg = "transparent", width = 20, height = 10)

###### Facet by abx group #######
lmer_emmeans_shannon_DOL <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon_DOL, "260202_lmer_emmeans_shannon_by_DOL.csv", row.names = F)

shannon_abx_dol2 <- ggplot(ps.filt.meta, aes(x = DOL_bin, y = Shannon, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ abx_group_names, nrow = 2, strip.position = "top") +
  labs(
    x = "DOL bin", 
    y = "Shannon Diversity", 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260202_P4_GM_Dev_Shannon_abx_group_facet_DOL_bp.pdf", plot = shannon_abx_dol2, bg = "transparent", width = 20, height = 10)

lmer_emmeans_shannon_DOL <- lmer_emmeans_shannon_DOL %>% separate(contrast, into = c("group1", "group2"), sep = "\\) - \\(") %>%
  mutate(
    group1 = gsub("^\\(|\\)$", "", group1),
    group2 = gsub("^\\(|\\)$", "", group2)
  )

shannon_abx_dol2_emm_lmer <- shannon_abx_dol2 + stat_pvalue_manual(
  lmer_emmeans_shannon_DOL,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.4, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Shannon_abx_group_facet_DOL_bp_emmeans.pdf", plot = shannon_abx_dol2_emm_lmer, bg = "transparent", width = 20, height = 10)


lmer_emmeans_richness_DOL <- perform_emmeans_lmer(df = ps.filt.meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness_DOL, "260202_lmer_emmeans_richness_by_DOL.csv", row.names = F)

richness_abx_dol2 <- ggplot(ps.filt.meta, aes(x = DOL_bin, y = Richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(width=0.2)+
  scale_fill_manual(values=custom_colors_names)+
  facet_wrap(~ abx_group_names, nrow = 2, strip.position = "top") +
  labs(
    x = "DOL bin", 
    y = "Species Richness", 
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = "none",
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260202_P4_GM_Dev_Richness_abx_group_facet_DOL_bp.pdf", plot = richness_abx_dol2, bg = "transparent", width = 20, height = 10)

lmer_emmeans_richness_DOL <- lmer_emmeans_richness_DOL %>% separate(contrast, into = c("group1", "group2"), sep = "\\) - \\(") %>%
  mutate(
    group1 = gsub("^\\(|\\)$", "", group1),
    group2 = gsub("^\\(|\\)$", "", group2)
  )

richness_abx_dol2_emm_lmer <- richness_abx_dol2 + stat_pvalue_manual(
  lmer_emmeans_richness_DOL,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.4, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_Richness_abx_group_facet_DOL_bp_emmeans.pdf", plot = richness_abx_dol2_emm_lmer, bg = "transparent", width = 20, height = 10)

###--- Run lmer and emmeans separately on term dataset ######
nterm_meta <- ps.filt.meta %>% filter(abx_group_names == "Near-term")
lmer_emmeans_richness_DOL_nterm <- perform_emmeans_lmer(df = nterm_meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_richness_DOL_nterm, "260203_lmer_emmeans_richness_by_DOL_nterm.csv", row.names = F)

lmer_emmeans_shannon_DOL_nterm <- perform_emmeans_lmer(df = nterm_meta, facet_var = "abx_group_names", x_var = "DOL_bin", y_var = "Shannon",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_shannon_DOL_nterm, "260203_lmer_emmeans_shannon_by_DOL_nterm.csv", row.names = F)

```

ARG Comparison (Related to Fig 1B, Supplementary Fig. 1E, G)
```{r}
library(rlang) # v1.1.7
library(dplyr) # v1.2.0
library(lme4) # v1.1.38
library(emmeans) # v2.0.1
library(rstatix) # v0.7.3
library(ggplot2) # v4.0.2
library(ggpubr) # v0.6.3

#look at overall ARG burden between groups
##load shortbred output
arg.full <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/250905_P4_GM_Dev_ShortBRED_merged_output_n2585.txt"))
arg.full <- arg.full %>% mutate(Sample_ID_R_final = gsub("-", ".", paste0("p4_", Sample)))
arg.p4 <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251022_GM_Dev_P4_shortBRED_joint_output.txt"))
arg.p4 <- arg.p4 %>% mutate(Sample_ID_R_final = paste0("p4_", gsub("_", ".", Sample)))
arg_df <- rbind(arg.full, arg.p4)

# starts_with_number <- grepl("^[0-9]", arg.full$Sample)
# modified_column <- arg.full$Sample
# modified_column[starts_with_number] <- paste0("p4_", gsub("-", ".", arg.full$Sample[starts_with_number]))
# arg.full$Sample_ID_R_final <- modified_column

arg.wide <- as.data.frame(dcast(arg_df, Sample_ID_R_final ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg.wide) <- arg.wide$Sample_ID_R_final
arg.wide <- arg.wide[, c(-1)]

##sum counts to get RPKM per sample
arg.wide$RPKM <- rowSums(arg.wide)

##get ARG richness
arg.wide$Richness <- specnumber(arg.wide[, 1:ncol(arg.wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg.wide$Sample_ID_R_final <- row.names(arg.wide)
# Subset arg.wide to columns we want
arg.want <- arg.wide %>%
  select(c(Sample_ID_R_final, RPKM, Richness)) %>% dplyr::rename(ShortBRED_richness = Richness)

# Merge with metadata
arg.df <- merge(ps_infants_meta, arg.want, by = "Sample_ID_R_final")

arg.df <- arg.df %>% filter(abx_group_updated != "After only") 

##### Read in term ARG data ######
arg_term <- as.data.frame(read.delim("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/251016_DrewG_term_n162_shortbred_merged.txt"))

arg_term_wide <- as.data.frame(dcast(arg_term, Sample ~ Family, value.var = 'Count', fun.aggregate = sum))
row.names(arg_term_wide) <- arg_term_wide$Sample
arg_term_wide <- arg_term_wide[, c(-1)]

##sum counts to get RPKM per sample
arg_term_wide$RPKM <- rowSums(arg_term_wide)

##get ARG richness
arg_term_wide$Richness <- specnumber(arg_term_wide[, 1:ncol(arg_term_wide)-1])

# Merge arg.wide to metadata 
## First add a paperID column to merge with metadata file
arg_term_wide$Sample <- row.names(arg_term_wide)
# Subset arg.wide to columns we want
arg_term_want <- arg_term_wide %>%
  select(c(Sample, RPKM, Richness)) %>% dplyr::rename(Sample_ID_R_final = Sample, ShortBRED_richness = Richness)

# Merge with metadata
term_arg.df <- merge(ps_infants_meta, arg_term_want, by = "Sample_ID_R_final") 

#### Merge with preterm ARG data ###
arg.df2 <- arg.df %>% 
  select(Sample_ID_R_final, Patient, Study, Site, DOL_bin, Gestational_age_weeks, abx_group_updated, abx_group_names, RPKM, ShortBRED_richness)
term_arg.df2 <- term_arg.df %>% select(Sample_ID_R_final, Patient, Study, Site, DOL_bin, Gestational_age_weeks, abx_group_updated, abx_group_names, RPKM, ShortBRED_richness)

arg_full <- rbind(arg.df2, term_arg.df2)
write.csv(arg_full, "260130_P4_GM_Dev_ARG_data.csv", row.names = F)

##### Merge ARG data with metadata for Supplementary table 1 ##################################
arg_meta <- arg_full %>% select(Sample_ID_R_final, RPKM, ShortBRED_richness) %>% dplyr::rename(ARG_abundance_RPKM = RPKM,
                                                                                               ARG_richness = ShortBRED_richness)
arg_merged_meta <- merge(alpha_merged_meta, arg_meta, by = "Sample_ID_R_final", all = TRUE)
#write.csv(arg_merged_meta, "260225_GM_Dev_alpha_arg_merged_meta_final.csv", row.names = F)
######################### ---------------------------------- ###################################

###### Plot and compare using linear mixed models ###########
lmer_emmeans_arg_rpkm <- perform_emmeans_lmer(df = arg_full, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "RPKM",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_arg_rpkm, "260202_lmer_emmeans_ARG_rpkm.csv", row.names = F)

counts_by_bin <- arg_full %>%
  filter(!is.na(RPKM), !is.na(abx_group_names)) %>%
  mutate(abx_group_names = droplevels(abx_group_names)) %>%
  count(DOL_bin, abx_group_names, name = "n") %>%
  group_by(DOL_bin) %>% mutate(groups_present = n_distinct(abx_group_names)) %>% ungroup()
write.csv(counts_by_bin, "260202_abx_exposure_n_by_group_shortbred_comparison.csv", row.names = F)

shortbred_hits_rpkm <- ggplot(arg_full, aes(x = abx_group_names, y = RPKM, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Total ShortBRED hits (RPKM)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

shortbred_hits_rpkm_stat  <- shortbred_hits_rpkm + stat_pvalue_manual(
  lmer_emmeans_arg_rpkm,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_ARG_RPKM_abx_group_DOL_bp_emmeans.pdf", plot = shortbred_hits_rpkm_stat, bg = "transparent", width = 20, height = 10)

########### Total ShortBRED Hits (Richness) #########
lmer_emmeans_arg_richness <- perform_emmeans_lmer(df = arg_full, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "ShortBRED_richness",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_arg_richness, "260202_lmer_emmeans_ARG_richness.csv", row.names = F)

shortbred_hits_richness <- ggplot(arg_full, aes(x = abx_group_names, y = ShortBRED_richness, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Total ShortBRED hits (Unique ARGs)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

shortbred_hits_richness_stat <- shortbred_hits_richness + stat_pvalue_manual(
  lmer_emmeans_arg_richness,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)
ggsave("260202_P4_GM_Dev_ARG_Richness_abx_group_DOL_bp_emmeans.pdf", plot = shortbred_hits_richness_stat, bg = "transparent", width = 20, height = 10)

###################-------------Richness normalized ARGs------------#######################
### Merge ARG data with phyloseq metadata ###
arg_full_want <- arg_full %>% select(Sample_ID_R_final, RPKM, ShortBRED_richness) %>% 
  dplyr::rename(ShortBRED_RPKM = RPKM)
ps.filt.meta2 <- merge(ps.filt.meta, arg_full_want, by = "Sample_ID_R_final")

ps.filt.meta2$norm_arg_abund <- ps.filt.meta2$ShortBRED_RPKM / ps.filt.meta2$Richness
#write.csv(ps.filt.meta2, "260202_P4_GM_Dev_all_infants_meta_alpha_ARGs.csv", row.names = F)

lmer_emmeans_norm_arg_richness <- perform_emmeans_lmer(df = ps.filt.meta2, facet_var = "DOL_bin", x_var = "abx_group_names", y_var = "norm_arg_abund",
                                              fixed_effects = c("Gestational_age_weeks"),
                                              random_effects = c("(1|Patient)", "(1|Site)", "(1|Study)"),
                                              p.adjust.method = "BH",
                                              df_method = "kenward-roger")
write.csv(lmer_emmeans_norm_arg_richness, "260206_lmer_emmeans_richness_norm_ARG_rpkm.csv", row.names = F)

norm_arg_abund <- ggplot(ps.filt.meta2, aes(x = abx_group_names, y = norm_arg_abund, fill = abx_group_names)) +
  geom_boxplot(outlier.shape=NA)+
  geom_jitter(shape = 17, aes(fill = abx_group_names), color = "black", width=0.2, size = 2, alpha = 0.4)+
  scale_fill_manual(values=custom_colors_names)+
  scale_color_manual(values=custom_colors_names) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  labs(
    x = "Antibiotic Exposure Group", 
    y = "Richness Normalized Total ShortBRED Hits (RPKM)", 
    fill = "Antibiotic Exposure Group",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 18, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
ggsave("260206_P4_GM_Dev_Richness_Norm_ARG_rpkm_abx_group_.pdf", plot = norm_arg_abund, bg = "transparent", width = 20, height = 10)

norm_arg_abund_stat <- norm_arg_abund + stat_pvalue_manual(
  lmer_emmeans_norm_arg_richness,
  label = "p.adj.signif",   # can use "p.adj" for exact values
  tip.length = 0.01,
  step.increase = 0.05,
  hide.ns = TRUE, bracket.size = 0.6, facet.y = "bin"
)

```

Repeat measures PERMANOVA (Related to Fig. 1C, D and Supplementary Fig. 1F)
This section was run through a high-throughput computing facility 
```{r}
library(phyloseq) # v1.50.0
library(vegan) # v.2.7.2
library(doParallel) # v1.0.17
library(foreach) # v1.5.2
library(permute) # v0.9.10
library(stringr) # v1.6.0
library(dplyr) # v1.2.0

source("GM_development_functions.R")
source("Repeat_measures_PERMANOVA_function.R")
load("260202_merged_ps.RData")
dim(merged_ps)

cat_var <- c("Sample_ID", "Patient", "Study", "abx_group_updated",	"abx_group_names","Necrotizing_enterocolitis","Sex","Delivery_route","Maternal_ethnicity","Maternal_race","Site","Final_disposition",
             "Para", "Maternal_smoking",	"Maternal_alcohol",	"Maternal_substance_abuse",	"Maternal_preeclampsia",	"Maternal_diabetes",	
             "Maternal_chorioamnionitis","Maternal_hypertension", "mat_inpt_antimicrobe","mat_outpt_antimicrobial","Multiple_births")

num_var <- c("DOL","MOL","Gestational_age_weeks", "Birthweight","Maternal_age_at_delivery","Gravida", "abx_exposure_count","acute_abx_exposure_count","antiviral_exposure", "aminoglycoside_exposure","penicillin_exposure","macrolide_exposure","firstgencephalosporin_exposure",
             "fourthgencephalosporin_exposure","thirdgencephalosporin_exposure","secondgencephalosporin_exposure","lincosamide_exposure",
             "antifungal_exposure","carbapenem_exposure","other_antimicrobials_exposure","other_non_antimicrobials_exposure","glycopeptide_exposure","med_exposure_count", "morphine_exposure", "fentanyl_exposure", "surfactant_exposure", "acetaminophen_exposure", "caffeine_exposure", "cholecalciferol_vit_d__exposure",	'vitamin_a_exposure',	'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	'bumetanide_bumax__exposure',	'other_exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	'sodium_bicarbonate_exposure','hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure',"acute_antiviral_exposure", 
"acute_aminoglycoside_exposure","acute_penicillin_exposure","acute_macrolide_exposure","acute_firstgencephalosporin_exposure",
             "acute_fourthgencephalosporin_exposure","acute_thirdgencephalosporin_exposure","acute_secondgencephalosporin_exposure","acute_lincosamide_exposure",
             "acute_antifungal_exposure","acute_carbapenem_exposure","acute_other_antimicrobials_exposure","acute_other_non_antimicrobials_exposure",
             "acute_glycopeptide_exposure","acute_med_exposure_count", "acute_morphine_exposure", "acute_fentanyl_exposure", "acute_surfactant_exposure", 
             "acute_acetaminophen_exposure", "acute_caffeine_exposure", "acute_cholecalciferol_vit_d__exposure",	'acute_vitamin_a_exposure',	
             'acute_famotidine_pepcid__exposure',	'acute_iron_exposure',	'acute_calcium_gluconate_exposure',	'acute_dopamine_exposure','acute_adenosine_exposure',	
             'acute_cosyntropin_cortrosyn_acth__exposure','acute_bumetanide_bumax__exposure',	'acute_other_exposure',	'acute_indomethacin_exposure',
             'acute_multi_vitamin_with_iron_exposure',	'acute_lasix_exposure',	'acute_atropine_exposure',	'acute_dobutamine_exposure',
             'acute_sodium_chloride_exposure',	'acute_levothyroxine_exposure',	'acute_dexamethasone_exposure',	'acute_phenobarbitol_exposure',	
             'acute_hydrocortisone_exposure',	'acute_ibuprofen_exposure',	'acute_chlorathiazide_diuril__exposure',	'acute_midazolam_versed__exposure',	
             'acute_potassium_chloride_exposure',	'acute_hepatitis_b_immune_globulin_exposure',	'acute_citric_acid_sodium_citrate_bicitra__exposure',	
             'acute_pancuronium_or_vecurionium_exposure',	'acute_insulin_exposure',	'acute_immunoglobulin_other_exposure',	'acute_multi_vitamin_no_iron_exposure', 
             'acute_levocarnitine_exposure','acute_digoxin_exposure','acute_hydraluronidase_hydase__exposure','acute_lorazapam_ativan__exposure',
             'acute_sodium_bicarbonate_exposure','acute_hydrochlorathiazide_hydrodiuril__exposure',
             'acute_spironolactone_hydrochlorathiazide_aldactazide__exposure','acute_tocopherol_acetate_vit_e__exposure','cum_days_formula',
             'cum_days_milk','frac_days_formula', 'frac_days_milk')

### Run Repeat measure PERMANOVA #####
ps_dol60_meta <- data.frame(phyloseq::sample_data(merged_ps))

### Determine metadata of interest 
meta_of_interest <- c("Patient", 'Gestational_age_weeks', 'Birthweight', 'Sex', 'Delivery_route', 'Necrotizing_enterocolitis', 
                      'Site', 'Final_disposition', 'Maternal_race', 'Maternal_age_at_delivery', 'Gravida', 'Para', 'Maternal_smoking', 
                      'Maternal_alcohol', 'Maternal_substance_abuse', 'Maternal_preeclampsia', 'Maternal_diabetes', 
                      'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe', 'mat_outpt_antimicrobial', 
                      'Multiple_births', num_var)


# Keep only rows without any NA
ps_dol60_meta_filt <- ps_dol60_meta %>% filter(if_all(all_of(meta_of_interest), ~ !is.na(.x)))
summary(ps_dol60_meta_filt$abx_group_updated)

# Rows with at least one NA --> 31 samples filtered out 
ps_dol60_meta_NAs <- ps_dol60_meta %>% filter(if_any(all_of(meta_of_interest), ~ is.na(.x)))

ps_filt2_dol60 <- phyloseq::subset_samples(merged_ps, phyloseq::sample_names(merged_ps) %in% ps_dol60_meta_filt$Sample_ID_R_final)

## Calculate bray curtis distance 
gm_dev_dol60.dist <- phyloseq::distance(ps_filt2_dol60, method="bray")

# Create subject_data which is a dataframe with per-subject metadata or data specific to each subject that don't change over time
gm_dev_subject_data <- ps_dol60_meta_filt %>%
  select(Patient, Gestational_age_weeks, Birthweight, Sex, Delivery_route, Necrotizing_enterocolitis, Site, Final_disposition, Maternal_race, 
  Maternal_age_at_delivery, Gravida, Para, Maternal_smoking, Maternal_alcohol, Maternal_substance_abuse, Maternal_preeclampsia, Maternal_diabetes, 
  Maternal_chorioamnionitis, Maternal_hypertension, mat_inpt_antimicrobe, mat_outpt_antimicrobial, Multiple_births)

gm_dev_subject_data <- gm_dev_subject_data %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(num_var, names(.))), as.numeric))

inherits(gm_dev_dol60.dist, "dist")
gm_dev_subject_data2 <- unique(gm_dev_subject_data)

rownames(gm_dev_subject_data2) <- gm_dev_subject_data2$Patient
gm_dev_subject_data2$Patient <- NULL

# Create sample_data file which is a dataframe with per-sample metadata
## do the rownames here also have to match?
gm_dev_sample_data <- ps_dol60_meta_filt %>%
  select(DOL, Study, abx_group_updated, cum_days_formula,cum_days_milk,frac_days_formula,frac_days_milk,abx_exposure_count,antiviral_exposure, aminoglycoside_exposure,	
         penicillin_exposure, macrolide_exposure,	firstgencephalosporin_exposure,	fourthgencephalosporin_exposure,	
         thirdgencephalosporin_exposure,	secondgencephalosporin_exposure,	lincosamide_exposure,	antifungal_exposure,	
         carbapenem_exposure,	other_antimicrobials_exposure,	other_non_antimicrobials_exposure, glycopeptide_exposure, med_exposure_count,
         morphine_exposure,	fentanyl_exposure,	surfactant_exposure,	acetaminophen_exposure,	caffeine_exposure,
         cholecalciferol_vit_d__exposure,	vitamin_a_exposure,	famotidine_pepcid__exposure,	iron_exposure,	
         calcium_gluconate_exposure,	dopamine_exposure,	adenosine_exposure,	cosyntropin_cortrosyn_acth__exposure,	
         bumetanide_bumax__exposure,	other_exposure,	indomethacin_exposure,	multi_vitamin_with_iron_exposure,	lasix_exposure,	
         atropine_exposure,	dobutamine_exposure,	sodium_chloride_exposure,	levothyroxine_exposure,	dexamethasone_exposure,	
         phenobarbitol_exposure,	hydrocortisone_exposure,	ibuprofen_exposure,	chlorathiazide_diuril__exposure,	midazolam_versed__exposure,	
         potassium_chloride_exposure,	hepatitis_b_immune_globulin_exposure,	citric_acid_sodium_citrate_bicitra__exposure,	
         pancuronium_or_vecurionium_exposure,	insulin_exposure,	immunoglobulin_other_exposure,	multi_vitamin_no_iron_exposure,	
         levocarnitine_exposure,	digoxin_exposure,	hydraluronidase_hydase__exposure,	lorazapam_ativan__exposure,	sodium_bicarbonate_exposure,	
         hydrochlorathiazide_hydrodiuril__exposure,	spironolactone_hydrochlorathiazide_aldactazide__exposure,	tocopherol_acetate_vit_e__exposure, acute_abx_exposure_count,acute_antiviral_exposure,
  acute_aminoglycoside_exposure,acute_penicillin_exposure,acute_macrolide_exposure,acute_firstgencephalosporin_exposure,acute_fourthgencephalosporin_exposure,
  acute_thirdgencephalosporin_exposure,acute_secondgencephalosporin_exposure,acute_lincosamide_exposure,acute_antifungal_exposure,
  acute_carbapenem_exposure,acute_other_antimicrobials_exposure,acute_other_non_antimicrobials_exposure,acute_glycopeptide_exposure,
  acute_med_exposure_count,acute_morphine_exposure,acute_fentanyl_exposure,acute_surfactant_exposure,acute_acetaminophen_exposure,acute_caffeine_exposure,
  acute_cholecalciferol_vit_d__exposure,acute_vitamin_a_exposure,acute_famotidine_pepcid__exposure,acute_iron_exposure, acute_calcium_gluconate_exposure,
  acute_dopamine_exposure, acute_adenosine_exposure, acute_cosyntropin_cortrosyn_acth__exposure, acute_bumetanide_bumax__exposure, acute_other_exposure,
  acute_indomethacin_exposure, acute_multi_vitamin_with_iron_exposure, acute_lasix_exposure, acute_atropine_exposure, acute_dobutamine_exposure, 
  acute_sodium_chloride_exposure, acute_levothyroxine_exposure, acute_dexamethasone_exposure, acute_phenobarbitol_exposure, acute_hydrocortisone_exposure,
  acute_ibuprofen_exposure, acute_chlorathiazide_diuril__exposure, acute_midazolam_versed__exposure, acute_potassium_chloride_exposure, 
  acute_hepatitis_b_immune_globulin_exposure, acute_citric_acid_sodium_citrate_bicitra__exposure, acute_pancuronium_or_vecurionium_exposure,
  acute_insulin_exposure, acute_immunoglobulin_other_exposure,	acute_multi_vitamin_no_iron_exposure, 
             acute_levocarnitine_exposure,acute_digoxin_exposure,acute_hydraluronidase_hydase__exposure,acute_lorazapam_ativan__exposure,
             acute_sodium_bicarbonate_exposure,acute_hydrochlorathiazide_hydrodiuril__exposure,
             acute_spironolactone_hydrochlorathiazide_aldactazide__exposure, acute_tocopherol_acetate_vit_e__exposure)
       
gm_dev_sample_data <- gm_dev_sample_data %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(num_var, names(.))), as.numeric))

# Create subject vector
gm_dev_subject_vector <- as.character(ps_dol60_meta_filt$Patient)
names(gm_dev_subject_vector) <- rownames(ps_dol60_meta_filt)

set.seed(42)
baseline_diet_var <- c('DOL','Gestational_age_weeks',	'Birthweight','Sex',	'Delivery_route',	'Cohort', 'Necrotizing_enterocolitis', 'Site','frac_days_milk','cum_days_formula', 
'cum_days_milk', 'frac_days_formula')
mat_var <- c('Maternal_race','Maternal_age_at_delivery',	'Gravida',	'Para',	'Maternal_smoking',	'Maternal_alcohol',	'Maternal_substance_abuse',	'Maternal_preeclampsia',
'Maternal_diabetes',	'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe',	'mat_outpt_antimicrobial',	'Multiple_births')

abx_var <- c('abx_group','abx_exposure_count','antiviral_exposure',	'aminoglycoside_exposure',	'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	
'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure','lincosamide_exposure',	'antifungal_exposure',	'carbapenem_exposure',
	'other_antimicrobials_exposure',	'other_non_antimicrobials_exposure',	'glycopeptide_exposure')
med_var <- c('med_exposure_count',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure')


acute_abx_var <- c('abx_group_updated','acute_abx_exposure_count','acute_antiviral_exposure', 'acute_aminoglycoside_exposure','acute_penicillin_exposure',
'acute_macrolide_exposure','acute_firstgencephalosporin_exposure','acute_fourthgencephalosporin_exposure','acute_thirdgencephalosporin_exposure',
'acute_secondgencephalosporin_exposure','acute_lincosamide_exposure','acute_antifungal_exposure','acute_carbapenem_exposure',
'acute_other_antimicrobials_exposure','acute_other_non_antimicrobials_exposure','acute_glycopeptide_exposure')

acute_med_var <- c('acute_med_exposure_count','acute_morphine_exposure','acute_fentanyl_exposure','acute_surfactant_exposure',
'acute_acetaminophen_exposure','acute_caffeine_exposure','acute_cholecalciferol_vit_d__exposure','acute_vitamin_a_exposure',
'acute_famotidine_pepcid__exposure','acute_iron_exposure', 'acute_calcium_gluconate_exposure', 'acute_dopamine_exposure', 'acute_adenosine_exposure', 
'acute_cosyntropin_cortrosyn_acth__exposure', 'acute_bumetanide_bumax__exposure', 'acute_other_exposure', 'acute_indomethacin_exposure', 
'acute_multi_vitamin_with_iron_exposure', 'acute_lasix_exposure', 'acute_atropine_exposure', 'acute_dobutamine_exposure', 'acute_sodium_chloride_exposure',
'acute_levothyroxine_exposure', 'acute_dexamethasone_exposure', 'acute_phenobarbitol_exposure', 'acute_hydrocortisone_exposure', 'acute_ibuprofen_exposure', 
'acute_chlorathiazide_diuril__exposure', 'acute_midazolam_versed__exposure', 'acute_potassium_chloride_exposure', 'acute_hepatitis_b_immune_globulin_exposure', 
'acute_citric_acid_sodium_citrate_bicitra__exposure', 'acute_pancuronium_or_vecurionium_exposure', 'acute_insulin_exposure', 'acute_immunoglobulin_other_exposure',	'acute_multi_vitamin_no_iron_exposure', 
             'acute_levocarnitine_exposure','acute_digoxin_exposure','acute_hydraluronidase_hydase__exposure','acute_lorazapam_ativan__exposure',
             'acute_sodium_bicarbonate_exposure','acute_hydrochlorathiazide_hydrodiuril__exposure',
             'acute_spironolactone_hydrochlorathiazide_aldactazide__exposure','acute_tocopherol_acetate_vit_e__exposure')

gm_dev_repeat_perm_mat <- PERMANOVA_repeat_measures_original(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(baseline_diet_var, mat_var),
  permutations=999, ncores=20)
print("Done with PERMANOVA on maternal, baseline, and diet variables")
write.csv(gm_dev_repeat_perm_mat, "260203_GM_dev_baseline_diet_mat_repeat_permanova_results.csv")

gm_dev_repeat_perm_abx <- PERMANOVA_repeat_measures_original(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(abx_var),
  permutations=999, ncores=20)
write.csv(gm_dev_repeat_perm_abx, "260203_GM_Dev_repeat_permanova_BH_adj_results_abx.csv")
print("Done with PERMANOVA on abx variables")

gm_dev_repeat_perm_med <- PERMANOVA_repeat_measures_original(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(med_var),
  permutations=999, ncores=20)
write.csv(gm_dev_repeat_perm_abx, "260203_GM_Dev_repeat_permanova_BH_adj_results_med.csv")
print("Done with PERMANOVA on med variables")

gm_dev_repeat_perm_acute_abx <- PERMANOVA_repeat_measures_original(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(acute_abx_var),
  permutations=999, ncores=20)
write.csv(gm_dev_repeat_perm_acute_abx, "260203_GM_Dev_repeat_permanova_BH_adj_results_acute_abx.csv")
print("Done with PERMANOVA on acute abx variables")

gm_dev_repeat_perm_acute_med <- PERMANOVA_repeat_measures_original(
  gm_dev_dol60.dist,
  gm_dev_subject_vector, subject_data = gm_dev_subject_data2,
  sample_data = gm_dev_sample_data,
  metadata_order = c(acute_med_var),
  permutations=999, ncores=20)
write.csv(gm_dev_repeat_perm_acute_med, "260203_GM_Dev_repeat_permanova_BH_adj_results_acute_med.csv")
print("Done with PERMANOVA on acute med variables")

```

Beta Diversity PCoA Plots (Related to Fig. 1C, Supplementary Figs. 1F,H)
```{r}
######## Plot ordination plot #####
gm_dev_ordination <- phyloseq::ordinate(merged_ps, method='PCoA', distance='bray')

pcoa_gm_dev <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                         type="samples", color= "DOL_bin")+
  geom_point(shape=19, size=3)+
  #scale_color_manual(values = dol_bin_color_vector) + 
  scale_color_manual(values=natparks.pals("Arches"))+ 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev <- ggMarginal(pcoa_gm_dev, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_DOL_bin_P4_joined.pdf", plot = pcoa_gm_dev, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_abx <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                             type="samples", color= "abx_group_updated")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_abx) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev_abx <- ggMarginal(pcoa_gm_dev_abx, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_abx_exposure_group_P4_joined.pdf", plot = pcoa_gm_dev_abx, bg = "transparent", width = 9, height = 6)

pcoa_gm_dev_abx_count <- phyloseq::plot_ordination(merged_ps, gm_dev_ordination,
                                             type="samples", color= "abx_exposure_count")+
  geom_point(shape=19, size=3)+
  scale_color_gradientn(colors = natparks.pals("Arches")) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 16),
        legend.text = element_text(size=14),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))
ggsave("260202_GM_Dev_PCoA_abx_exposure_count_P4_joined.pdf", plot = pcoa_gm_dev_abx_count, bg = "transparent", width = 9, height = 6)


### Including term infants #####
gm_dev_ordination2 <- phyloseq::ordinate(ps_infants, method='PCoA', distance='bray')

pcoa_gm_dev_abx_term <- phyloseq::plot_ordination(ps_infants, gm_dev_ordination2,
                                             type="samples", color= "abx_group_names")+
  geom_point(shape=19, size=3)+
  stat_ellipse()+
  scale_color_manual(values = custom_colors_names) + 
  theme_classic()+
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        legend.title = element_text(size = 18),
        legend.text = element_text(size=16),
        legend.position = 'left',
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=3))

pcoa_gm_dev_abx_term <- ggMarginal(pcoa_gm_dev_abx_term, type='boxplot', groupFill=TRUE, size=10)
ggsave("260202_GM_Dev_PCoA_abx_exposure_group_tern_joined.pdf", plot = pcoa_gm_dev_abx_term, bg = "transparent", width = 9, height = 6)

```


## Beta diversity comparisons (DELETE in Public github)
```{r}

## Pairwise comparisons by DOL bin ##
group_levels <- levels(factor(ps_infants_meta$abx_group_names))
group_pairs <- combn(group_levels, 2, simplify = FALSE)

# Loop through each pair
for (pair in group_pairs) {
  # Subset the phyloseq object
  print(pair)
  ps_subset <- subset_samples(ps_infants, abx_group_names %in% pair)
  ps_subset <- prune_samples(sample_sums(ps_subset) > 0, ps_subset)  # avoid empty samples
  ps_subset
  
  # Perform ordination
  ord <- ordinate(ps_subset, method = "PCoA", distance = "bray")
  
  # Base PCoA plot
  p <- plot_ordination(ps_subset, ord, type = "samples", color = "abx_group_names") +
    geom_point(shape = 19, size = 3) +
    stat_ellipse() +
    scale_color_manual(values = custom_colors_names[pair]) +
    theme_classic() +
    theme(
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 16),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      legend.position = "left",
      plot.title = element_text(size = 16, hjust = 0.5),
      panel.border = element_rect(fill = NA, colour = "black", size = 1)
    ) +
    labs(title = paste(pair[1], "vs", pair[2]))
  
  # Add marginal boxplots
  p_with_marginals <- ggMarginal(p, type = "boxplot", groupFill = TRUE, size = 10)
  
  # Save the plot
  file_name <- paste0("PCoA_", pair[1], "_vs_", pair[2], ".pdf")
  ggsave(file.path(file_name), plot = p_with_marginals, bg = "transparent", width = 9, height = 6)
  
  message("Saved: ", file_name)
}


ps_infants1 <- ps_infants %>% subset_samples(DOL_bin == "DOL 1 - 10")
plot_pcoa(ps_obj = ps_infants1, group_var = "abx_group_names", custom_colors = custom_colors_names, 
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants2 <- ps_infants %>% subset_samples(DOL_bin == "DOL 11 - 20")
plot_pcoa(ps_obj = ps_infants2, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants3 <- ps_infants %>% subset_samples(DOL_bin == "DOL 21 - 30")
plot_pcoa(ps_obj = ps_infants3, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants4 <- ps_infants %>% subset_samples(DOL_bin == "DOL 31 - 40")
plot_pcoa(ps_obj = ps_infants4, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants5 <- ps_infants %>% subset_samples(DOL_bin == "DOL 41 - 50")
plot_pcoa(ps_obj = ps_infants5, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")

ps_infants6 <- ps_infants %>% subset_samples(DOL_bin == "DOL 51 - 60")
plot_pcoa(ps_obj = ps_infants6, group_var = "abx_group_names", custom_colors = custom_colors_names,
          adonisformula = "ps.dist ~ Gestational_age_weeks + Site + abx_group_names")


```

Repeat Measure PERMANOVA Summary Figure (Related to Fig. 1D)
```{r}
DOL <- c("DOL")
baseline_var <- c('Gestational_age_weeks',	'Birthweight','Sex',	'Delivery_route',	'Cohort', 'Necrotizing_enterocolitis', 'Site')
diet_var <- c('cum_days_formula', 'cum_days_milk', 'frac_days_formula','frac_days_milk')
mat_var <- c('Maternal_race','Maternal_age_at_delivery',	'Gravida',	'Para',	'Maternal_smoking',	'Maternal_alcohol',	'Maternal_substance_abuse',	'Maternal_preeclampsia',
'Maternal_diabetes',	'Maternal_chorioamnionitis', 'Maternal_hypertension', 'mat_inpt_antimicrobe',	'mat_outpt_antimicrobial',	'Multiple_births')
abx_var <- c('abx_group_updated','abx_exposure_count','aminoglycoside_exposure',	'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	
'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure','lincosamide_exposure',		'carbapenem_exposure',
	'other_antimicrobials_exposure', 'glycopeptide_exposure','acute_abx_exposure_count', 'acute_aminoglycoside_exposure', 'acute_penicillin_exposure','acute_macrolide_exposure', 'acute_firstgencephalosporin_exposure','acute_fourthgencephalosporin_exposure',
'acute_thirdgencephalosporin_exposure','acute_secondgencephalosporin_exposure','acute_lincosamide_exposure','acute_carbapenem_exposure','acute_other_antimicrobials_exposure',
'acute_other_non_antimicrobials_exposure','acute_glycopeptide_exposure')
medications_var <- c('med_exposure_count', 'other_non_antimicrobials_exposure',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure', 'antifungal_exposure', 'acute_med_exposure_count','acute_morphine_exposure','acute_fentanyl_exposure','acute_surfactant_exposure',
'acute_acetaminophen_exposure','acute_caffeine_exposure','acute_cholecalciferol_vit_d__exposure',
'acute_vitamin_a_exposure','acute_famotidine_pepcid__exposure','acute_iron_exposure',
'acute_calcium_gluconate_exposure','acute_dopamine_exposure','acute_bumetanide_bumax__exposure',
'acute_other_exposure','acute_indomethacin_exposure','acute_multi_vitamin_with_iron_exposure',
'acute_lasix_exposure','acute_atropine_exposure','acute_dobutamine_exposure',
'acute_sodium_chloride_exposure','acute_levothyroxine_exposure','acute_dexamethasone_exposure',
'acute_phenobarbitol_exposure','acute_hydrocortisone_exposure','acute_ibuprofen_exposure',
'acute_chlorathiazide_diuril__exposure','acute_midazolam_versed__exposure',
'acute_potassium_chloride_exposure','acute_citric_acid_sodium_citrate_bicitra__exposure',
'acute_pancuronium_or_vecurionium_exposure','acute_insulin_exposure', "acute_multi_vitamin_no_iron_exposure", "acute_levocarnitine_exposure", "acute_digoxin_exposure", "acute_lorazapam_ativan__exposure", "acute_sodium_bicarbonate_exposure",
'acute_antifungal_exposure', 'acute_antiviral_exposure', 'antiviral_exposure')

#### Repeat measures PERMANOVA results were compiled manually ######
results <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/Beta_diversity/260203_GM_Dev_repeat_permanova_results_combined.xlsx")
results$p_adj_BH_redo <- p.adjust(results[["Pr(>F)"]], method = "BH")
results$p_adj_BH_redo <- as.numeric(results$p_adj_BH_redo)
write.csv(results, "260209_Repeat_measures_PERMANOVA_results_p_adj_redo.csv", row.names = F)

results <- results %>%
  mutate(significant = p_adj_BH_redo < 0.05,
         Category = case_when(Variable %in% DOL ~ "DOL",
                              Variable %in% baseline_var ~ "Baseline",
                              Variable %in% diet_var ~ "Diet",
                              Variable %in% mat_var ~ "Maternal",
                              Variable %in% abx_var ~ "Antibiotic",
                              Variable %in% medications_var ~ "Non-antibiotic medication"))
var_order <- c("DOL","Baseline", "Diet", "Maternal","Antibiotic", "Non-antibiotic medication")

results$Category <- factor(results$Category, levels = var_order)

results_plot <- results %>% mutate(R2_percent = R2 * 100,
                                   sig_label = case_when(p_adj_BH_redo <= 0.001 ~ "***",
                                                         p_adj_BH_redo <= 0.01 ~ "**",
                                                         p_adj_BH_redo <= 0.05 ~ "*")) %>%
  filter(p_adj_BH_redo < 0.05) %>%
  arrange(desc(R2_percent)) %>%
  mutate(Label = factor(Label, levels = unique(Label))) %>%
  ungroup()

perm_cat_vector = c("DOL" = "#3A1F46","Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

permanova_sign_var_p <- ggplot(results_plot, aes(x = Label, y = R2_percent, fill = Category)) +
  geom_col(width = 0.8, color = "black") + 
  geom_text(aes(label = sig_label, y = R2_percent + 0.15), size = 5, vjust = 0) +
  scale_fill_manual(values=perm_cat_vector) + 
  labs(x = NULL, y = "Variance explained (%)", fill = NULL) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    axis.line.x = element_line(color = "black", linewidth = 0.6),
    axis.ticks.x = element_line(color = "black", linewidth = 0.4),
    plot.margin = ggplot2::margin(t = 10, r = 10, b = 40, l = 10))  # extra space for category bar
ggsave("260205_GM_Dev_repeat_measure_PERMANOVA_sign_var.pdf", plot = permanova_sign_var_p, bg = "transparent", width = 12, height = 6)
```


Run Linear Mixed Models (MaAslin2) to test association between species and metadata (Related to Fig. 2A-C, Supplementary Fig. 2)
```{r}
############### ----------------------------- Taxa ----------------------##################################################

# results_plot can be found in the code block above
repeat_permanova_sign_var <- results_plot$Variable

pterm_infants_otu <- data.frame(otu_table(merged_ps), check.names = FALSE) 
pterm_infants_meta <- data.frame(sample_data(merged_ps), check.names = FALSE)

cat_var <- c("Patient",'Sex','Delivery_route', 'Necrotizing_enterocolitis', 'Site','Maternal_substance_abuse','Maternal_preeclampsia','Maternal_diabetes', 'Maternal_chorioamnionitis', 'Maternal_hypertension','mat_inpt_antimicrobe', 'mat_outpt_antimicrobial','Multiple_births','Cohort', 'Site', 'abx_group_updated',
             'abx_group_names', 'Study')

num_var <- c('Gestational_age_weeks', 'Birthweight','DOL','abx_exposure_count', 'antiviral_exposure', 'antiviral_exposure',	'aminoglycoside_exposure',	
'penicillin_exposure',	'macrolide_exposure',	'firstgencephalosporin_exposure',	'fourthgencephalosporin_exposure',	'thirdgencephalosporin_exposure',	'secondgencephalosporin_exposure',	
'lincosamide_exposure',	'antifungal_exposure',	'carbapenem_exposure',	'other_antimicrobials_exposure',	'other_non_antimicrobials_exposure','glycopeptide_exposure', 'med_exposure_count',
	'morphine_exposure',	'fentanyl_exposure',	'surfactant_exposure',	'acetaminophen_exposure',	'caffeine_exposure',	'cholecalciferol_vit_d__exposure',	'vitamin_a_exposure',	
  'famotidine_pepcid__exposure',	'iron_exposure',	'calcium_gluconate_exposure',	'dopamine_exposure',	'adenosine_exposure',	'cosyntropin_cortrosyn_acth__exposure',	
  'bumetanide_bumax__exposure',	'indomethacin_exposure',	'multi_vitamin_with_iron_exposure',	'lasix_exposure',	'atropine_exposure',	'dobutamine_exposure',	'sodium_chloride_exposure',	
  'levothyroxine_exposure',	'dexamethasone_exposure',	'phenobarbitol_exposure',	'hydrocortisone_exposure',	'ibuprofen_exposure',	'chlorathiazide_diuril__exposure',	'midazolam_versed__exposure',	
  'potassium_chloride_exposure',	'hepatitis_b_immune_globulin_exposure',	'citric_acid_sodium_citrate_bicitra__exposure',	'pancuronium_or_vecurionium_exposure',	'insulin_exposure',	
  'immunoglobulin_other_exposure',	'multi_vitamin_no_iron_exposure',	'levocarnitine_exposure',	'digoxin_exposure',	'hydraluronidase_hydase__exposure',	'lorazapam_ativan__exposure',	
  'sodium_bicarbonate_exposure',	'hydrochlorathiazide_hydrodiuril__exposure',	'spironolactone_hydrochlorathiazide_aldactazide__exposure',	'tocopherol_acetate_vit_e__exposure', 
  'cum_days_formula', 'cum_days_milk', 'frac_days_formula','frac_days_milk', "acute_fentanyl_exposure", "acute_cholecalciferol_vit_d__exposure", "acute_dopamine_exposure", "acute_levothyroxine_exposure", "acute_hydrocortisone_exposure", "acute_insulin_exposure", "acute_surfactant_exposure", "acute_iron_exposure", "acute_indomethacin_exposure", "acute_dexamethasone_exposure",
"acute_citric_acid_sodium_citrate_bicitra__exposure", "acute_lorazapam_ativan__exposure", )

pterm_infants_meta <- pterm_infants_meta %>%
  mutate(across(all_of(intersect(cat_var, names(.))), as.factor)) %>%
  mutate(across(all_of(intersect(results_plot$Variable, names(.))), as.numeric))

pterm_infants_meta$abx_group_updated <- factor(pterm_infants_meta$abx_group_names, levels = c("ABN", "EEO", "EAS"))

maaslin_taxa_results <- Maaslin2(input_data = pterm_infants_otu,
                                input_metadata = pterm_infants_meta,
                                transform = "LOG",
                                normalization = "NONE",
                                min_abundance = 0.0001,
                                min_prevalence = 0.1,
                                output = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/MaAslin2/Taxa/all_sign_var",
                                fixed_effects = c("abx_group_updated",repeat_permanova_sign_var),
                                random_effects = c("Patient", "Site", "Study"))

############ --------------- Coefficient Plots ------------------------------#############################################
taxa_sign_out <- read_excel("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/MaAslin2/Taxa/all_sign_var/significant_results.xlsx")

num_samples <- 2406

taxa_sign_out_filt <- taxa_sign_out %>%
  filter(qval < 0.05) %>%
  mutate(
    # grab text AFTER "g__" up to the next dot
    genus  = str_extract(feature, "(?<=g__)[^.]+"),
    genus  = if_else(is.na(genus) | genus == "", "Unknown", genus),
    value  = as.factor(value),
    feature = gsub(".*s__", "", feature)
  )
write.csv(taxa_sign_out_filt, "260205_MaAslin2_all_sign_var_results.csv", row.names = F)

unique_taxa <- unique(taxa_sign_out_filt$feature)
unique_genera <- unique(taxa_sign_out_filt$genus)

maaslin_taxa_abxGroup <- taxa_sign_out %>% filter(metadata == "abx_group_updated") %>% 
  #filter(qval < 0.05) %>%
  mutate(
    # grab text AFTER "g__" up to the next dot
    genus  = str_extract(feature, "(?<=g__)[^.]+"),
    genus  = if_else(is.na(genus) | genus == "", "Unknown", genus),
    value  = as.factor(value),
    feature = gsub(".*s__", "", feature)
  ) %>%
  arrange(desc(coef))

#obtain min and max coefficient values (needed for plotting later)
min <- min(maaslin_taxa_abxGroup$coef)
max <- max(maaslin_taxa_abxGroup$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
maaslin_taxa_abxGroup <- maaslin_taxa_abxGroup %>% mutate(species = fct_reorder(feature, coef))

maaslin_taxa_abxGroup$coefmax <- maaslin_taxa_abxGroup$coef + maaslin_taxa_abxGroup$stderr
maaslin_taxa_abxGroup$coefmin <- maaslin_taxa_abxGroup$coef - maaslin_taxa_abxGroup$stderr

maaslin_taxa_abxGroup$CI95upper <- maaslin_taxa_abxGroup$coef + 1.96*maaslin_taxa_abxGroup$stderr
maaslin_taxa_abxGroup$CI95lower <- maaslin_taxa_abxGroup$coef - 1.96*maaslin_taxa_abxGroup$stderr

maaslin_taxa_abxGroup$value <- factor(maaslin_taxa_abxGroup$value, levels = c("Early only", "Early and after"))
write.csv(maaslin_taxa_abxGroup, "260205_MaAslin2_sign_taxa_abx_group.csv", row.names = F)

# Plot coefficients
x_min_lim <- (floor(min / 5) * 6) 
x_max_lim <- (ceiling(max / 5) * 6) 
colors_list <- c("#44637D","#AAC9ED")

abxgroup_coeff_p <- ggplot(maaslin_taxa_abxGroup, aes(x=coef, y=species,color=value))+
  geom_point(size = 3)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.5)+
  scale_color_manual(values=colors_list)+
  theme_linedraw()+
  theme(axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=17),
        axis.title.x = element_text(size=17),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 16),
        legend.position = "bottom",
        plot.title = element_text(size=15),
        panel.border = element_rect(fill=NA, colour="black", size=1))+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  xlab('Coefficient') + ggtitle("Antibiotic Exposure Group")
ggsave("260205_GM_dev_abx_group_MaAslin2_taxa_coeff.pdf", plot = abxgroup_coeff_p, bg = "transparent", width = 11, height = 12)

#### Generate prevalence plot ###
### Access species abundance data 
abund_gmdev <- psmelt(merged_ps)

abund_gmdev <- abund_gmdev %>%  mutate(
    genus = str_extract(OTU, "g__[^|]+"),  # extract 'g__Genus'
    genus = str_remove(genus, "g__"),          # remove 'g__' prefix
    genus = if_else(is.na(genus), "Unknown", genus),
    OTU=gsub(".*s__", "", OTU))

abund_abx_group <- subset(abund_gmdev, OTU %in% maaslin_taxa_abxGroup$feature)

# Re-order feature abs abundances according to coefficient value
# This order is preserved in filename, which was previously ordered: 
order_abx_group <- data.frame(OTU = unique(maaslin_taxa_abxGroup$feature))
# sort = F makes sure that the order is based on order_index df which is ordered by coefficient value
abund_abx_group.df <- merge(order_abx_group, abund_abx_group, by="OTU", sort = F)
abund_abx_group.df$OTU <- as.factor(abund_abx_group.df$OTU)
abund_abx_group.df <- abund_abx_group.df %>% mutate(OTU = fct_relevel(OTU, order_abx_group$OTU))

# For each significant feature in each sample, set AbunNotZero == 1 if its 
# abundance is greater than 0. 
abund_abx_group.df$AbunNotZero <- 1*(abund_abx_group.df$Abundance > 0)

# Summarize pathway prevalence by variable of interest
abund_abx_group.df.n0 <- abund_abx_group.df %>% 
  group_by(OTU, abx_group) %>%
  dplyr::summarise(N = n(), SumNotZero = sum(AbunNotZero)) %>%
  mutate(PercentNotZero = 100*(SumNotZero / N)) %>%
  ungroup()
write.csv(abund_abx_group.df.n0, "260205_GM_dev_abx_group_maaslin2_species_prevalence.csv", 
          row.names = F)

prev_p_abx_group <- ggplot(abund_abx_group.df.n0, aes(x = abx_group, y=OTU, fill=PercentNotZero))+
  geom_tile()+
  scale_fill_distiller(palette ='RdBu', type='seq', breaks = c(0,50,100), 'Prevalence (%)')+
  theme_classic()+
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(size=17, angle=45, hjust=1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=17),
        legend.text = element_text(size = 18),
        panel.border = element_rect(fill=NA, colour="black", size=1))+
  scale_y_discrete(limits=rev)
ggsave("260205_GM_dev_abx_group_MaAslin2_taxa_prev.pdf", plot = prev_p_abx_group, bg = "transparent", width = 11, height = 12)

combined_abx_group <- abxgroup_coeff_p | prev_p_abx_group 
ggsave("260205_GM_dev_MaAslin2_taxa_plot.pdf", plot = combined_abx_group, bg = "transparent", width = 12, height = 6)

############---------------- Other numerical variables ------------------------------------------#########
library(tidytext) # v0.4.3
library(stringr) # v1.6.0
library(ggh4x) # v0.3.1

perm_cat_vector = c("DOL" = "#3A1F46", "Baseline" = "#7F4B89", "Diet" = "#F3DAE4", "Antibiotic" = "#E3A5D6", "Non-antibiotic medication" = "#7C5467")

taxa_sign_out_num <- taxa_sign_out_filt %>% filter(metadata != "abx_group_updated") %>% 
  mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 
write.csv(taxa_sign_out_num, "260205_sign_MaAslin2_taxa.csv", row.names = F)

taxa_sign_out_num <- read_excel("260205_sign_MaAslin2_taxa.xlsx")
taxa_sign_out_num <- taxa_sign_out_num %>% mutate(direction = if_else(coef > 0, "Positive", "Negative"),
         feature = reorder_within(feature, coef, metadata)) 


taxa_sign_out_num$metadata_category <- factor(taxa_sign_out_num$metadata_category, levels = names(perm_cat_vector))

#obtain min and max coefficient values (needed for plotting later)
min <- min(taxa_sign_out_num$coef)
max <- max(taxa_sign_out_num$coef)

### Generate coefficient plot ####
# Calculate 95% CI for coefficients
#maaslin_taxa <- taxa_sign_out_num %>% mutate(species = fct_reorder(feature, coef))

taxa_sign_out_num$coefmax <- taxa_sign_out_num$coef + taxa_sign_out_num$stderr
taxa_sign_out_num$coefmin <- taxa_sign_out_num$coef - taxa_sign_out_num$stderr

taxa_sign_out_num$CI95upper <- taxa_sign_out_num$coef + 1.96*taxa_sign_out_num$stderr
taxa_sign_out_num$CI95lower <- taxa_sign_out_num$coef - 1.96*taxa_sign_out_num$stderr

taxa_sign_out_num$direction <- factor(taxa_sign_out_num$direction, levels = c("Positive", "Negative"))

genus_color_vector <- c(
  "Veillonella"      = "#A8A55C",  
  "Streptococcus"      = "#68434E",  
  "Clostridium"    = "#B07159",  
  "Klebsiella"     = "#2F4B7A",  
  "Enterobacter"      = "#8C2B0E",  
  "Staphylococcus"      = "#B47E83",  
  "Enterococcus"  = "#F2A450",  
  "Escherichia"    = "#9F7E59",  
  "Cutibacterium"          = "#6EA57B", 
  "Finegoldia"       = "#8DA0CB",  
  "Clostridioides"          = "#088096"  
)

taxa_dol_coeff <- taxa_sign_out_num %>% filter(metadata_category == "DOL") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#3A1F46", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260205_GM_dev_MaAslin2_taxa_DOL_var_coeff_plot.pdf", plot = taxa_dol_coeff, bg = "transparent", width = 5, height = 4)

taxa_baseline_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Baseline") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "#7F4B89", color = "white"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260205_GM_dev_MaAslin2_taxa_baseline_var_coeff_plot.pdf", plot = taxa_baseline_coeff, bg = "transparent", width = 4, height = 3)

taxa_diet_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Diet") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#F3DAE4", color = "black"),
    strip.text = element_text(size = 11, face = "bold", color = "black"),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.box = "horizontal")
ggsave("260205_GM_dev_MaAslin2_taxa_diet_var_coeff_plot.pdf", plot = taxa_diet_coeff, bg = "transparent", width = 9, height = 4)

taxa_abx_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Antibiotic") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#E3A5D6", color = "black"),
    strip.text = element_text(size = 12, face = "bold", color = "black"),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.box = "horizontal")
ggsave("260205_GM_dev_MaAslin2_taxa_abx_var_coeff_plot.pdf", plot = taxa_abx_coeff, bg = "transparent", width = 18, height = 6)

taxa_med_coeff <- taxa_sign_out_num %>% filter(metadata_category == "Non-antibiotic medication") %>% ggplot(., aes(x = coef , y = feature)) +
  geom_point(aes(color = genus), alpha = 0.9, size = 3) +
  geom_vline(xintercept = 0, linetype='dotted', color='grey', size=1.2)+
  geom_errorbarh(aes(xmax = CI95upper, xmin = CI95lower, height =0.1))+
  facet_wrap2(~ Label, labeller = as_labeller(function(x) {str_wrap(x, width = 20)}), scales = "free_y", ncol = 4) +
  scale_color_manual(values = genus_color_vector, name = "Genus") +
  scale_size_continuous(range = c(2, 10)) +
  scale_y_reordered() +
  theme_linedraw(base_size = 13)+
  labs(
    x = "Model Coefficient",
    y = NULL,
    color = "Direction") +
  theme(
    strip.background = element_rect(fill = "#7C5467", color = "black"),
    strip.text = element_text(size = 9, face = "bold", color = "white"),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 12, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none")
ggsave("260205_GM_dev_MaAslin2_taxa_med_var_coeff_plot.pdf", plot = taxa_med_coeff, bg = "transparent", width = 22, height = 8)

###################### ---------------------------- ShortBRED-------------------###########################################
## Merge term infant ARGs with class
arg.map <- read.delim('~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Shortbred/ShortBRED_mapping.txt',header = F)

## Merge preterm infant ARGs with class
arg.byclass <- merge(arg_df, arg.map, by.x = 'Family', by.y = 'V1') %>% 
  dplyr::rename(Class = V2)

arg.byclass_preterm <- arg.byclass %>% 
  dplyr::group_by(Sample_ID_R_final, Class) %>% 
  dplyr::summarise(Count = sum(Count))

arg.byclass_preterm.wide <- dcast(arg.byclass_preterm, Sample_ID_R_final ~ Class, value.var = 'Count')
arg.byclass_preterm.wide[is.na(arg.byclass_preterm.wide)] <- 0
row.names(arg.byclass_preterm.wide) <- arg.byclass_preterm.wide$Sample_ID_R_final
arg.byclass_preterm.wide$Sample_ID_R_final <- NULL


# Create metadata file
merged_ps_meta <- data.frame(sample_data(merged_ps))

merged_ps_meta_filt <- merged_ps_meta %>% select(Sample_ID_R_final, Patient, Cohort, Site, DOL, DOL_bin, Gestational_age_weeks, Birthweight, Sex, Delivery_route, abx_exposure_count, abx_group_updated, abx_group_names)

merged_ps_meta_filt$abx_group_updated <- factor(merged_ps_meta_filt$abx_group_updated, levels = c("None", "Early only", "Early and after"))

############--------- Run MaAslin2 -------------------------------############################

shortbred_preterm <- Maaslin2(input_data = arg.byclass_preterm.wide,
                                input_metadata = merged_ps_meta_filt,
                                transform = "LOG",
                                normalization = "NONE",
                                min_abundance = 0.0001,
                                min_prevalence = 0.1,
                                output = "~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/Re-grouped/MaAslin2/ShortBRED/Preterm_site_cohort_random_effect_v2",
                                fixed_effects = c("Gestational_age_weeks","abx_group_updated, None"),
                                random_effects = c("Patient", "Site", "Cohort"))


```


Random Forest Analysis to predict microbiota age (Related to Fig. 2D,E and Supplementary Fig. 3)
```{r}
library(randomForest) # v4.7.1.2
library(dplyr) #  v1.2.0
library(stringr) # v1.6.0

load("~/Library/CloudStorage/Box-Box/pppp/GM_development_cohort/Analyses/Manuscript/No_AfterOnly/GA_matched/260109_Preterm_GA_matched_PS_object.RData")

# 1. Apply random forests regression to regress relative abundances of all species predicted by MetaPhlAn3 in infant stool samples against their chronological age 
## Default parameters used: ntree = 10,000, importance = TRUE
# 2. Fivefold cross-validation was performed using 'rfcv' function over 100 iterations to estimate the min number of features (taxa) needed to accurately predict microbiota age. This was done using 100 iterations of the "importance" function 
### Importance determined based on the percentage increase in mean-squared error of microbiota age prediction when the relative abundance values of each taxon were randomly permuted 
# 3. Sparse model consisting of the 50 most important features constructed and trained on half of abx-naive near-term infants that were randomly selected 
# 4. Validate model in remaining half of abx-naive near-term infants 
# 5. Apply model to each preterm infant group 

# Renormalize ps object to ensure that relative abundance sums to 100
ps_filt_dol60_renorm  <- transform_sample_counts(merged_ps, function(x) (x / sum(x))*100 )

# # Access taxa relative abundances, and sample metadata
# mtg_relabund.df <- data.frame(otu_table(ps_m3_mtg_long_renorm), check.names = FALSE)

# Load your taxa relative abundance dataframe and chronological age vector
relabund_meta_full.df <- psmelt(ps_filt_dol60_renorm) 
# Rename each OTU so that the name only has species names not the full phylogenetic breakdown 
relabund_meta.df <- relabund_meta_full.df %>%
    mutate(OTU = sapply(str_split(OTU, "\\|"), function(x) if(length(x) >= 7) x[7] else NA))

# Subset to only rows from ABN samples  
relabund_meta_naive.df <- relabund_meta.df %>%
  filter(abx_group_updated == "None")

# randomForest requires that each row represent a sample and columns represent relative abundance of taxa
## we will also need a column for the response variable, age
# This means we will need to reformat the output of psmelt() which outputs a long-format dataframe where:
## each row is a specific taxon in a sample and columns include sample ID and metadata like age 

# First pivot dataframe to wide format and 
taxa_abundance <- relabund_meta_naive.df %>%
  select(Sample, OTU, Abundance) %>%
  pivot_wider(names_from = OTU, values_from = Abundance, values_fill = 0) # values_fill = 0 makes sure that missing values (e.g., taxa absent in a sample) are set to 0

# Add age as a separate column
taxa_abundance <- taxa_abundance %>%
  left_join(
    relabund_meta_naive.df %>% select(Sample, DOL) %>% distinct(),
    by = "Sample"
  )

# 2. Randomly split the data upfront into training and validation sets
set.seed(142)  # For reproducibility
train_indices <- sample(1:nrow(taxa_abundance), size = 0.6 * nrow(taxa_abundance))
train_data <- taxa_abundance[train_indices, ]
validation_data <- taxa_abundance[-train_indices, ]

# Separate predictors and response for training set
x_train <- train_data %>% select(-Sample, -DOL)
y_train <- train_data$DOL

# Step 2: Perform 100 iterations for feature importance and RFCV
iterations <- 100
importance_list <- vector("list", iterations)
rfcv_results_list <- vector("list", iterations)

for (i in 1:iterations) {
  cat("Iteration:", i, "\n")
  
  # Fit Random Forest model on the training set
  rf_model <- randomForest(
    x = x_train,
    y = y_train,
    ntree = 10000,
    importance = TRUE
  )
  
  # Extract feature importance
  importance_df <- as.data.frame(importance(rf_model))
  importance_df <- importance_df %>%
    mutate(Taxon = rownames(importance_df),
           Iteration = i) %>%
    arrange(desc(IncNodePurity))
  
  # Append this iteration's importance data to the list
  importance_list[[i]] <- importance_df
  
  # Perform RFCV for this iteration
  rfcv_results <- rfcv(
    trainx = x_train,
    trainy = y_train,
    cv.fold = 10,
    step = 0.5
  )
  
  # Store the rfcv results with the iteration number
  rfcv_results_list[[i]] <- data.frame(
    Iteration = i,
    n_features = rfcv_results$n.var,
    error = rfcv_results$error.cv
  )
}

# Combine all feature importance results
feature_importance_df <- bind_rows(importance_list)

# Aggregate feature importance (Mean and SD for both %IncMSE and IncNodePurity)
aggregated_importance <- feature_importance_df %>%
  group_by(Taxon) %>%
  summarise(
    MeanIncNodePurity = mean(IncNodePurity, na.rm = TRUE),
    SDIncNodePurity = sd(IncNodePurity, na.rm = TRUE),
    MeanPercIncMSE = mean(`%IncMSE`, na.rm = TRUE),
    SDPercIncMSE = sd(`%IncMSE`, na.rm = TRUE)
  ) %>%
  arrange(desc(MeanPercIncMSE))
#write.csv(aggregated_importance, "260225_GM_Dev_RandomForests_aggregated_importance.csv", row.names = F)

# Combine all rfcv results
rfcv_results_df <- bind_rows(rfcv_results_list)

# Summarize rfcv results across iterations
aggregated_rfcv <- rfcv_results_df %>%
  group_by(n_features) %>%
  summarise(
    MeanError = mean(error, na.rm = TRUE),
    SDError = sd(error, na.rm = TRUE)
  )
#write.csv(aggregated_rfcv, "260225_GM_Dev_RandomForests_aggregated_rfcv.csv", row.names = F)

# Plot aggregated RFCV results
rfcv_error_variables_plot <- ggplot(aggregated_rfcv, aes(x = n_features, y = MeanError)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = MeanError - SDError, ymax = MeanError + SDError), width = 5) +
  theme_minimal() +
  labs(
    x = "Number of Variables",
    y = "Cross-validation error",
    title = "RFCV Error Across Variable Subsets"
  )
ggsave("260202_GMdev_RF_RFCV_aggregated_results_plot.pdf", plot = rfcv_error_variables_plot, bg = "transparent", width = 6, height = 6)

# Based on rfcv results, minimal improvement in predictive performance was observed when including taxa beyond the top N
top_taxa_n <- 50
top_taxa_rf <- aggregated_importance %>%
  arrange(desc(MeanPercIncMSE)) %>%
  dplyr::slice(1:top_taxa_n) %>%
  pull(Taxon)  # Select top N taxa based on mean importance

top50 <- aggregated_importance %>% arrange(desc(MeanPercIncMSE)) %>% slice_head(n=50)
write.csv(top50, "260225_GM_Dev_RandomForests_top50_aggregated_importance.csv", row.names = F)

## Plot the top 50 taxa by %IncMSE
top50_taxa_plot <- ggplot(top50, aes(x = reorder(Taxon, MeanPercIncMSE), y = MeanPercIncMSE)) +
  geom_point() +
  geom_errorbar(aes(ymin = MeanPercIncMSE - SDPercIncMSE, ymax = MeanPercIncMSE + SDPercIncMSE), width = 0.3) +
  coord_flip() +
  labs(
    x = "",
    y = "Increase in Mean Squared Error (%)",
    title = "Top 50 Age-Discriminatory Taxa"
  ) +
  theme_minimal()
ggsave("260202_GMdev_RF_top50_taxa_by_MSE_plot.pdf", plot = top50_taxa_plot, bg = "transparent", width = 6, height = 8)

# Step 3: Train sparse model on training set and validate on validation set
x_train_sparse <- x_train[, top_taxa_rf, drop = FALSE]

sparse_rf_model <- randomForest(
  x = x_train_sparse,
  y = y_train,
  ntree = 10000,
  importance = TRUE
)

#############-------------Apply sparse model to Early only and Early and After infants--------------#################
ps_early_only <- ps_filt_dol60_renorm %>% subset_samples(abx_group_updated == "Early only") # 801 samples; 171 infants
ps_early_and_after <- ps_filt_dol60_renorm %>% subset_samples(abx_group_updated == "Early and after") # 1312 samples; 183 infants
ps_abxN <- ps_filt_dol60_renorm %>% subset_samples(abx_group_updated == "None") # 282 samples; 38 infants
ps_abxN_val <- ps_abxN %>% subset_samples(Sample_ID_R_final %in% validation_data$Sample) # 113 samples; 35 infants 

predict_early_only <- predict_microbiota_age(ps_early_only, sparse_rf_model, top_taxa_rf)
predict_early_and_after <- predict_microbiota_age(ps_early_and_after, sparse_rf_model, top_taxa_rf)
predict_abxN_val <- predict_microbiota_age(ps_abxN_val, sparse_rf_model, top_taxa_rf)

predict_early_only <- data.frame(predict_early_only)
predict_early_and_after <- data.frame(predict_early_and_after)
predict_abxN_val <- data.frame(predict_abxN_val)

# Combine all datasets into one
combined_df <- bind_rows(
  predict_early_only %>% mutate(Group = "Early only"),
  predict_early_and_after %>% mutate(Group = "Early and after"),
  predict_abxN_val %>% mutate(Group = "None"))
#write.csv(combined_df, "260225_GM_Dev_RF_predicted_microbiota_ages.csv", row.names = F)

rf_plot <- ggplot(combined_df, aes(x = DOL, y = Predicted_Microbiota_Age, color = Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  labs(
    x = "Chronological age (days)",
    y = "Microbiota age (days)",
    title = "Microbiota Maturation Across Groups"
  ) +
  theme_minimal() +
  scale_color_manual(values = custom_colors_abx) + 
  theme(
    axis.title = element_text(size = 16, face = "bold"),      
    axis.text = element_text(size = 14),                      
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )
ggsave("260202_GMdev_RF_abxN_model_validation.pdf", plot = rf_plot, bg = "transparent", width = 10, height = 10)

####----------- Delete --------------------#################
########## Calculating and comparing MAZ values ##########
# Bin DOL into age bins (month-based)
combined_df2 <- combined_df %>%
  mutate(
    Month = case_when(
      DOL <= 30 ~ "Month 1",
      DOL <= 60 ~ "Month 2",
      TRUE ~ NA_character_
    )
  )

# Calculate mean and sd of predicted microbiota age for each bin in reference group
reference_group <- "None"
reference_stats <- combined_df2 %>%
  filter(abx_group == reference_group) %>%
  group_by(DOL_bin) %>%
  summarise(
    mu_ref = mean(Predicted_Microbiota_Age, na.rm = TRUE),
    sigma_ref = sd(Predicted_Microbiota_Age, na.rm = TRUE),
    .groups = "drop"
  )

# Merge stats back and calculate MAZ
combined_df2 <- combined_df2 %>%
  left_join(reference_stats, by = "DOL_bin") %>%
  mutate(MAZ = (Predicted_Microbiota_Age - mu_ref) / sigma_ref)

combined_df2$abx_group <- factor(combined_df2$abx_group, levels = c("Early and after", "Early only", "None"))

### Plot MAZ per group
MAZ_p <- ggplot(combined_df2, aes(x = abx_group, y = MAZ, fill = abx_group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  scale_color_manual(values = custom_colors_abx) + 
  scale_fill_manual(values = custom_colors_abx) + 
  stat_compare_means(method = "wilcox.test", label = "p.signif", ref.group = reference_group) +
  facet_wrap(~ Month) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 16, face = "bold"),
    legend.position = "none"
  ) +
  ylab("MAZ (Microbiota-for-Age Z-score)") +
  xlab(NULL)
ggsave("251211_GMdev_predicted_MAZ_Comp.pdf", plot = MAZ_p, bg = "transparent", width = 10, height = 8)


#save.image("251020_GM_dev_wP4_random_forests.RData")
```

Prevalence of top taxa over time (Related to Fig. 3A)
```{r}
otu_df <- as.data.frame(otu_table(merged_ps))
## Extract only species names from the taxa names
rownames(otu_df) <- sapply(str_split(rownames(otu_df), "\\|"), function(x) if(length(x) >= 7) x[7] else NA)
otu_df <- t(otu_df)
otu_df <- data.frame(otu_df)
otu_df <- otu_df %>% rownames_to_column("Sample_ID_R_final")

## Get taxonomy at desired level (genus, species)
tax_df <- as.data.frame(tax_table(ps_filt_dol60)) 
rownames(tax_df) <- sapply(str_split(rownames(tax_df), "\\|"), function(x) if(length(x) >= 7) x[7] else NA)
tax_df <- tax_df %>% rownames_to_column("TaxaID")

## Choose desired taxonomic rank
tax_level <- "Genus"

## Map TaxaID to taxonomic name (e.g., genus)
tax_map <- tax_df %>% select(TaxaID, !!sym(tax_level)) %>%
  mutate(Taxon = if_else(is.na(!!sym(tax_level)), TaxaID,!!sym(tax_level)))

## Reshape to long format
otu_long <- otu_df %>% pivot_longer(-Sample_ID_R_final, names_to = "TaxaID", values_to = "Abundance") %>%
  left_join(tax_map, by = "TaxaID")

## Add prevalence column 
otu_long <- otu_long %>% mutate(Prevalence = as.integer(Abundance > 0))

## Merge with sample metadata
meta_df <- as.data.frame(sample_data(merged_ps))

otu_long <- otu_long %>% left_join(meta_df, by = "Sample_ID_R_final")
otu_long_earlyOnly <- otu_long %>% filter(abx_group_updated == "Early only")
otu_long_earlyAfter <- otu_long %>% filter(abx_group_updated == "Early and after")
otu_long_naive <- otu_long %>% filter(abx_group_updated == "None")

custom_colors_prev <- c("g__Clostridioides" = "#E69F00", "g__Clostridium" = "#F0E442", "g__Escherichia" = "#56B4E9",
                                          "g__Enterobacter" = "#0072B2", "g__Klebsiella" = "#009E73", "g__Cutibacterium" = "#D55E00",
                                          "g__Veillonella" = "#CC79A7", "g__Staphylococcus" = "#2B2D42", "g__Enterococcus" = "#FC8D62",
                                          "g__Finegoldia" = "#8DA0CB")
custom_colors_top6 <- c("g__Escherichia" = "#9F7E59","g__Enterobacter" = "#8C2B0E", "g__Klebsiella" = "#2F4B7A", "g__Staphylococcus" = "#B47E83","g__Finegoldia" = "#8DA0CB","g__Enterococcus" = "#F2A450")

## Select top taxa based on average prevalence across all timepoints
top_taxa <- otu_long %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa, "260203_GM_Dev_top10_prevalent_taxa.csv", row.names = F)

top_taxa_eeo <- otu_long_earlyOnly %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa_eeo, "260203_GM_Dev_top10_prevalent_taxa_EEO.csv", row.names = F)

top_taxa_eas <- otu_long_earlyAfter %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa_eas, "260203_GM_Dev_top10_prevalent_taxa_EAS.csv", row.names = F)

top_taxa_naive <- otu_long_naive %>%
  group_by(Taxon) %>%
  summarise(AvgPrev = mean(Prevalence) * 100) %>%
  arrange(desc(AvgPrev)) %>%
  slice_head(n = 10) 
write.csv(top_taxa_naive, "260203_GM_Dev_top10_prevalent_taxa_ABN.csv", row.names = F)

## Select top 6 for plotting
top_taxa <- top_taxa %>% slice_head(n = 6) 

## prevalence summary over time 
prevalence_summary <- otu_long %>% filter(Taxon %in% top_taxa$Taxon) %>% group_by(Taxon, DOL, abx_group_names) %>% summarise(Prevalence = mean(Prevalence) * 100, mean(Prevalence), .groups = "drop") %>% 
  mutate(Taxon = factor(Taxon, levels = c("g__Escherichia", "g__Klebsiella", "g__Enterobacter", "g__Staphylococcus","g__Finegoldia","g__Enterococcus")))

prevalence_DOL_plot <- ggplot(prevalence_summary, aes(x = DOL, y = Prevalence, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  scale_color_manual(values = custom_colors_top6) + 
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Postnatal age [days]",
    y = "Prevalence [% of cohort]",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )
ggsave("260203_GM_Dev_NoAfterOnly_top6_prevalent_taxa_DOL.pdf", plot = prevalence_DOL_plot, bg = "transparent", width = 10, height = 8)


prevalence_DOL_plot_facet <- ggplot(prevalence_summary, aes(x = DOL, y = Prevalence, color = Taxon)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, size = 1.2) +
  facet_wrap(~ abx_group_names, nrow = 1) +
  scale_color_manual(values = custom_colors_top6) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Postnatal age [days]",
    y = "Prevalence [% of cohort]",
    color = "Taxon"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )
ggsave("260203_GM_Dev_NoAfterOnly_top6_prevalent_taxa_DOL_faceted.pdf", plot = prevalence_DOL_plot_facet, bg = "transparent", width = 10, height = 6)
```

Stacked barplots at the genus level for preterm and near-term infants (Related to Supplementary Fig. 4)
```{r}
## Barplot of taxonomic composition (Genus level) by Enterotype #######
# Step 1: Collapse to Genus level and transform to relative abundance
ps_genus <- tax_glom(ps_infants, taxrank = "Genus")
ps_relabun <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Step 2: Extract OTU table
otu_df <- as.data.frame(otu_table(ps_relabun))
if (taxa_are_rows(ps_relabun)) {
  otu_df <- t(otu_df)
}
otu_df <- as.data.frame(otu_df) %>%
  tibble::rownames_to_column("Sample_ID_R_final")

# Step 3: Pivot to long format
otu_long <- otu_df %>%
  pivot_longer(-Sample_ID_R_final, names_to = "OTU", values_to = "Abundance")

# Step 4: Merge sample metadata
sample_df <- data.frame(sample_data(ps_relabun)) 
otu_long <- left_join(otu_long, sample_df, by = "Sample_ID_R_final")

# Step 5: Add Genus information from tax_table
tax_df <- as.data.frame(tax_table(ps_relabun)) %>%
  tibble::rownames_to_column("OTU")

otu_long <- left_join(otu_long, tax_df[, c("OTU", "Genus")], by = "OTU")

# Step 6: Summarize mean relative abundance
otu_summary <- otu_long %>%
  filter(!is.na(Genus), !is.na(abx_group_names)) %>%
  group_by(abx_group_names, Genus, DOL_bin) %>%
  summarise(Mean_RelativeAbund = mean(Abundance), .groups = "drop")

# Step 7: Identify top 15 genera across all enterotypes
top_genera <- otu_summary %>%
  group_by(Genus) %>%
  summarise(Total = sum(Mean_RelativeAbund)) %>%
  arrange(desc(Total)) %>%
  head(15) %>%
  pull(Genus)

# Step 8: Group others into "Other"
otu_summary <- otu_summary %>%
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other")) %>%
  group_by(abx_group_names, Genus, DOL_bin) %>%
  summarise(Mean_RelativeAbund = sum(Mean_RelativeAbund), .groups = "drop")


### Create a color-palette with genera belonging to the same families/order with similar hues
custom_genus_colors <- c(

  # --- Bacteroidota: blue-greens ---
  "g__Bacteroides"      = "#088096",
  "g__Parabacteroides"  = "#7AD4E4",

  # --- Actinobacteriota: warm teals ---
  "g__Bifidobacterium"  = "#9AC8A4",
  "g__Cutibacterium"    = "#6EA57B",

  # --- Firmicutes Clostridia: orange-browns ---
  "g__Clostridium"        = "#B07159",
  "g__Erysipelatoclostridium" = "#6C3322",

  # --- Firmicutes Bacilli: pink-reds ---
  "g__Enterococcus"     = "#F2A450",
  "g__Staphylococcus"   = "#B47E83",
  "g__Streptococcus"    = "#68434E",

  # --- Proteobacteria: blue-purple tones ---
  "g__Citrobacter"      = "#DFCB91",
  "g__Enterobacter"     = "#8C2B0E",
  "g__Escherichia"      = "#9F7E59",
  "g__Klebsiella"       = "#2F4B7A",
  "g__Serratia"         = "#7C97C2",

  # --- Negativicutes (Veillonellaceae): olive-greens ---
  "g__Veillonella"      = "#A8A55C",
    # --- Other: neutral gray ---
  "Other"               = "#A9A9A9"   # darkgray
)

genus_order <- c(
  # Bacteroidota
  "g__Bacteroides",
  "g__Parabacteroides",

  # Actinobacteriota
  "g__Bifidobacterium",
  "g__Cutibacterium",

  # Firmicutes: Clostridia
  "g__Clostridium",
  "g__Erysipelatoclostridium",

  # Firmicutes: Bacilli
  "g__Enterococcus",
  "g__Staphylococcus",
  "g__Streptococcus",

  # Proteobacteria (Enterobacterales)
  "g__Citrobacter",
  "g__Enterobacter",
  "g__Escherichia",
  "g__Klebsiella",
  "g__Serratia",

  # Negativicutes
  "g__Veillonella",

  # Other
  "Other"
)
otu_summary$Genus <- factor(otu_summary$Genus, levels = genus_order)

# Step 9: Plot
abx_group_DOL_bin_genus_bp <- ggplot(otu_summary, aes(x = abx_group_names, y = Mean_RelativeAbund, fill = Genus)) +
  geom_bar(stat = "identity", width = 0.85, color = "black") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = custom_genus_colors) +
  facet_wrap(~ DOL_bin, nrow = 2, strip.position = "top") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 16, face = "bold"),  # Larger facet titles
    panel.spacing = unit(1.2, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  labs(
    title = "Average Genus-Level Composition by Enterotype",
    x = NULL, y = "Relative Abundance"
  )
ggsave("260204_P4_GM_Dev_near-term_barplot_genus.pdf", plot = abx_group_DOL_bin_genus_bp, bg = "transparent", width = 20, height = 8)
```

